### 2022-06-06 README.Rmd
paprpkgr::new_paper()

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Copeland.2022.hypoxia.flux

<!-- badges: start -->
<!-- badges: end -->

### Compendium URL

### Author
William Oldham (wm.oldham@gmail.com)

### Contents
This research compendium contains all data, code, and text associated with our publication:

>

The paper is available:

### Use

### Licenses

| **Text / figures :** [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)
| **Code :** See the [DESCRIPTION](DESCRIPTION) file.
| **Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/). Attribution requested in reuse.

### 2022-06-06 R/1_utils.R

@@ -1,0 +1,181 @@
+# utils.R
+
+#' Negate %in%
+#'
+#' Binary operator that returns a logical vector returning `TRUE` if there is
+#' not a match for its left operand.
+#'
+#' @param x The values to be matched
+#' @param y The values to be matched against
+#'
+"%nin%" <- function(x, table) {
+  match(x, table, nomatch = 0L) == 0L
+}
+
+path_to_data <- function(nm) {
+  dir(
+    path = "data-raw/",
+    pattern = nm,
+    all.files = TRUE,
+    full.names = TRUE,
+    recursive = TRUE,
+    include.dirs = TRUE
+  )
+}
+
+path_to_reports <- function(nm) {
+  stringr::str_c("analysis/", nm)
+}
+
+path_to_manuscript <- function(nm) {
+  stringr::str_c("manuscript/", nm)
+}
+
+path_to_plots <- function(nm) {
+  path <- stringr::str_c("analysis/figures/", nm)
+
+  if (dir.exists(path)) unlink(path, recursive = TRUE)
+
+  if (!dir.exists(path)) dir.create(path = path, recursive = TRUE)
+
+  path
+
+}
+
+read_multi_excel <- function(excel_file) {
+  sheets <- readxl::excel_sheets(excel_file)
+  purrr::map(sheets, ~readxl::read_excel(excel_file, sheet = .x)) %>%
+    rlang::set_names(sheets)
+}
+
+clean_technical_replicates <- function(tbl) {
+  tidyr::pivot_longer(
+    data = tbl,
+    cols = .data$a:.data$c,
+    names_to = "replicate",
+    values_to = "value"
+  ) %>%
+    dplyr::group_by(
+      dplyr::across(-c(.data$replicate, .data$value))
+    ) %>%
+    dplyr::mutate(value = replace_outliers(value)) %>%
+    dplyr::summarise(value = mean(.data$value, na.rm = TRUE)) %>%
+    dplyr::ungroup()
+}
+
+make_std_curves <- function(df, fo = NULL) {
+
+  if (is.null(fo)){
+    fo <- ~lm(value ~ conc, data = .x, na.action = modelr::na.warn)
+  }
+
+  df %>%
+    dplyr::filter(!is.na(.data$conc)) %>%
+    dplyr::select(where(~all(!is.na(.)))) %>%
+    dplyr::group_by(dplyr::across(-c(.data$conc, .data$value))) %>%
+    tidyr::nest() %>%
+    dplyr::mutate(
[…]

### 2022-06-06 _targets.R

@@ -1,0 +1,100 @@
+# setup -------------------------------------------------------------------
+
+devtools::load_all()
+library(targets)
+library(tarchetypes)
+
+invisible(
+  lapply(
+    list.files(path = "R", pattern = "\\.R$", full.names = TRUE),
+    source
+  )
+)
+
+conflicted::conflict_prefer("filter", "dplyr")
+
+options(
+  tidyverse.quiet = TRUE,
+  usethis.quiet = TRUE
+)
+
+future::plan(future.callr::callr(workers = future::availableCores() - 1))
+
+# target-specific options
+tar_option_set(
+  packages = c("tidyverse", "patchwork"),
+  # packages = c("tidyverse", "patchwork", "xcms"),
+  # imports = c("rnaseq.lf.hypoxia.molidustat"),
+  format = "qs"
+)
+
+# list of targets ---------------------------------------------------------
+
+list(
+
+  # write manuscript --------------------------------------------------------
+
+  # tar_target(
+  #   template,
+  #   system.file("manuscript/template.docx", package = "McGarrity.2022.hypoxia.omics"),
+  #   format = "file"
+  # ),
+  # tar_target(
+  #   pkgs,
+  #   system.file("manuscript/packages.bib", package = "McGarrity.2022.hypoxia.omics"),
+  #   format = "file"
+  # ),
+  # tar_target(
+  #   bib,
+  #   system.file("manuscript/library.json", package = "McGarrity.2022.hypoxia.omics"),
+  #   format = "file"
+  # ),
+  # tar_target(
+  #   csl,
+  #   system.file("manuscript/cell-metabolism.csl", package = "McGarrity.2022.hypoxia.omics"),
+  #   format = "file"
+  # ),
+  # tar_render(
+  #   manuscript,
+  #   path = path_to_manuscript("manuscript.Rmd"),
+  #   output_dir = path_to_manuscript(""),
+  #   output_format = bookdown::word_document2(
+  #     reference_docx = template,
+  #     df_print = "kable",
+  #     fig_caption = TRUE,
+  #     number_sections = FALSE,
+  #     pandoc_args = c(
+  #       "--lua-filter=scholarly-metadata.lua",
+  #       "--lua-filter=author-info-blocks.lua",
+  #       "--lua-filter=pagebreak.lua"
+  #     )
+  #   ),
+  #   params = list(
+  #     bibliography = c(bib, pkgs),
[…]

### 2022-06-06 R/1_figures.R
run check()

@@ -1,11 +1,6 @@
-# figures.R
+# 1_figures.R
 
 # plot setup
-pair_clrs <- RColorBrewer::brewer.pal(8, "Paired")
-names(pair_clrs) <-c("CTL.CTL", "TGFβ.CTL", "CTL.AZD3965", "TGFβ.AZD3965", "CTL.VB124", "TGFβ.VB124", "CTL.Dual", "TGFβ.Dual")
-
-clrs <- pair_clrs[1:2]
-names(clrs) <- c("CTL", "TGFβ")
 
 theme_plots <- function() {
   list(

### 2022-06-06 R/1_utils.R

@@ -1,16 +1,4 @@
-# utils.R
-
-#' Negate %in%
-#'
-#' Binary operator that returns a logical vector returning `TRUE` if there is
-#' not a match for its left operand.
-#'
-#' @param x The values to be matched
-#' @param y The values to be matched against
-#'
-"%nin%" <- function(x, table) {
-  match(x, table, nomatch = 0L) == 0L
-}
+# 1_utils.R
 
 path_to_data <- function(nm) {
   dir(

### 2022-06-06 analysis/dna-per-cell.Rmd
dna per cell number analysis

---
title: "Cell-DNA Standard Curves"
subtitle: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::pdf_document2:
    number_sections: true
    toc: true
    toc_depth: 1
    df_print: kable
    dev: cairo_pdf
    latex_engine: xelatex
geometry: margin = 0.5in
fontsize: 11pt
mainfont: Calibri
---

```{r setup, include = FALSE}
# load libraries
suppressPackageStartupMessages({
  devtools::load_all()
  library(tidyverse)
  library(wmo)
  library(targets)
})

# set global chunk options
knitr::opts_chunk$set(
	echo = FALSE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE,
	out.width = "49%"
)

options(knitr.table.format = function() {
  if (knitr::is_latex_output()) 
    "latex" else "html"
})

theme_set(theme_wmo(base_family = "Calibri"))
```

```{r}
withr::with_dir(here::here(), {
  std <- tar_read(dna_per_cell_std)
  sample <- tar_read(dna_per_cell_clean)
  cells_per_dna <- tar_read(cells_per_dna)
})
```

\newpage

# Overview

Cell counts were estimated from total cellular DNA measured by PicoGreen fluorescence (Quant-iT PicoGreen dsDNA Assay Kit, P11496, Thermo). Here, we determine the relationship between total DNA and cell count by quantifying DNA extracted from cells seeded at different densities in basal growth medium. For the lung fibroblasts, cells were extracted with 100 μL or 200 μL of TE buffer with 0.2% Triton X-100. Earlier experiments used the smaller volume of extraction buffer and this is adjusted for during the extracellular flux calculations.

# Analysis

## Standard Curves

```{r}
std |> 
  unnest(c(summary)) |> 
  ggplot() +
  aes(y = reorder(date, desc(date)), x = r.squared) +
  geom_point() +
  labs(
    y = "Date", 
    x = expression(paste("R"^2))
  )
```

## Samples

```{r}
sample |> 
  ggplot() +
  aes(x = cells, y = conc) +
  facet_wrap(~ cell_type * volume, labeller = as_labeller(toupper)) +
  geom_point(alpha = 0.3) +
  stat_summary(
    geom = "pointrange", 
    fun.data = "mean_se"
  ) +
  geom_smooth(
    method = "lm", 
    formula = y ~ 0 + x, 
    se = FALSE
  ) +
  labs(
    x = "Cell count", 
    y = "DNA (ng)"
  )
```

These plots suggest that, for the 100 μL extraction buffer, the maximum cell count is ~ 250,000. We will exclude the 300,000 value prior to calculate the slopes of the curves. For the PASMC samples, the 400,000 cell sample is quite high compared to the other samples. Will also remove these values prior to calculation of the slopes.

\newpage

```{r}
usethis::use_data(cells_per_dna, overwrite = TRUE)

cells_per_dna  |> 
  my_kable(
    digits = 0, 
    col.names = c("Cell Type", "Volume", "Cell / DNA"), 
    align = "c", 
    caption = "Cells per DNA (ng)"
  ) |> 
  kableExtra::kable_styling(position = "center") 
```

### 2022-06-06 R/1_dna-per-cell.R

@@ -1,0 +1,25 @@
+# 1_dna-per-cell.R
+
+clean_dna_per_cell <- function(filename) {
+  filename |>
+    read_multi_excel() |>
+    purrr::map(clean_technical_replicates) |>
+    dplyr::bind_rows(.id = "id") |>
+    tidyr::separate(id, c("cell_type", "volume"), sep = "_", convert = TRUE) |>
+    dplyr::mutate(cells = 1000 * cells)
+}
+
+calculate_cells_per_dna <- function(tbl) {
+  tbl %>%
+    dplyr::filter(!(cell_type == "lf" & volume == "100" & cells == 300000)) %>%
+    dplyr::filter(!(cell_type == "pasmc" & volume == "200" & cells == 400000)) %>%
+    dplyr::group_by(cell_type, volume) %>%
+    tidyr::nest() %>%
+    dplyr::mutate(
+      model = map(data, ~lm(conc ~ 0 + cells, data = .x, na.action = modelr::na.warn)),
+      glance = map(model, broom::tidy)
+    ) %>%
+    tidyr::unnest(c(glance)) %>%
+    dplyr::select(cell_type, volume, slope = estimate) %>%
+    dplyr::mutate(slope = 1/slope)
+}

### 2022-06-06 R/1_utils.R

@@ -2,7 +2,7 @@
 
 path_to_data <- function(nm) {
   dir(
-    path = "data-raw/",
+    path = "data-raw",
     pattern = nm,
     all.files = TRUE,
     full.names = TRUE,
@@ -32,7 +32,7 @@ path_to_plots <- function(nm) {
 
 read_multi_excel <- function(excel_file) {
   sheets <- readxl::excel_sheets(excel_file)
-  purrr::map(sheets, ~readxl::read_excel(excel_file, sheet = .x)) %>%
+  purrr::map(sheets, ~readxl::read_excel(excel_file, sheet = .x)) |>
     rlang::set_names(sheets)
 }
 
@@ -42,35 +42,43 @@ clean_technical_replicates <- function(tbl) {
     cols = .data$a:.data$c,
     names_to = "replicate",
     values_to = "value"
-  ) %>%
+  ) |>
     dplyr::group_by(
       dplyr::across(-c(.data$replicate, .data$value))
-    ) %>%
-    dplyr::mutate(value = replace_outliers(value)) %>%
-    dplyr::summarise(value = mean(.data$value, na.rm = TRUE)) %>%
+    ) |>
+    dplyr::mutate(value = replace_outliers(value)) |>
+    dplyr::summarise(value = mean(.data$value, na.rm = TRUE)) |>
     dplyr::ungroup()
 }
 
-make_std_curves <- function(df, fo = NULL) {
+replace_outliers <- function(vec) {
+  if (mad(vec, na.rm = TRUE) == 0) return (vec)
+  replace(
+    vec,
+    abs(vec - median(vec, na.rm = TRUE)) / mad(vec, na.rm = TRUE) > 2,
+    NA
+  )
+}
 
+make_std_curves <- function(df, fo = NULL) {
   if (is.null(fo)){
     fo <- ~lm(value ~ conc, data = .x, na.action = modelr::na.warn)
   }
 
   df %>%
-    dplyr::filter(!is.na(.data$conc)) %>%
-    dplyr::select(where(~all(!is.na(.)))) %>%
-    dplyr::group_by(dplyr::across(-c(.data$conc, .data$value))) %>%
+    dplyr::filter(!is.na(.data$conc)) |>
+    dplyr::select(where(~all(!is.na(.)))) |>
+    dplyr::group_by(dplyr::across(-c(.data$conc, .data$value))) |>
     tidyr::nest() %>%
     dplyr::mutate(
[…]

### 2022-06-06 _targets.R

@@ -15,7 +15,8 @@ conflicted::conflict_prefer("filter", "dplyr")
 
 options(
   tidyverse.quiet = TRUE,
-  usethis.quiet = TRUE
+  usethis.quiet = TRUE,
+  dplyr.summarise.inform = FALSE
 )
 
 future::plan(future.callr::callr(workers = future::availableCores() - 1))
@@ -31,6 +32,35 @@ tar_option_set(
 # list of targets ---------------------------------------------------------
 
 list(
+
+  # dna per cell ------------------------------------------------------------
+
+  tar_target(
+    dna_per_cell_file,
+    path_to_data("dna-per-cell-number.xlsx"),
+    format = "file"
+  ),
+  tar_target(
+    dna_per_cell_raw,
+    clean_dna_per_cell(dna_per_cell_file)
+  ),
+  tar_target(
+    dna_per_cell_std,
+    make_std_curves(dna_per_cell_raw)
+  ),
+  tar_target(
+    dna_per_cell_clean,
+    interp_data(dna_per_cell_raw, dna_per_cell_std)
+  ),
+  tar_target(
+    cells_per_dna,
+    calculate_cells_per_dna(dna_per_cell_clean)
+  ),
+  tar_render(
+    dna_per_cell_report,
+    path = path_to_reports("dna-per-cell.Rmd"),
+    output_dir = system.file("analysis/pdfs", package = "Copeland.2022.hypoxia.flux")
+  ),
 
   # write manuscript --------------------------------------------------------
 

### 2022-06-06 analysis/extracellular-fluxes.Rmd
added extracellular flux calculations

---
title: "Extracellular Fluxes"
subtitle: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::pdf_document2:
    number_sections: true
    toc: true
    toc_depth: 1
    df_print: kable
    dev: cairo_pdf
    latex_engine: xelatex
geometry: margin = 0.5in
fontsize: 11pt
mainfont: Calibri
---

```{r setup, include = FALSE}
# load libraries
suppressPackageStartupMessages({
  devtools::load_all()
  library(tidyverse)
  library(wmo)
  library(targets)
})

# resolve conflicts
conflicted::conflict_prefer("filter", "dplyr")

# set global chunk options
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE, 
  warning = FALSE, 
  fig.align = "center", 
  out.width = "49%"
)

options(knitr.table.format = function() {
  if (knitr::is_latex_output()) 
    "latex" else "html"
})

theme_set(theme_wmo(base_family = "Calibri"))
```

```{r}
withr::with_dir(here::here(), {
  data_raw <- tar_read(conc_raw)
  std <- tar_read(conc_std)
  std_fld <- tar_read(conc_std_clean_fld)
  std_clean <- tar_read(conc_std_clean)
  conc <- tar_read(conc_with_missing)
  evap <- tar_read(evap_raw)
  flux_measurements <- tar_read(flux_measurements)
  growth_curves <- tar_read(growth_curves)
  growth_rates <- tar_read(growth_rates)
  degradation_rates <- tar_read(degradation_rates)
  k <- tar_read(k)
  fluxes <- tar_read(fluxes)
})
```

\newpage

# Overview

Extracellular fluxes are determined from serial measurements of cell count and metabolite mass in conditioned medium accounting for evaporation and metabolite degradation/accumulation. Here, we calculate the extracellular fluxes included in the manuscript. Metabolic flux models assume metabolic steady state associated with exponential cell growth. Pilot experiments yielded the following conditions to monitor this time period:

*Lung fibroblasts.* Lung fibroblasts (LFs) were seeded at 25,000 cells per well of six-well plates on Day -1. Treatment was initiated on this day by placing plates in hypoxia (0.2% or 0.5% ambient oxygen) or treating with the prolyl hydroxylase inhibitor molidustat (10 μM, BAY, BAY85-3934, Cayman) with 0.1% DMSO as the vehicle control. The medium was changed on Day 0 and samples were collected this day and every 24 h for 72 h. The cell culture medium was MCDB131 (genDepot) supplemented with glucose (8 mM) and glutamine (1 mM), similar to medium used for isotope labeling experiments.

*Pulmonary artery smooth muscle cells.* Pulmonary artery smooth muscle cells (PASMCs) were seeded at 25,000 cells per well of six-well plates on Day -1. Treatment was initiated on this day by placing plates in hypoxia (0.5% ambient oxygen). The medium was changed on Day 0 and samples were collected at this time and every 12 h for 48 h. The cell culture medium was MCDB131 (genDepot) supplemented with glucose (5.551 mM) and glutamine (10 mM), similar to medium used for isotope labeling experiments.

*Hypoxia plus BAY.* When hypoxia was combined with BAY treatment to measure glucose and lactate fluxes, cells were cultured in complete growth medium (*e.g.*, FGM-2 for LFs).

```{r}
data_raw %>% 
  select(cell_type, experiment, date) %>% 
  distinct() %>% 
  group_by(cell_type, experiment) %>% 
  summarise(count = n()) %>% 
  my_kable(
    col.names = c("Cell type", "Experiment", "Count"), 
    caption = "Biological replicates"
  )
```

# Data Acquisition

Raw data are indexed to metadata by sample number, which corresponds to a unique combination of measurement type, treatment, time, and well. The sample type, `cells` or `empty`, refers to whether the medium was obtained from wells containing cells for flux measurements (`cells`) or from empty wells for degradation rate measurements (`empty`). The raw data for each experiment are contained in multi-sheet Excel files.

## Cell count

Cell counts were estimated from total cellular DNA measured by PicoGreen fluorescence (Quant-iT PicoGreen dsDNA Assay Kit, P11496, Thermo). The relationship between total DNA and cell count was determined by measuring DNA from cells seeded at different densities in basal growth medium. In the table below, the DNA standards were adjusted to incorporate the conversion from [DNA] to cell count. Lysis buffer volume changed between the two experimental batches, which affected the slope of the [DNA] *v*. cell count standard curve. Here, `conc` refers to cell count.

## Lactate

Lactate concentration was determined using an enzymatic assay (\textsc{l}-Lactate Assay Kit, 700510, Cayman). For batch `a`, samples were deproteinized with MPA per the manufacturer's protocol (2.1-fold dilution) and then diluted 5-fold prior to analysis (10.5-fold dilution total). Control experiments indicated that, for medium samples, deproteinization is not necessary (*i.e.*, signal intensities for unconditioned and conditioned medium were the same regardless of the deproteinization step). For subsequent batches, samples were diluted 10-fold in PBS prior to analysis. As for cell count, the lactate standard concentrations were adjusted to account for sample dilution.

## Glucose

Glucose concentration was determined using an enzymatic assay (Glucose Colorimetric Assay Kit, 10009582, Cayman). All samples were diluted 10-fold in water (batch `a`) or PBS (batches `b` and `c`) prior to analysis. Preliminary studies suggested no significant degradation of glucose. Glucose standard concentrations were adjusted to reflect unit conversion from mg/dL to μM and sample dilution prior to analysis.

## Pyruvate

Pyruvate concentration was measured multiple ways. For hypoxia batch `a` samples, it was determined by HPLC analysis of deproteinized OPD-derivatized samples. This analysis utilized ketovalerate (`KV`) as an internal standard (10 μL, 100 μM final concentration), which was added at the same time as the amino acid internal standards. For `bay` batch `a`, pyruvate concentration was determined by LC-MS analysis with `d8-valine` as an internal standard. For all batch `b` and `c` samples, pyruvate concentration was determined by enzymatic assay (Pyruvate Assay Kit, 700470, Cayman). For the enzymatic assay, samples were diluted 20-fold in PBS.

## Amino acids

Amino acid concentrations were determined by HPLC analysis of FMOC- and OPA-derivatized samples, producing chemicals detectable by fluorescent or UV/visible spectrophotometery, respectively. Prior to HPLC, internal standards (norvaline and sarcosine, 10 μL, 100 μM final concentration) were added to the samples (180 μL for batch `a` samples where pyruvate was measured by chromatography or 190 μL for all other samples) and they were deproteinized using ice-cold acetone (400 μL). Insoluble material was removed by centrifugation and the supernatant was evaporated to < 50% of the initial volume (< 100 μL) prior to HPLC.

\newpage

# Data Processing

All measured values are interpolated to concentration in μM for metabolites or cell count for DNA measurements.

## Standard curves

A review of the r^2^ values for the standard curves will identify problematic fits.

```{r, results = "hide", fig.show = "hold", fig.align = "left"}
plot_r_squared <- function(df, title) {
  ggplot(df) +
    aes(
      y = reorder(metabolite, desc(metabolite)), 
      x = r.squared, 
      color = interaction(date, batch, run)
    ) +
    geom_point(
      aes(shape = detector), 
      size = 3, 
      alpha = 0.6
    ) +
    labs(
      title = title, 
      x = expression(paste("r"^2)), 
      y = "Metabolite"
    ) +
    guides(color = guide_legend(title = NULL, nrow = 10))
}

std %>% 
  select(metabolite:run, summary) %>% 
  unnest(c(summary)) %>% 
  group_by(cell_type, experiment) %>% 
  nest() %>% 
  mutate(
    title = str_c(cell_type, experiment, sep = " "), 
    plots = map2(data, title, plot_r_squared)
  ) %>% 
  pull(plots)
```

These data suggest worse fits from fluorescent detection.

```{r, results = "hide", fig.show = "hold", fig.align = "left", out.width = "24.5%"}
std %>% 
  filter(detector == "fld", date == "2018-02-12", experiment == "02") %>% 
  pull(plots)
```

The fluorescence signal is plateauing between 250 and 1000 μM. The highest standard concentration will be removed from the curves.

In addition, there were a few standard curves that had poor fits:

```{r, fig.show = "hold", fig.align = "left", results = "hide"}
std_fld %>% 
  select(metabolite:run, summary) %>% 
  unnest(c(summary)) %>% 
  group_by(cell_type, experiment) %>% 
  nest() %>% 
  mutate(
    title = str_c(cell_type, experiment, sep = " "), 
    plots = map2(data, title, plot_r_squared)
  ) %>% 
  pull(plots)
```

After removing the high fluorescence standard, the remaining poorest fits were:

```{r}
poor_fits <- 
  std_fld %>% 
  ungroup() %>% 
  select(metabolite:run, summary, plots) %>% 
  unnest(c(summary)) %>% 
  slice_min(order_by = r.squared, n = 7)

poor_fits %>% 
  select(metabolite:r.squared) %>% 
  my_kable(caption = "Poor standard curve fits")
```

```{r, fig.show = "hold", fig.align = "left", results = "hide", out.width = "32.9%"}
poor_fits %>% 
  pull(plots)
```

When an apparent outlier was identified, it was removed from the standard curve.

```{r, fig.show = "hold", fig.align = "left", results = "hide", out.width = "32.9%"}
std_clean %>% 
  semi_join(poor_fits, by = c("metabolite", "experiment", "date", "batch", "run", "detector")) %>% 
  pull(plots)
```

## Missing data

The sample data frame was expanded to identify missing samples.

- For the 0.5% oxygen experiments, only three `empty` time 0 samples were collected and analyzed. The measurements from these samples will be used as time 0 to calculate degradation and accumulation rates for both treatments, 21% and 0.5%.

- For the LF 0.5% oxygen experiment performed 2017-11-06, the 48 h 21% `empty` sample was lost.

- Specific `empty` samples were not collected for the BAY treatment condition, only for the DMSO condition. Thus, the DMSO degradation rates will be duplicated and used for BAY treatment in the flux calculations.

- For the LF BAY experiment performed 2018-11-02, only one of three time 0 `empty` samples were collected.

# Analysis

## Pyruvate assays

Pyruvate was ultimately measured by three different methods, HPLC, LC-MS, and enzymatic assay.

```{r}
pyruvate <- 
  conc %>% 
  filter(metabolite == "pyruvate") %>%
  mutate(detector = factor(
    detector, 
    levels = c("enzyme", "hplc", "lcms"), 
    labels = c("Enzyme", "HPLC", "LC-MS")
  ))

pyruvate %>% 
  filter(conc < 1500) %>% 
  filter(time == 0) %>% 
  ggplot() + 
  aes(x = detector, y = conc, color = detector) +
  geom_jitter(show.legend = FALSE) +
  labs(
    title = "Pyruvate concentrations at time 0", 
    x = "Detection method", 
    y = "Concentration (μM)"
  )
```

Compared to the enzymatic assay, the HPLC measurements tended to underestimate the pyruvate concentration while the LC-MS measurements tended to overestimate the pyruvate concentration. I suspect the pyruvate determination using the enzymatic assay is more accurate and we will preferentially use these measurements to determine the fluxes. Unfortunately, for the 0.2% oxygen experiments, the only pyruvate measurements available were obtained using the HPLC assay, and so this assay will be used to estimate fluxes from those experiments.

## Amino acid measurements

Amino acid metabolites were measured using two detectors, fluorescence (`fld`) and multi-wavelength (`mwd`). We need to decide which detector is optimal for which amino acids based on the metabolite and its concentration in medium. Cystine is derivatized poorly by FMOC, and so it not detectable by fluorescence.

The fluorescence response is linear up to a concentration of ~ 250 μM on the basis of the standard curves. First, we will assess the metabolite concentrations across all of the samples.

```{r, fig.show = "hold"}
amino_acids <-
  conc %>% 
  filter(detector %in% c("mwd", "fld"))

plot_aa <- function(df) {
  df %>% 
    filter(!is.na(conc)) %>% 
    ggplot() + 
    aes(y = conc, x = metabolite, color = detector) + 
    geom_hline(yintercept = 300, linetype = 2, color = "gray") +
    geom_jitter(alpha = 0.1) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    labs(
      x = "Amino acid", 
      y = "Concentration (μM)"
    ) + 
    scale_color_discrete(name = "Detector", labels = c("FLD", "MWD")) +
    guides(color = guide_legend(override.aes = list(alpha = 1)))
}

amino_acids %>% 
  filter(cell_type == "lf") %>% 
  plot_aa() + ggtitle("LF amino acids")

amino_acids %>% 
  filter(cell_type == "pasmc") %>% 
  plot_aa() + ggtitle("PASMC amino acids")
```

Fluorescence detection appears to be associated with smaller variance when amino acid concentration is low. At higher concentrations, fluorescence detection is likely underestimating the concentration based on the standard curve analysis performed above. Based on these data, we will use concentrations determined by fluorescence detection for metabolites with an average concentration < 300 μM.

```{r}
amino_acids %>% 
  group_by(metabolite) %>% 
  summarise(conc = mean(conc, na.rm = TRUE)) %>% 
  mutate(detector = case_when(
    metabolite == "cystine" ~ "mwd", 
    conc > 300 ~ "mwd", 
    conc <= 300 ~ "fld"
  )) %>% 
  filter(detector == "mwd") %>% 
  select(metabolite) %>% 
  my_kable(caption = "MWD detection", col.names = "Metabolite")
```

## Evaporation

Preliminary experiments suggested that medium evaporation rates would have a significant impact on the interpretation of metabolite concentration measurements and that the evaporation rates differed between the normoxia and hypoxia incubators. To determine evaporation rate, a six-well plate was filled with medium and weighed during each sample collection. For 0.2% oxygen experiments, no evaporation controls were measured and so 0.5% oxygen rates were used instead. For BAY experiments, only one evaporation plate containing medium supplemented with 0.1% DMSO was used and this evaporation rate was applied to both conditions. Predicted well volumes were calculated from experimental measurements of plate weight assuming a starting volume of 2 mL per well.

```{r}
evap %>% 
  ggplot() +
  aes(
    x = time, 
    y = volume, 
    color = interaction(oxygen, experiment, sep = " | ")
  ) +
  geom_point(
    alpha = 0.3, 
    size = 2
  ) +
  geom_smooth(
    method = "lm", 
    formula = y ~ x, 
    se = FALSE, 
    show.legend = FALSE
  ) +
  stat_summary(
    geom = "pointrange", 
    fun.data = "mean_se", 
    alpha = 1, 
    fatten = 1, 
    show.legend = FALSE
  ) +
  guides(color = guide_legend(
    title = "Treatment", 
    override.aes = list(alpha = 1)
  )) +
  labs(
    tite = "Predicted well volumes", 
    x = "Time (h)", 
    y = "Volume (ml)"
  )
```

Specific evaporation rates for BAY-treated cells were not obtained, so the DMSO condition was duplicated. Evaporation controls were not obtained for BAY batch `a` or 0.2% oxygen experiments and so the average evaporation rates from BAY batch `b` and 0.5% oxygen experiments were duplicated and substituted, respectively. The predicted well volumes are attached to the interpolated sample concentrations or cell counts and made available as `flux_measurements`.

```{r}
usethis::use_data(flux_measurements, overwrite = TRUE)
```

# Flux Calculations

Determination of fluxes is based on the solution of the following differential equations:

$$ 
\begin{aligned}
\frac{dX}{dt} &= \mu X \\[6 pt]
\frac{dM}{dt} &= -kM+vX 
\end{aligned}
$$

where $X$ is the cell number, $\mu$ is cell growth rate, $M$ is moles of substrate, $k$ is the metabolite degradation rate, and $v$ is the extracellular flux rate.

These equations are solved as follows:

$$
\begin{aligned}
X &= X_0e^{\mu t} \\[6 pt]
Me^{k t} &= \frac{v X_0}{\mu + k} (e^{(\mu + k)t}-1) + M_0
\end{aligned}
$$

The data can be fit to a linear equation and the slope is used to calculate $v$.

## Growth rates

```{r, fig.show = "hold", fig.align = "left", results = "hide", out.width = "32.9%"}
growth_curves %>% 
  select(cell_type:data) %>% 
  unnest(c(data)) %>% 
  group_by(cell_type, experiment, batch, date, oxygen, virus, treatment, time) %>%
  summarize(conc = mean(conc, na.rm = TRUE)) %>% 
  filter(!is.na(time)) %>% 
  group_by(cell_type, experiment) %>% 
  nest() %>% 
  mutate(plot = map2(
    data, 
    cell_type, 
    ~ggplot(.x) + 
      aes(
        x = time, 
        y = conc/1000, 
        color = interaction(oxygen, treatment, virus, sep = " | ")
      ) +
      stat_summary(
        geom = "pointrange", 
        fun.data = "mean_se"
      ) +
      stat_summary(
        geom = "line", 
        fun.data = "mean_se", 
        show.legend = FALSE
      ) +
      labs(
        title = toupper(.y), 
        x = "Time (h)", 
        y = expression(paste("Cell count (\U00D7", "10"^3, ")")), 
        color = "condition"
      )
  )) %>% 
  pull(plot)
```

Robust linear models were fit to the logarithm of cell count *v*. time to determine growth rate (μ) and initial cell count (X0). The calculated growth rates are available as `growth_rates`.

```{r}
usethis::use_data(growth_rates, overwrite = TRUE)
```

```{r, fig.show = "hold", out.width = "49%"}
plot_growth_info <- function(df, y){
  ggplot(df) +
    aes(
      x = interaction(oxygen, treatment, virus, sep = "\n"), 
      y = .data[[y]], 
      color = interaction(oxygen, treatment)
    ) +
    facet_grid(
      ~cell_type, 
      scales = "free_x", 
      space = "free_x"
    ) +
    geom_point(
      aes(shape = batch), 
      size = 4, 
      alpha = 0.5, 
      show.legend = FALSE
    ) +
    stat_summary(
      fun.data = "mean_se", 
      geom = "crossbar", 
      width = 0.3, 
      show.legend = FALSE
    ) +
    labs(
      x = "Treatment", 
      y = y
    ) 
}

plot_growth_info(growth_rates, "mu") + ylab(expression(paste("Growth rate (h"^-1, ")")))
plot_growth_info(growth_rates, "X0") + ylab("Cell count at time 0")
```

## Degradation rates

Metabolite concentrations from unconditioned medium (`type == empty`) were used to calculate metabolite degradation or accumulation rates. Glutamine, for example, is known to degrade over time. When available, 96 h samples were included in these analyses to improve fits. Metabolite concentration was multiplied by predicted well volume to determine metabolite mass in nanomoles.

```{r, out.width = "100%"}
ggplot(degradation_rates) +
  aes(
    x = reorder(abbreviation, k), 
    y = k, 
    color = interaction(oxygen, treatment, sep = " | ")
  ) +
  geom_hline(yintercept = 0) +
  geom_point(
    aes(shape = batch), 
    alpha = 0.3
  ) +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  labs(
    title = "Degradation and accumulation rates", 
    x = "Metabolite", 
    y = expression(paste("Rate constant (h"^"-1", ")")), 
    color = "condition"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

There appears to be some treatment-dependent effects on degradation rates. Will group rates by treatment, remove outliers, and determine if the mean rates are different from 0. If so, these rates will be included in the metabolic flux calculations.

```{r}
usethis::use_data(k, overwrite = TRUE)
```

## Fluxes

```{r}
usethis::use_data(fluxes, overwrite = TRUE)
```

```{r, out.width = "75%"}
plot_fluxes <- function(df, title) {
  ggplot(df) + 
    aes(
      x = reorder(abbreviation, flux),
      y = flux,
      color = interaction(oxygen, treatment, sep = " | ")
    ) +
    geom_hline(yintercept = 0) +
    geom_point(
      aes(group = interaction(oxygen, treatment)),
      position = position_dodge(width = 0.5),
      size = 2,
      alpha = 0.3
    ) +
    stat_summary(
      fun.data = "mean_se",
      geom = "crossbar",
      width = 0.3,
      position = position_dodge(width = 0.5),
      show.legend = FALSE
    ) +
    labs(
      title = title,
      x = "Metabolite",
      y = "Flux (fmol / cell / h)",
      color = "Condition"
    ) +
    guides(colour = guide_legend(override.aes = list(alpha = 1))) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

fluxes %>%
  filter(metabolite %in% c("glucose", "lactate")) %>%
  plot_fluxes("Metabolite flux") + facet_wrap(~cell_type, nrow = 1)
```

```{r, results = "hide", out.width = "100%"}
fluxes %>% 
  filter(!(metabolite %in% c("glucose", "lactate"))) %>% 
  group_by(cell_type, experiment) %>% 
  nest() %>% 
  mutate(plots = map2(data, toupper(cell_type), plot_fluxes)) %>% 
  pull(plots)
```

\newpage

```{r}
fluxes %>%
  ungroup() %>%
  group_by(metabolite, cell_type, oxygen, treatment, virus) %>%
  summarise(
    v = mean(flux, na.rm = TRUE),
    sd = sd(flux, na.rm = TRUE),
    n = n(),
    se = sd / sqrt(n()),
    cv = sd / abs(v)
  ) %>%
  arrange(cell_type, metabolite, oxygen, treatment, virus) %>%
  my_kable(
    caption = "Flux rate summary",
    col.names = c(
      "Metabolite", 
      "Cell type", 
      "Oxygen", 
      "Treatment", 
      "Virus", 
      "Flux", 
      "SD", 
      "n", 
      "SEM", 
      "CV"
    ),
    align = "lccccccccc",
    digits = 2,
    longtable = TRUE,
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("repeat_header"),
    font_size = 8
  )
```

### 2022-06-06 R/1_extracellular-fluxes.R

@@ -1,0 +1,704 @@
+# 1_extracellular-fluxes.R
+
+clean_flux_meta <- function(file_list){
+  file_list %>%
+    rlang::set_names(stringr::str_extract(basename(.), "^.+(?=_)")) |>
+    purrr::map_dfr(
+      readr::read_csv,
+      col_types = "dccccdc",
+      .id = "experiment"
+    ) |>
+    tidyr::separate(experiment, c("cell_type", "experiment"), "_")
+}
+
+assemble_flux_data <- function(file_list){
+  nms <- sub("\\..*$", "", basename(file_list))
+  sheets <- unique(unlist(purrr::map(file_list, readxl::excel_sheets)))
+  data_list <- purrr::map(file_list, read_multi_excel)
+
+  purrr::map(sheets, ~purrr::map(data_list, .x)) |>
+    rlang::set_names(sheets) |>
+    purrr::map(rlang::set_names, nms) |>
+    purrr::map(dplyr::bind_rows, .id = "experiment")
+}
+
+clean_fluxes <- function(data_list){
+  df <-
+    data_list[c("dna", "glc", "lac", "pyr")] |>
+    dplyr::bind_rows(.id = "metabolite") |>
+    dplyr::filter(!is.na(.data$a)) %>%
+    dplyr::select(where(~any(!is.na(.)))) |>
+    clean_technical_replicates() |>
+    tidyr::separate(.data$experiment, c("cell_type", "experiment", "batch", "date"), "_")
+
+  dna <-
+    df |>
+    dplyr::filter(.data$metabolite == "dna") |>
+    dplyr::mutate(volume = dplyr::if_else(.data$date <= "2018-05-25", 100, 200)) |>
+    dplyr::left_join(cells_per_dna, by = c("cell_type", "volume")) |>
+    dplyr::mutate(
+      conc = .data$conc * slope,
+      metabolite = "dna",
+      detector = "picogreen"
+    ) |>
+    dplyr::select(-c(.data$volume, .data$slope))
+
+  others <-
+    df |>
+    dplyr::filter(.data$metabolite != "dna") |>
+    dplyr::mutate(
+      conc = dplyr::case_when(
+        metabolite == "lac" & batch == "a" ~ .data$conc * 10.5,
+        metabolite == "lac" & batch != "a" ~ .data$conc * 10,
+        metabolite == "glc" ~ .data$conc * 555.074,
+        metabolite == "pyr" ~ .data$conc * 20
+      ),
[…]

### 2022-06-06 R/data.R

@@ -13,3 +13,114 @@
 #'   \item{slope}{cells / ng DNA}
 #'   }
 "cells_per_dna"
+
+
+#' Measurements for extracellular flux calculations
+#'
+#' A data set containing the interpolated metabolite concentrations, cell counts,
+#' and evaporation volumes for extracellular flux determinations.
+#'
+#' \describe{
+#'   \item{cell_type}{
+#'   `lf` = lung fibroblasts \cr
+#'   `pasmc` = pulmonary artery smooth muscle cells}
+#'   \item{experiment}{
+#'   `02` = 0.2% oxygen for hypoxia \cr
+#'   `05` = 0.5% oxygen for hypoxia \cr
+#'   `bay` = molidustat treatment \cr
+#'   `05-bay` = 0.5% oxygen plus molidustat}
+#'   \item{batch}{a group of biological replicates analyzed similarly}
+#'   \item{date}{start date of an experiment}
+#'   \item{metabolite}{name of measurement}
+#'   \item{abbreviation}{abbreviated metabolite name}
+#'   \item{detector}{
+#'   `fld` = HPLC fluorescence detector \cr
+#'   `mwd` = HPLC multi-wavelength detector \cr
+#'   `enzyme` = enzymatic assay \cr
+#'   `hplc` = OPD-derivatized HPLC detection \cr
+#'   `lcms` = liquid chromatography-mass spectrometry assay \cr
+#'   `picogreen` = fluorescence dye labeling of DNA}
+#'   \item{type}{
+#'   `cells` = conditioned medium \cr
+#'   `empty` = unconditioned medium}
+#'   \item{oxygen}{ambient oxygen level}
+#'   \item{treatment}{
+#'   `None` = no treatment \cr
+#'   `DMSO` = 0.1% DMSO \cr
+#'   `BAY` = 10 μM molidustat}
+#'   \item{time}{hours}
+#'   \item{well}{denotes technical replicates in each experiment}
+#'   \item{conc}{measured in number for cells and μM for metabolites}
+#'   \item{volume}{measured in mL, extrapolated from evaporation controls}
+#'   \item{nmol}{metabolite mass}
+#'   }
+"flux_measurements"
+
+#' Calculated fluxes
+#'
+#' A data set containing metabolite fluxes.
+#'
+#'  \describe{
+#'   \item{metabolite}{name of measurement}
[…]

### 2022-06-06 _targets.R

@@ -23,7 +23,7 @@ future::plan(future.callr::callr(workers = future::availableCores() - 1))
 
 # target-specific options
 tar_option_set(
-  packages = c("tidyverse", "patchwork"),
+  packages = c("wmo", "tidyverse", "patchwork"),
   # packages = c("tidyverse", "patchwork", "xcms"),
   # imports = c("rnaseq.lf.hypoxia.molidustat"),
   format = "qs"
@@ -59,6 +59,133 @@ list(
   tar_render(
     dna_per_cell_report,
     path = path_to_reports("dna-per-cell.Rmd"),
+    output_dir = system.file("analysis/pdfs", package = "Copeland.2022.hypoxia.flux")
+  ),
+
+  # extracellular fluxes ----------------------------------------------------
+
+  tar_target(
+    fluxes_meta_files,
+    path_to_data("(lf|pasmc)_.*_meta\\.csv"),
+    format = "file"
+  ),
+  tar_target(
+    fluxes_meta,
+    clean_flux_meta(fluxes_meta_files)
+  ),
+  tar_target(
+    fluxes_data_files,
+    path_to_data("(lf|pasmc)_.*_[a-z]_\\d{4}-\\d{2}-\\d{2}\\.xlsx"),
+    format = "file"
+  ),
+  tar_target(
+    fluxes_data,
+    assemble_flux_data(fluxes_data_files)
+  ),
+  tar_target(
+    conc_raw,
+    clean_fluxes(fluxes_data)
+  ),
+  tar_target(
+    conc_std,
+    make_std_curves(conc_raw)
+  ),
+  tar_target(
+    conc_std_plots,
+    print_plots(conc_std$plots, conc_std$title, "fluxes/01_standard_curves"),
+    format = "file"
+  ),
+  tar_target(
+    conc_std_clean_fld,
+    make_std_curves(dplyr::filter(conc_raw, !(detector == "fld" & conc > 900)))
+  ),
+  tar_target(
+    conc_std_clean,
+    clean_flux_std(conc_raw)
+  ),
+  tar_target(
+    conc_interp,
+    interp_data(conc_raw, conc_std_clean)
+  ),
+  tar_target(
+    conc_with_missing,
+    fill_missing_fluxes(conc_interp, fluxes_meta)
+  ),
+  tar_target(
+    conc_clean,
+    filter_assays(conc_with_missing)
+  ),
+  tar_target(
+    evap_raw,
[…]

### 2022-06-06 analysis/extracellular-fluxes.Rmd
add q bias correction analysis

@@
 ```{r}
-data_raw %>% 
-  select(cell_type, experiment, date) %>% 
-  distinct() %>% 
-  group_by(cell_type, experiment) %>% 
-  summarise(count = n()) %>% 
+data_raw |> 
+  select(cell_type, experiment, date) |> 
+  distinct() |> 
+  group_by(cell_type, experiment) |> 
+  summarise(count = n()) |> 
   my_kable(
@@
 
-std %>% 
-  select(metabolite:run, summary) %>% 
-  unnest(c(summary)) %>% 
-  group_by(cell_type, experiment) %>% 
-  nest() %>% 
+std |> 
+  select(metabolite:run, summary) |> 
+  unnest(c(summary)) |> 
+  group_by(cell_type, experiment) |> 
+  nest() |> 
   mutate(
@@
     plots = map2(data, title, plot_r_squared)
-  ) %>% 
+  ) |> 
   pull(plots)
@@
 ```{r, results = "hide", fig.show = "hold", fig.align = "left", out.width = "24.5%"}
-std %>% 
-  filter(detector == "fld", date == "2018-02-12", experiment == "02") %>% 
+std |> 
+  filter(detector == "fld", date == "2018-02-12", experiment == "02") |> 
   pull(plots)
@@
 ```{r, fig.show = "hold", fig.align = "left", results = "hide"}
-std_fld %>% 
-  select(metabolite:run, summary) %>% 
-  unnest(c(summary)) %>% 
-  group_by(cell_type, experiment) %>% 
-  nest() %>% 
+std_fld |> 
+  select(metabolite:run, summary) |> 
+  unnest(c(summary)) |> 
+  group_by(cell_type, experiment) |> 
+  nest() |> 
   mutate(
@@
 poor_fits <- 
-  std_fld %>% 
-  ungroup() %>% 
-  select(metabolite:run, summary, plots) %>% 
-  unnest(c(summary)) %>% 
+  std_fld |> 
+  ungroup() |> 
+  select(metabolite:run, summary, plots) |> 
+  unnest(c(summary)) |> 
   slice_min(order_by = r.squared, n = 7)
 
-poor_fits %>% 
-  select(metabolite:r.squared) %>% 
+poor_fits |> 
+  select(metabolite:r.squared) |> 
   my_kable(caption = "Poor standard curve fits")
@@
 ```{r, fig.show = "hold", fig.align = "left", results = "hide", out.width = "32.9%"}
-poor_fits %>% 
+poor_fits |> 
   pull(plots)
@@
 ```{r, fig.show = "hold", fig.align = "left", results = "hide", out.width = "32.9%"}
-std_clean %>% 
-  semi_join(poor_fits, by = c("metabolite", "experiment", "date", "batch", "run", "detector")) %>% 
+std_clean |> 
+  semi_join(poor_fits, by = c("metabolite", "experiment", "date", "batch", "run", "detector")) |> 
   pull(plots)
@@
 pyruvate <- 
-  conc %>% 
-  filter(metabolite == "pyruvate") %>%
+  conc |> 
+  filter(metabolite == "pyruvate") |>
   mutate(detector = factor(
@@
 
-pyruvate %>% 
-  filter(conc < 1500) %>% 
-  filter(time == 0) %>% 
+pyruvate |> 
+  filter(conc < 1500) |> 
+  filter(time == 0) |> 
   ggplot() + 
@@
 amino_acids <-
-  conc %>% 
+  conc |> 
   filter(detector %in% c("mwd", "fld"))
@@
 plot_aa <- function(df) {
-  df %>% 
-    filter(!is.na(conc)) %>% 
+  df |> 
+    filter(!is.na(conc)) |> 
     ggplot() + 
@@
 
-amino_acids %>% 
-  filter(cell_type == "lf") %>% 
+amino_acids |> 
+  filter(cell_type == "lf") |> 
   plot_aa() + ggtitle("LF amino acids")
 
-amino_acids %>% 
-  filter(cell_type == "pasmc") %>% 
+amino_acids |> 
+  filter(cell_type == "pasmc") |> 
   plot_aa() + ggtitle("PASMC amino acids")
@@
 ```{r}
-amino_acids %>% 
-  group_by(metabolite) %>% 
-  summarise(conc = mean(conc, na.rm = TRUE)) %>% 
+amino_acids |> 
+  group_by(metabolite) |> 
+  summarise(conc = mean(conc, na.rm = TRUE)) |> 
   mutate(detector = case_when(
@@
     conc <= 300 ~ "fld"
-  )) %>% 
-  filter(detector == "mwd") %>% 
-  select(metabolite) %>% 
+  )) |> 
+  filter(detector == "mwd") |> 
+  select(metabolite) |> 
   my_kable(caption = "MWD detection", col.names = "Metabolite")
@@
 ```{r}
-evap %>% 
+evap |> 
   ggplot() +
@@
 ```{r, fig.show = "hold", fig.align = "left", results = "hide", out.width = "32.9%"}
-growth_curves %>% 
-  select(cell_type:data) %>% 
-  unnest(c(data)) %>% 
-  group_by(cell_type, experiment, batch, date, oxygen, virus, treatment, time) %>%
-  summarize(conc = mean(conc, na.rm = TRUE)) %>% 
-  filter(!is.na(time)) %>% 
-  group_by(cell_type, experiment) %>% 
-  nest() %>% 
+growth_curves |> 
+  select(cell_type:data) |> 
+  unnest(c(data)) |> 
+  group_by(cell_type, experiment, batch, date, oxygen, virus, treatment, time) |>
+  summarize(conc = mean(conc, na.rm = TRUE)) |> 
+  filter(!is.na(time)) |> 
+  group_by(cell_type, experiment) |> 
+  nest() |> 
   mutate(plot = map2(
@@
       )
-  )) %>% 
+  )) |> 
   pull(plot)
@@
 
-fluxes %>%
-  filter(metabolite %in% c("glucose", "lactate")) %>%
+fluxes |>
+  filter(metabolite %in% c("glucose", "lactate")) |>
   plot_fluxes("Metabolite flux") + facet_wrap(~cell_type, nrow = 1)
@@
 ```{r, results = "hide", out.width = "100%"}
-fluxes %>% 
-  filter(!(metabolite %in% c("glucose", "lactate"))) %>% 
-  group_by(cell_type, experiment) %>% 
-  nest() %>% 
-  mutate(plots = map2(data, toupper(cell_type), plot_fluxes)) %>% 
+fluxes |> 
+  filter(!(metabolite %in% c("glucose", "lactate"))) |> 
+  group_by(cell_type, experiment) |> 
+  nest() |> 
+  mutate(plots = map2(data, toupper(cell_type), plot_fluxes)) |> 
   pull(plots)
@@
 ```{r}
-fluxes %>%
-  ungroup() %>%
-  group_by(metabolite, cell_type, oxygen, treatment, virus) %>%
+fluxes |>
+  ungroup() |>
+  group_by(metabolite, cell_type, oxygen, treatment, virus) |>
   summarise(
@@
     cv = sd / abs(v)
-  ) %>%
-  arrange(cell_type, metabolite, oxygen, treatment, virus) %>%
+  ) |>
+  arrange(cell_type, metabolite, oxygen, treatment, virus) |>
   my_kable(
@@
     longtable = TRUE,
-  ) %>%
+  ) |>
   kableExtra::kable_styling(

### 2022-06-06 analysis/qbias-correction-factors.Rmd
add q bias correction analysis

---
title: "Q Bias Correction Factors"
subtitle: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::pdf_document2:
    number_sections: true
    toc: true
    toc_depth: 1
    df_print: kable
    dev: cairo_pdf
    latex_engine: xelatex
geometry: margin = 0.5in
fontsize: 11pt
mainfont: Calibri
---

```{r setup, include = FALSE}
# load libraries
suppressPackageStartupMessages({
  devtools::load_all()
  library(tidyverse)
  library(wmo)
  library(targets)
  library(mzrtools)
})

# resolve conflicts
conflicted::conflict_prefer("filter", "dplyr")

# set global chunk options
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE, 
  warning = TRUE, 
  fig.align = "center", 
  out.width = "49%"
)

options(knitr.table.format = function() {
  if (knitr::is_latex_output()) 
    "latex" else "html"
})

theme_set(theme_wmo(base_family = "Calibri"))
```

```{r}
withr::with_dir(here::here(), {
  qbias_ratios <- tar_read(qbias_ratios)
  correction_factors <- tar_read(correction_factors)
})
```

\newpage

# Overview

A selected ion monitoring method was utilized to increase sensitivity for isotope detection. Others have noted quadrupole bias on the Q Exactive using this detection method (*i.e.*, the sensitivity for each isotope differs relative to its position within the m/*z* window). To account for this, naturally labeled standards were analyzed using methods where the m/*z* window is shifted -1.003355 units, the ^13^C mass defect.

# Analysis

The M1/M0 ratio was calculated for each frame shift. The resulting data were fit with a third-degree polynomial function.

```{r}
make_plots <- function(data, batch, metabolite) {
  ggplot(data) +
    aes(x = window, y = ratio) +
    geom_smooth(
      method = MASS::rlm, 
      formula = y ~ poly(x, 3),
      color = "blue", 
      se = FALSE
    ) +
    geom_point(
      size = 2, 
      alpha = 0.5, 
      color = "black"
    ) +
    geom_point(
      size = 5, 
      stat = "summary", 
      fun.data = "mean_se", 
      color = "blue"
    ) +
    labs(
      title = str_c(metabolite, batch, sep = " "), 
      x = "Window", 
      y = "Ratio"
    ) 
}
```

```{r, results = "hide", out.width = "24.9%", fig.show = "hold"}
qbias_ratios |> 
  nest() |> 
  mutate(plots = pmap(list(data, batch, metabolite), make_plots)) |> 
  pull(plots)
```

```{r}
qbias_ratios |> 
  select(-ratio) |> 
  pivot_longer(c(M0, M1), names_to = "isotope", values_to = "value") |> 
  mutate(
    M = str_extract(isotope, "\\d") |> as.numeric(), 
    M = M + window, 
    iso = str_c("M", M)
  ) |> 
  left_join(correction_factors, by = c("metabolite", "batch", "iso" = "M")) |> 
  mutate(value = value * cf) |> 
  select(-c(cf, iso, M)) |> 
  pivot_wider(names_from = isotope, values_from = value) |> 
  mutate(ratio = M1 / M0) |> 
  group_by(metabolite, batch) |> 
  nest() |> 
  mutate(plots = pmap(list(data, batch, metabolite), make_plots)) |> 
  pull(plots)
```

```{r}
usethis::use_data(correction_factors, overwrite = TRUE)
```

### 2022-06-06 R/1_qbias.R

@@ -1,0 +1,115 @@
+# 1_qbias.R
+
+isotope_library <-
+  tibble::tribble(
+    ~ metabolite, ~ formula, ~ polarity,
+    "2HG", "C5H8O5", "negative",
+    "2OG", "C5H6O5", "negative",
+    "alanine", "C3H7NO2", "negative",
+    "aspartate", "C4H7NO4", "negative",
+    "citrate", "C6H8O7", "negative",
+    "glutamate", "C5H9NO4", "negative",
+    "glutamine", "C5H10N2O3", "negative",
+    "lactate", "C3H6O3", "negative",
+    "malate", "C4H6O5", "negative",
+    "pyruvate", "C3H4O3", "negative",
+    "serine", "C3H7NO3", "negative",
+    "succinate", "C4H6O4", "negative",
+    "3PG", "C3H7O7P", "negative",
+    "aconitate", "C6H6O6", "negative",
+    "FBP", "C6H14O12P2", "negative",
+    "G3P", "C3H9O6P", "negative",
+    "palmitate", "C16H32O2", "negative",
+    "PEP", "C3H5O6P", "negative",
+    "sedoheptulose", "C7H14O7", "negative",
+    "DHAP", "C3H7O6P", "negative",
+    "GAP", "C3H7O6P", "negative",
+    "G1P", "C6H13O9P", "negative",
+    "G6P", "C6H13O9P", "negative",
+    "R5P", "C5H11O8P", "negative"
+  )
+
+calculate_ratios <- function(file_name){
+  readr::read_csv(file_name) |>
+    dplyr::filter(!is.na(area)) |>
+    tidyr::separate(
+      .data$filename,
+      into = c(NA, "window", "replicate"),
+      sep = c(1, 2),
+      convert = TRUE
+    ) |>
+    tidyr::separate(
+      .data$ion,
+      into = c("metabolite", "isotope"),
+      sep = " ",
+      convert = TRUE
+    ) |>
+    dplyr::mutate(carbons = case_when(
+      .data$metabolite %in% c("citrate") ~ 6,
+      .data$metabolite %in% c("2HG", "2OG", "glutamate", "glutamine") ~ 5,
+      .data$metabolite %in% c("aspartate", "malate", "succinate") ~ 4,
+      .data$metabolite %in% c("lactate", "pyruvate", "alanine", "serine") ~ 3
+    )) |>
+    dplyr::filter(.data$window <= .data$carbons) |>
[…]

### 2022-06-06 R/data.R

@@ -124,3 +124,18 @@
 #'   \item{k}{accumulation or decay rate per hour, positive values indicate decay}
 #'   }
 "k"
+
+#' Quadrupole bias correction factors
+#'
+#' A data set containing experimentally determined correction factors to adjust
+#' peak areas for isotopes detected in selected ion monitoring mode. Isotope peak
+#' areas are multiplied by the correction factor. The corrected peak area is
+#' subsequently used to determine the mass isotope distribution.
+#'
+#'  \describe{
+#'    \item{batch}{correction factors associated with specific experiments}
+#'    \item{metabolite}{name of measurement}
+#'    \item{M}{isotope}
+#'    \item{cf}{correction factor}
+#'   }
+"correction_factors"

### 2022-06-06 _targets.R

@@ -189,6 +189,31 @@ list(
     output_dir = system.file("analysis/pdfs", package = "Copeland.2022.hypoxia.flux")
   ),
 
+  # q bias correction -------------------------------------------------------
+
+  tar_target(
+    qbias_files,
+    path_to_data("q-bias-correction"),
+    format = "file"
+  ),
+  tar_target(
+    qbias_ratios,
+    import_qbias(qbias_files)
+  ),
+  tar_target(
+    pred_ratios,
+    calculate_predicted_ratios()
+  ),
+  tar_target(
+    correction_factors,
+    calculate_correction_factors(qbias_ratios, pred_ratios)
+  ),
+  tar_render(
+    qbias_correction_factor_report,
+    path = path_to_reports("qbias-correction-factors.Rmd"),
+    output_dir = system.file("analysis/pdfs", package = "Copeland.2022.hypoxia.flux")
+  ),
+
   # write manuscript --------------------------------------------------------
 
   # tar_target(

### 2022-06-06 R/1_mids.R
add mids

@@ -1,0 +1,196 @@
+# 1_mids.R
+
+clean_mids <- function(mid_files){
+  read_function <- function(filename){
+    readr::read_csv(filename) %>%
+      replace(is.na(.), 0)
+  }
+
+  mid_files %>%
+    rlang::set_names(sub("\\..*$", "", basename(.))) |>
+    purrr::map_dfr(read_function, .id = "experiment") |>
+    tidyr::pivot_longer(
+      cols = .data$`2HG M0`:tidyselect::last_col(),
+      names_to = "ion",
+      values_to = "peak_area",
+      values_drop_na = TRUE
+    ) |>
+    dplyr::mutate(
+      batch = stringr::str_extract(.data$experiment, "a|b|c"),
+      method = stringr::str_extract(.data$experiment, "fs|sim"),
+      cell_type = stringr::str_extract(.data$experiment, "lf|pasmc"),
+      date = stringr::str_extract(.data$experiment, "\\d{4}-\\d{2}-\\d{2}"),
+      .before = .data$experiment
+    ) |>
+    dplyr::select(-.data$experiment) |>
+    tidyr::separate(
+      .data$id,
+      c("tracer", "treatment", "time", "well"),
+      sep = " "
+    ) |>
+    tidyr::separate(.data$ion, into = c("metabolite", "isotope")) |>
+    dplyr::mutate(
+      peak_area = tidyr::replace_na(peak_area, 0),
+      oxygen = dplyr::case_when(
+        .data$treatment %in% c("21%", "dmso", "bay") ~ "21%",
+        .data$treatment == "0.5%" ~ "0.5%",
+        .data$treatment == "0.2%" ~ "0.2%"
+      ),
+      treatment = dplyr::case_when(
+        .data$treatment %in% c("21%", "0.5%", "0.2%") ~ "None",
+        TRUE ~ toupper(.data$treatment)
+      ),
+      oxygen = factor(.data$oxygen, levels = c("21%", "0.5%", "0.2%")),
+      treatment = factor(.data$treatment, levels = c("None", "DMSO", "BAY")),
+      time = as.double(.data$time),
+      time = dplyr::case_when(
+        .data$cell_type == "lf" ~ time * 24,
+        .data$cell_type == "pasmc" ~ time * 12
+      )
+    ) |>
+    dplyr::relocate(.data$oxygen, .before = .data$treatment) |>
+    dplyr::select(-filename)
+}
+
+correct_mid <- function(mid_clean){
+  correction_factors <-
[…]

### 2022-06-06 R/1_utils.R

@@ -1,5 +1,9 @@
 # 1_utils.R
 
+"%nin%" <- function(x, table) {
+  match(x, table, nomatch = 0L) == 0L
+}
+
 path_to_data <- function(nm) {
   dir(
     path = "data-raw",

### 2022-06-06 _targets.R

@@ -23,7 +23,7 @@ future::plan(future.callr::callr(workers = future::availableCores() - 1))
 
 # target-specific options
 tar_option_set(
-  packages = c("wmo", "tidyverse", "patchwork"),
+  packages = c("tidyverse", "patchwork"),
   # packages = c("tidyverse", "patchwork", "xcms"),
   # imports = c("rnaseq.lf.hypoxia.molidustat"),
   format = "qs"
@@ -213,6 +213,44 @@ list(
     path = path_to_reports("qbias-correction-factors.Rmd"),
     output_dir = system.file("analysis/pdfs", package = "Copeland.2022.hypoxia.flux")
   ),
+
+  # mids --------------------------------------------------------------------
+
+  tar_target(
+    mid_files,
+    path_to_data("(a|b)_(fs|sim)_(lf|pasmc)_.*\\.csv"),
+    format = "file"
+  ),
+  tar_target(
+    mid_clean,
+    clean_mids(mid_files)
+  ),
+  tar_target(
+    mid_correct,
+    correct_mid(mid_clean)
+  ),
+  tar_target(
+    mids,
+    remove_mid_outliers(mid_correct)
+  ),
+  tar_target(
+    mid_curves,
+    plot_mid_curves(mids)
+  ),
+  tar_target(
+    mid_curve_plots,
+    print_plots(mid_curves$plots, mid_curves$title, "mids"),
+    format = "file"
+  ),
+  tar_render(
+    mid_report,
+    path = path_to_reports("mass-isotope-distributions.Rmd"),
+    output_dir = system.file("analysis/pdfs", package = "Copeland.2022.hypoxia.flux")
+  ),
+  # tar_target(
+  #   pasmc_m5,
+  #   get_m5_citrate(pruned_mids)
+  # ),
 
   # write manuscript --------------------------------------------------------
 

### 2022-06-06 analysis/mass-isotope-distributions.Rmd
add biomass equations

---
title: "Mass Isotopomer Distributions"
subtitle: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::pdf_document2:
    number_sections: true
    toc: true
    toc_depth: 1
    df_print: kable
    dev: cairo_pdf
    latex_engine: xelatex
geometry: margin = 0.5in
fontsize: 11pt
mainfont: Calibri
---

```{r setup, include = FALSE}
# load libraries
suppressPackageStartupMessages({
  devtools::load_all()
  library(tidyverse)
  library(wmo)
  library(targets)
})

# resolve conflicts
conflicted::conflict_prefer("filter", "dplyr")

# set global chunk options
knitr::opts_chunk$set(
	echo = FALSE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE,
	out.width = "49%"
)

options(knitr.table.format = function() {
  if (knitr::is_latex_output()) 
    "latex" else "html"
})

theme_set(theme_wmo(base_family = "Calibri"))
```

```{r}
withr::with_dir(here::here(), {
  mids <- tar_read(mids)
})
```

```{r}
usethis::use_data(mids, overwrite = TRUE)
```

\newpage

# Overview

Mass isotopomer distributions (MIDs) were determined for key intracellular metabolites. Cells were seeded on Day -1 in "light" labeling medium. On Day 0, medium was changed to "heavy" labeling medium containing an equimolar concentration of the stable isotope substrate. These experiments utilized the following substrates: [1,2-^13^C~2~] glucose, [U-^13^C~6~] glucose, [U-^13^C~5~] glutamine, and [U-^13^C~3~] lactate. Intracellular metabolites for every time point were extracted in 80% MeOH pre-cooled to -80 °C and analyzed by LC-MS. Each sample was run using two different LC-MS methods. First, selected ion monitoring (SIM) was used to enhance the sensitivity for key metabolites used in modeling. Second, a polarity-switching, full scan (FS) method was used. The full scan method provided labeling insights for metabolites that were not previously targeted.

# Data Processing

Peaks were interpolated using either Thermo Xcalibur or TraceFinder software.

## Quadrupole bias correction

A selected ion monitoring method was utilized to increase sensitivity for isotope detection. Others have noted quadrupole bias on the Q Exactive using this detection method (*i.e.*, the sensitivity for each isotope differs relative to its position within the m/*z* window). To account for this, naturally labeled standards were analyzed using methods where the m/*z* window is shifted -1.003355 units, the ^13^C mass defect. The M1/M0 ratio was calculated for each frame shift and corrected to the theoretical ratio. These correction factors are applied to the raw MID data.

## Natural abundance correction matrices

Isotope correction matrices are generated from the metabolite's molecular formula using a custom R package (`wmoldham/mzrtools`).

## Correct for natural isotope abundance

MIDs are calculated from the peak areas corrected for quadrupole bias for each metabolite within each run. The correction matrices are are then used to adjust experiment MIDs for the natural abundance of ^13^C.

# Analysis

```{r}
mids_filtered <-
  mids |> 
  group_by(across(-c(date, mid))) |> 
  remove_nested_outliers(mid, remove = TRUE)
```

```{r}
mids_filtered |> 
  pivot_wider(names_from = method, values_from = mid) |> 
  filter(!is.na(sim)) |>  
  ggplot() + 
  aes(
    x = sim, 
    y = fs, 
    color = isotope
  ) +
  facet_wrap(~metabolite) +
  geom_abline(slope = 1) +
  geom_point() + 
  scale_color_manual(values = viridis::viridis(7)) +
  labs(
    x = "Selected ion monitoring", 
    y = "Full scan", 
    title = "Method comparison"
  )
```

There is reasonably good agreement between full scan and SIM methods for several of the metabolites. This observation suggests that those metabolites detected in full scan mode only are reasonable approximations. Certainly, SIM mode is preferred for metabolites detected with it, particularly as the MID correction seems to be better.

```{r remove-excess-fs}
fs <-
  mids_filtered |> 
  filter(method == "fs") |> 
  pull(metabolite) |> 
  unique()

sim <- 
  mids_filtered |> 
  filter(method == "sim") |> 
  pull(metabolite) |> 
  unique()

fs <- setdiff(fs, sim)

method_filter <-
  tibble(
    metabolite = c(fs, sim), 
    method = c(rep("fs", length(fs)), rep("sim", length(sim)))
  )

mids_method <-
  mids_filtered |> 
  semi_join(method_filter, by = c("metabolite", "method"))
```

```{r tracer-label-key}
tracer_labels <- c(
  expression(paste("[1,2-"^13, "C"[2], "] glucose")),
  expression(paste("[U-"^13, "C"[6], "] glucose")), 
  expression(paste("[U-"^13, "C"[5], "] glutamine")), 
  expression(paste("[U-"^13, "C"[3], "] lactate"))
)
tracer_levels <- c("glc2", "glc6", "q5", "lac3")
```

```{r, results = "hide", fig.show = "hold", out.width = "49%"}
mids_method |> 
  arrange(metabolite) |> 
  filter((cell_type == "lf" & time == 72) | (cell_type == "pasmc" & time == 48)) |> 
  group_by(method, cell_type, metabolite) |> 
  mutate(tracer = factor(tracer, levels = tracer_levels, labels = tracer_labels)) |> 
  nest() |> 
  mutate(plots = pmap(
    list(data, metabolite, method, cell_type),   
    ~ ggplot(data = ..1) + 
      aes(
        x = isotope, 
        y = mid, 
        color = interaction(oxygen, treatment, sep = " | ")
      ) +
      facet_wrap(
        ~tracer, 
        nrow = 2, 
        ncol = 2, 
        scales = "free_x", 
        labeller = label_parsed
      ) +
      stat_summary(
        geom = "pointrange", 
        position = position_dodge(width = 0.3), 
        fun.data = "mean_se"
      ) +
      labs(
        x = "Isotope", 
        y = "Mole fraction", 
        title = stringr::str_c(toupper(..3), " ", ..2, " ", toupper(..4)), 
        color = NULL
      )
  )) |> 
    pull(plots)
```

### 2022-06-06 R/1_biomass.R

@@ -1,0 +1,169 @@
+# 1_biomass.R
+
+clean_biomass <- function(biomass_file) {
+  readr::read_csv(biomass_file) |>
+    dplyr::mutate(
+      diff = post-pre,
+      cell_mass = diff / cell_number * 1E12
+    )
+}
+
+cell_composition <-
+  tibble::tribble(
+    ~ metabolite, ~ μmol_per_gDW,
+    "ALA", 0.6,
+    "ARG", 0.377,
+    "ASP", 0.359,
+    "ASN", 0.288,
+    "CYS", 0.145,
+    "GLN", 0.322,
+    "GLU", 0.386,
+    "GLY", 0.538,
+    "HIS", 0.143,
+    "ILE", 0.324,
+    "LEU", 0.564,
+    "LYS", 0.57,
+    "MET", 0.138,
+    "PHE", 0.219,
+    "PRO", 0.313,
+    "SER", 0.43,
+    "THR", 0.386,
+    "TRP", 0.044,
+    "TYR", 0.182,
+    "VAL", 0.416,
+    "glycogen", 0.279,
+    "dAMP", 0.0148,
+    "dGMP", 0.0099,
+    "dCMP", 0.0099,
+    "dTMP", 0.0148,
+    "AMP", 0.033,
+    "GMP", 0.0624,
+    "CMP", 0.0551,
+    "UMP", 0.033,
+    "cholesterol", 0.018,
+    "phosphatidylcholine", 0.069,
+    "phosphatidylethanolamine", 0.026,
+    "phosphatidylinositol", 0.01,
+    "phosphatidylserine", 0.003,
+    "phosphatidylglycerol", 0.001,
+    "sphingomyelin", 0.008,
+    "cardiolipin", 0.003
+  ) |>
+  dplyr::mutate(class = dplyr::case_when(
+    .data$metabolite %in% c("dAMP", "dGMP", "AMP", "GMP") ~ "purine",
+    .data$metabolite %in% c("dCMP", "CMP", "UMP") ~ "pyrimidine",
+    .data$metabolite == "dTMP" ~ "thymine",
+    TRUE ~ metabolite
+  ))
+
+metabolite_composition <-
+  tibble::tribble(
+    ~ class, ~ component, ~ stoichiometry,
+    "purine", c("P5P", "GLY", "CO2", "MEETHF"), c(1, 1, 1, 2),
[…]

### 2022-06-06 _targets.R

@@ -252,6 +252,36 @@ list(
   #   get_m5_citrate(pruned_mids)
   # ),
 
+  # biomass -----------------------------------------------------------------
+
+  tar_target(
+    biomass_file,
+    path_to_data("cell-mass.csv"),
+    format = "file"
+  ),
+  tar_target(
+    biomass_clean,
+    clean_biomass(biomass_file)
+  ),
+  tar_target(
+    biomass,
+    calculate_biomass(biomass_clean)
+  ),
+  tar_target(
+    biomass_equations,
+    calculate_biomass_equations(biomass)
+  ),
+  tar_target(
+    biomass_equations_out,
+    write_matlab_input(biomass_equations, coefs, "_biomass.csv"),
+    format = "file"
+  ),
+  tar_render(
+    biomass_report,
+    path = path_to_reports("biomass.Rmd"),
+    output_dir = system.file("analysis/pdfs", package = "Copeland.2022.hypoxia.flux")
+  ),
+
   # write manuscript --------------------------------------------------------
 
   # tar_target(

### 2022-06-06 R/1_matlab.R
add matlab input formatting

@@ -1,0 +1,170 @@
+# 1_matlab.R
+
+format_reactions <- function(reactions_file){
+  model_reactions <-
+    readr::read_csv(reactions_file) |>
+    dplyr::filter(!is.na(.data$name)) |>
+    dplyr::mutate(
+      pathway = factor(pathway, levels = c(
+        "transport",
+        "glycolysis",
+        "pentose phosphate pathway",
+        "anaplerosis",
+        "tricarboxylic acid cycle",
+        "amino acid metabolism",
+        "biomass",
+        "mixing",
+        "dilution"
+      ))
+    )
+
+  usethis::use_data(model_reactions, overwrite = TRUE)
+
+  model_reactions
+}
+
+format_fluxes <- function(growth_rates, fluxes){
+  growth <-
+    growth_rates |>
+    dplyr::mutate(metabolite = "biomass") |>
+    dplyr::select(-.data$X0) |>
+    dplyr::rename(flux = .data$mu)
+
+  fluxes_final <-
+    fluxes |>
+    dplyr::filter(.data$experiment %in% c("05", "02", "bay")) |>
+    dplyr::filter(!(.data$experiment == "02" & metabolite == "pyruvate")) |>
+    dplyr::filter(!(.data$experiment == "02" & oxygen == "0.2%")) |>
+    dplyr::select(-.data$abbreviation) |>
+    dplyr::bind_rows(growth) |>
+    dplyr::group_by(.data$metabolite, .data$cell_type, .data$oxygen, .data$treatment) |>
+    wmo::remove_nested_outliers(flux, remove = TRUE)
+
+  flux_names <- tribble(
+    ~ metabolite, ~ name,
+    "biomass", "BIOMASS",
+    "alanine", "ALAR",
+    "glucose", "GLUT",
+    "glutamine", "GLNR",
+    "glutamate", "GLUR",
+    "lactate", "MCT",
+    "pyruvate", "PYRR",
+    "serine", "SERR",
+    "cystine", "CYSR",
+    "glycine", "GLYR",
+    "aspartate", "ASPR"
+  )
+
+  model_fluxes <- c(
+    "biomass",
+    "alanine",
+    "glucose",
+    "glutamine",
+    "glutamate",
+    "lactate",
+    "pyruvate",
+    "serine",
+    "cystine",
+    "glycine",
+    "aspartate"
+  )
+
+  fluxes_final |>
[…]

### 2022-06-06 R/data.R

@@ -139,3 +139,15 @@
 #'    \item{cf}{correction factor}
 #'   }
 "correction_factors"
+
+#' Model reactions
+#'
+#' A data table of model reactions formatted for INCA analysis.
+#'
+#'  \describe{
+#'   \item{index}{reaction identifier}
+#'   \item{name}{reaction name}
+#'   \item{equation}{equation for INCA input}
+#'   \item{pathway}{metabolic pathway}
+#'   }
+"model_reactions"

### 2022-06-06 _targets.R

@@ -282,6 +282,40 @@ list(
     output_dir = system.file("analysis/pdfs", package = "Copeland.2022.hypoxia.flux")
   ),
 
+  # matlab ------------------------------------------------------------------
+
+  tar_target(
+    reactions_file,
+    path_to_reports("modeling/matlab-input/reactions.csv"),
+    format = "file"
+  ),
+  tar_target(
+    model_reactions,
+    format_reactions(reactions_file)
+  ),
+  tar_target(
+    model_fluxes,
+    format_fluxes(growth_rates, fluxes)
+  ),
+  tar_target(
+    model_fluxes_out,
+    write_matlab_input(model_fluxes, data, "_fluxes.csv"),
+    format = "file"
+  ),
+  tar_target(
+    pruned_mids,
+    format_mids(mids)
+  ),
+  tar_target(
+    model_mids,
+    summarize_mids(pruned_mids)
+  ),
+  tar_target(
+    model_mids_out,
+    write_matlab_input(model_mids, data, "_mids.csv"),
+    format = "file"
+  ),
+
   # write manuscript --------------------------------------------------------
 
   # tar_target(

### 2022-06-06 R/1_viability.R
add viability

@@ -1,0 +1,11 @@
+# 1_viability.R
+
+clean_viability <- function(viability_file) {
+  readr::read_csv(viability_file) |>
+    dplyr::mutate(
+      viability = 100 * live / (dead + live),
+      oxygen = factor(oxygen, levels = c("21%", "0.5%"))
+    ) |>
+    dplyr::group_by(time, oxygen) |>
+    wmo::remove_nested_outliers(viability, remove = TRUE)
+}

### 2022-06-06 _targets.R

@@ -316,6 +316,18 @@ list(
     format = "file"
   ),
 
+  # viability ---------------------------------------------------------------
+
+  tar_target(
+    viability_file,
+    path_to_data("cell-viability.csv"),
+    format = "file"
+  ),
+  tar_target(
+    viability,
+    clean_viability(viability_file)
+  ),
+
   # write manuscript --------------------------------------------------------
 
   # tar_target(

### 2022-06-06 R/1_expression.R
add immunoblots

@@ -1,0 +1,62 @@
+# 1_expression.R
+
+read_data <- function(data_files){
+  data_files[stringr::str_detect(data_files, "\\.csv$")] %>%
+    rlang::set_names(stringr::str_extract(., "(lf|pasmc)_(02|05-bay|05|bay|)")) |>
+    purrr::map_dfr(read_csv, .id = "experiment") |>
+    dplyr::mutate(
+      oxygen = factor(oxygen, levels = c("21%", "0.5%", "0.2%"), ordered = TRUE),
+      treatment = factor(treatment, levels = c("None", "DMSO", "BAY"), ordered = TRUE)
+    )
+}
+
+plot_blot <- function(blot_image){
+  blot_image <- magick::image_read(blot_image)
+
+  ggplot2::ggplot() +
+    ggpubr::background_image(blot_image) +
+    ggplot2::coord_fixed() +
+    wmo::theme_wmo(
+      base_family = "Calibri",
+      base_size = 8
+    ) +
+    ggplot2::theme(
+      plot.margin = ggplot2::margin(5, 5, 5, 5),
+      panel.border = ggplot2::element_blank(),
+      plot.tag = ggplot2::element_text(face = "bold")
+    )
+}
+
+normalize_densities <- function(blot_raw){
+  blot_raw |>
+    dplyr::filter(.data$time < 96) |>
+    dplyr::filter(!(experiment == "lf_05-bay" & gel %in% c("b", "e", "f"))) |>
+    tidyr::pivot_longer(
+      blot:tidyselect::last_col(),
+      names_to = "protein",
+      values_to = "value",
+      values_drop_na = TRUE
+    ) |>
+    dplyr::group_by(.data$experiment, .data$gel, .data$protein) |>
+    dplyr::mutate(norm = value / mean(value, na.rm = TRUE)) |>
+    dplyr::group_by(dplyr::across(.data$experiment:.data$time)) |>
+    dplyr::mutate(density = norm / norm[protein == "blot"]) |>
+    dplyr::filter(protein != "blot") |>
+    dplyr::group_by(.data$experiment, .data$protein) |>
+    dplyr::mutate(
+      fold_change = density /
+        mean(density[oxygen == min(oxygen) & treatment %in% c("None", "DMSO") & time == min(time)]),
[…]

### 2022-06-06 _targets.R

@@ -29,11 +29,11 @@ tar_option_set(
   format = "qs"
 )
 
-# list of targets ---------------------------------------------------------
+# targets -----------------------------------------------------------------
 
 list(
 
-  # dna per cell ------------------------------------------------------------
+  # dna curve ---------------------------------------------------------------
 
   tar_target(
     dna_per_cell_file,
@@ -62,7 +62,7 @@ list(
     output_dir = system.file("analysis/pdfs", package = "Copeland.2022.hypoxia.flux")
   ),
 
-  # extracellular fluxes ----------------------------------------------------
+  # fluxes ------------------------------------------------------------------
 
   tar_target(
     fluxes_meta_files,
@@ -189,7 +189,7 @@ list(
     output_dir = system.file("analysis/pdfs", package = "Copeland.2022.hypoxia.flux")
   ),
 
-  # q bias correction -------------------------------------------------------
+  # qbias -------------------------------------------------------------------
 
   tar_target(
     qbias_files,
@@ -328,7 +328,23 @@ list(
     clean_viability(viability_file)
   ),
 
-  # write manuscript --------------------------------------------------------
+  # blots -------------------------------------------------------------------
+
+  tar_target(
+    blot_files,
+    path_to_data("immunoblots"),
+    format = "file"
+  ),
+  tar_target(
+    blot_raw,
+    read_data(blot_files)
+  ),
+  tar_target(
+    blot_norm,
+    normalize_densities(blot_raw)
+  ),
+
+  # manuscript --------------------------------------------------------------
 
   # tar_target(
   #   template,

### 2022-06-06 R/1_expression.R
add mrna

@@ -60,3 +60,40 @@ normalize_densities <- function(blot_raw){
     wmo::remove_nested_outliers(fold_change, remove = TRUE) |>
     dplyr::relocate(group, .after = treatment)
 }
+
+normalize_qpcr <- function(raw_mrna){
+  raw_mrna |>
+    dplyr::mutate(gene = tolower(gene)) |>
+    dplyr::group_by(dplyr::across(c(experiment:gene))) |>
+    dplyr::summarize(ct = mean(ct, na.rm = TRUE)) |>
+    tidyr::pivot_wider(names_from = gene, values_from = ct) |>
+    dplyr::ungroup() %>%
+    dplyr::mutate(
+      dplyr::across(-c(experiment:time, actin), ~ . - actin)
+    ) |>
+    dplyr::select(-actin) |>
+    tidyr::pivot_longer(
+      -c(experiment:time),
+      names_to = "protein",
+      values_to = "dct"
+    ) |>
+    dplyr::filter(!is.na(dct)) |>
+    dplyr::group_by(.data$experiment, .data$protein) |>
+    dplyr::mutate(
+      ddct = dct - mean(
+        dct[oxygen == min(oxygen) & treatment %in% c("None", "DMSO") & time == 0]
+      ),
+      fold_change = 2 ^ -ddct,
+      group = dplyr::case_when(
+        experiment %in% c("lf_02", "lf_05", "lf_bay", "pasmc_05") & treatment == "None" & oxygen == "21%" ~ "21%",
+        experiment %in% c("lf_02", "lf_05", "lf_bay", "pasmc_05") & treatment == "DMSO" ~ "DMSO",
+        experiment %in% c("lf_02", "lf_05", "lf_bay", "pasmc_05") & treatment == "BAY" ~ "BAY",
+        experiment %in% c("lf_02", "lf_05", "lf_bay", "pasmc_05") & oxygen == "0.5%" ~ "0.5%",
+        experiment %in% c("lf_02", "lf_05", "lf_bay", "pasmc_05") & oxygen == "0.2%" ~ "0.2%",
+      ),
+      group = factor(group, levels = c("21%", "0.5%", "0.2%", "DMSO", "BAY"))
+    ) |>
+    dplyr::group_by(experiment, oxygen, treatment, group, time, protein) |>
+    wmo::remove_nested_outliers(fold_change, remove = TRUE) |>
+    dplyr::relocate(group, .after = treatment)
+}

### 2022-06-06 _targets.R

@@ -344,6 +344,22 @@ list(
     normalize_densities(blot_raw)
   ),
 
+  # mrna --------------------------------------------------------------------
+
+  tar_target(
+    mrna_files,
+    path_to_data("mrna"),
+    format = "file"
+  ),
+  tar_target(
+    mrna_raw,
+    read_data(mrna_files)
+  ),
+  tar_target(
+    mrna_norm,
+    normalize_qpcr(mrna_raw)
+  ),
+
   # manuscript --------------------------------------------------------------
 
   # tar_target(

### 2022-06-06 R/1_model.R
analyze model results

@@ -1,0 +1,106 @@
+# 1_model.R
+
+clean_model_fluxes <- function(map_flux_files, model_reactions){
+  pathways <-
+    model_reactions |>
+    dplyr::select(-equation) |>
+    dplyr::add_row(name = "BIOMASS", pathway = factor("biomass"), index = 0)
+
+  map_flux_files[stringr::str_detect(map_flux_files, "(lf|pasmc)_.*_model\\.csv")] %>%
+    rlang::set_names(stringr::str_extract(basename(.), pattern = ".*(?=_model\\.csv)")) |>
+    purrr::map_dfr(readr::read_csv, .id = "experiment") |>
+    tidyr::separate(
+      experiment,
+      c("cell_type", "treatment"),
+      sep = "_"
+    ) |>
+    dplyr::mutate(
+      treatment = dplyr::case_when(
+        treatment == "21" ~ "21%",
+        treatment == "05" ~ "0.5%",
+        treatment == "dmso" ~ "DMSO",
+        treatment == "bay" ~ "BAY"
+      ),
+      treatment = factor(treatment, levels = c("21%", "0.5%", "DMSO", "BAY"))
+    ) |>
+
+    # identify net and exchange fluxes
+    tidyr::separate(id, c("id", "type"), sep = " ", fill = "right") |>
+    dplyr::mutate(type = replace(type, is.na(type), "net")) |>
+    dplyr::select(-se) |>
+
+    # add pathway info
+    dplyr::left_join(pathways, by = c("id" = "name")) |>
+    dplyr::select(
+      cell_type,
+      treatment,
+      pathway,
+      index,
+      id,
+      type,
+      equation,
+      flux = value,
+      lb,
+      ub
+    ) |>
+    dplyr::group_by(cell_type, treatment) |>
+    dplyr::arrange(treatment, pathway)
+}
+
+calculate_flux_differences <- function(map_fluxes, cell, control, experiment){
+  map_fluxes |>
+    dplyr::filter(cell_type == cell) |>
+    dplyr::filter(treatment %in% c(control, experiment)) |>
+    dplyr::mutate(
+      treatment = dplyr::case_when(
+        treatment == control ~ "ctl",
+        treatment == experiment ~ "exp"
+      )
+    ) |>
+    tidyr::pivot_wider(names_from = treatment, values_from = c(flux, lb, ub)) |>
[…]

### 2022-06-06 _targets.R

@@ -360,6 +360,27 @@ list(
     normalize_qpcr(mrna_raw)
   ),
 
+  # model -------------------------------------------------------------------
+
+  tar_target(
+    map_flux_files,
+    path_to_data("model"),
+    format = "file"
+  ),
+  tar_target(
+    map_fluxes,
+    clean_model_fluxes(map_flux_files, model_reactions)
+  ),
+  tar_target(
+    map_flux_differences,
+    assemble_flux_differences(map_fluxes)
+  ),
+  tar_render(
+    map_flux_difference_report,
+    path = path_to_reports("flux-differences.Rmd"),
+    output_dir = system.file("analysis/pdfs", package = "Copeland.2022.hypoxia.flux")
+  ),
+
   # manuscript --------------------------------------------------------------
 
   # tar_target(

### 2022-06-06 R/1_nad.R
add nad(h) analysis

@@ -1,0 +1,151 @@
+# nad.R
+
+clean_nad <- function(nad_data){
+  nad_data |>
+    dplyr::bind_rows(.id = "metabolite") |>
+    dplyr::rename(rep = replicate) |>
+    clean_technical_replicates() |>
+    tidyr::separate(.data$experiment, c(NA, "date"), "_")
+}
+
+finalize_nad <- function(nad_interp, cells_per_dna){
+
+  x <-
+    dplyr::filter(cells_per_dna, cell_type == "lf" & volume == 200) |>
+    dplyr::pull(slope)
+
+  nad_interp |>
+    dplyr::group_by(metabolite, date, oxygen, treatment, nucleotide) |>
+    dplyr::summarise(conc = mean(conc)) |>
+    dplyr::mutate(
+      conc = dplyr::case_when(
+        metabolite == "dna" ~ conc * x,
+        metabolite == "nad" ~ conc * 720
+      ),
+      nucleotide = replace(nucleotide, is.na(nucleotide), "Count")
+    ) |>
+    tidyr::pivot_wider(-c(metabolite), names_from = nucleotide, values_from = conc) |>
+    dplyr::mutate(
+      oxygen = factor(oxygen, levels = c("21%", "0.5%")),
+      treatment = factor(treatment, levels = c("none", "DMSO", "BAY")),
+      treatment = forcats::fct_recode(treatment, "None" = "none"),
+      Ratio = NADH/NAD,
+      dplyr::across(tidyselect::contains("NAD"), ~. / Count * 1000)
+    ) |>
+    dplyr::select(-Count) |>
+    tidyr::pivot_longer(c(NAD, NADH, Ratio), names_to = "measurement", values_to = "value") |>
+    dplyr::arrange(measurement, oxygen, treatment)
+
+}
+
+annot_nad <- function(nad_final){
+  nad_final |>
+    dplyr::filter(treatment != "None") |>
+    dplyr::group_by(measurement) |>
+    tidyr::nest() |>
+    dplyr::mutate(
+      m = purrr::map(data, ~lmerTest::lmer(value ~ treatment * oxygen + (1 | date), data = .x)),
+      s = purrr::map(
+        m,
+        ~emmeans::emmeans(.x, ~ treatment * oxygen) |>
+          pairs(simple = "each", combine = TRUE) |>
+          broom::tidy()
+      )
+    ) |>
+    tidyr::unnest(c(s)) |>
[…]

### 2022-06-06 _targets.R

@@ -381,6 +381,42 @@ list(
     output_dir = system.file("analysis/pdfs", package = "Copeland.2022.hypoxia.flux")
   ),
 
+  # nad ---------------------------------------------------------------------
+
+  tar_target(
+    nad_files,
+    path_to_data("nad-assay_.*\\.xlsx"),
+    format = "file"
+  ),
+  tar_target(
+    nad_data,
+    assemble_flux_data(nad_files)
+  ),
+  tar_target(
+    nad_raw,
+    clean_nad(nad_data)
+  ),
+  tar_target(
+    nad_conc_std,
+    make_std_curves(nad_raw, fo = ~MASS::rlm(value ~ poly(conc, 2, raw = TRUE), data = .x, , maxit = 1000))
+  ),
+  tar_target(
+    nad_conc_std_plots,
+    print_plots(nad_conc_std$plots, nad_conc_std$title, "nad/01_standard_curves"),
+    format = "file"
+  ),
+  tar_target(
+    nad_interp,
+    interp_data(nad_raw, nad_conc_std)
+  ),
+  tar_target(
+    nad_final,
+    finalize_nad(nad_interp, cells_per_dna)
+  ),
+  tar_target(
+    nad_annot,
+    annot_nad(nad_final)
+  ),
   # manuscript --------------------------------------------------------------
 
   # tar_target(

### 2022-06-08 analysis/flux-differences.Rmd
start rnaseq data analysis and update tfea approach

---
title: "Model Analysis"
subtitle: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::pdf_document2:
    number_sections: true
    toc: true
    toc_depth: 1
    df_print: kable
    dev: cairo_pdf
    latex_engine: xelatex
geometry: margin = 0.5in
fontsize: 11pt
mainfont: Calibri
---

```{r setup, include = FALSE}
# load libraries
suppressPackageStartupMessages({
  devtools::load_all()
  library(tidyverse)
  library(wmo)
  library(targets)
})

# resolve conflicts
conflicted::conflict_prefer("filter", "dplyr")

# set global chunk options
knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center",
  message = FALSE,
  warning = FALSE,
  out.width = "49%"
)

options(knitr.table.format = function() {
  if (knitr::is_latex_output()) 
    "latex" else "html"
})

theme_set(theme_wmo(base_family = "Calibri"))
```

```{r}
withr::with_dir(here::here(), {
  model_fluxes <- tar_read(model_fluxes)
  flux_differences <- tar_read(map_flux_differences)
})
```

\newpage

# Overview

In this document, we will analyze the output from the metabolic flux maps.

```{r}
usethis::use_data(model_fluxes, overwrite = TRUE)
usethis::use_data(flux_differences, overwrite = TRUE)
```

# Analysis

## Significant Differences

Identify reactions that are significantly different based on whether the confidence intervals overlap. This is more conservative than p < 0.05.

```{r}
calculate_ratio <- function(norm) {
  flux_differences %>% 
    filter(normalization == norm) %>% 
    # filter(!is.na(ratio)) %>% 
    select(-c(normalization, equation, ctl, contains(c("flux", "lb", "ub")))) %>% 
    pivot_wider(names_from = exp, values_from = ratio)
}

print_flux_table <- function(df, caption) {
  df %>% 
    mutate(across(c(`0.5%`, BAY), ~as.character(signif(., 3)))) %>% 
    kableExtra::kable(
      caption = caption, 
      col.names = c("Cell type", "Pathway", "Index", "ID", "Type", "0.5%", "BAY"), 
      align = "clccccc", 
      escape = TRUE, 
      digits = 2, 
      booktabs = TRUE, 
      longtable = TRUE, 
      linesep = ""
    ) %>% 
    kableExtra::kable_styling(font_size = 9)
}
```

```{r}
calculate_ratio("none") %>% 
  print_flux_table("Significant Flux Differences")
```

Do any of these differences have a negative ratio?

```{r}
calculate_ratio("none") %>% 
  filter(`0.5%` < 0 | BAY < 0) %>% 
  print_flux_table("Reversed fluxes")
```

For both of these models, some of the PPP fluxes reverse.

## Normalized differences

It doesn't look like there is any dramatic reprogramming of the intracellular metabolic pathways secondary to hypoxia or BAY treatment. In other words, most of the differences are related to absolute rates rather than differential usage. Another way to look at this might be to normalize the fluxes to glucose uptake or biomass fluxes. This would show all fluxes in terms of total glucose use or growth rate, for example.

```{r}
calculate_ratio("growth") %>% 
  print_flux_table("Significant differences in growth-normalized fluxes")
```

```{r}
calculate_ratio("glucose") %>% 
  print_flux_table("Significant differences in glucose-normalized fluxes")
```

### 2022-06-08 analysis/rnaseq.Rmd
start rnaseq data analysis and update tfea approach

---
title: "RNA-seq Analysis"
subtitle: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::pdf_document2:
    number_sections: true
    toc: true
    toc_depth: 1
    df_print: kable
    dev: cairo_pdf
    latex_engine: xelatex
geometry: margin = 0.5in
fontsize: 11pt
mainfont: Calibri
---

```{r setup, include = FALSE}
# load libraries
suppressPackageStartupMessages({
  devtools::load_all()
  library(tidyverse)
  library(wmo)
  library(targets)
})

# resolve conflicts
conflicted::conflict_prefer("filter", "dplyr")

# set global chunk options
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE, 
  warning = FALSE, 
  fig.align = "center", 
  out.width = "49%"
)

options(knitr.table.format = function() {
  if (knitr::is_latex_output()) 
    "latex" else "html"
})

theme_set(theme_wmo(base_family = "Calibri"))
```

```{r}
withr::with_dir(here::here(), {
  pca <- tar_read(rnaseq_pca)
  deg <- tar_read(deg_int)
  results <- tar_read(deg_int)
  gsea_res <- tar_read(gsea_int)
})
```

\newpage

# Overview

We observed that proliferating primary cells exposed to hypoxia do not increase glucose uptake and lactate efflux despite up-regulation of glucose transporters and glycolytic genes. When these cells are treated with the prolyl hydroxylase inhibitor molidustat in normoxia, the expected increases in glycolytic flux are observed. Interestingly, when molidustat-treated cells are cultured in hypoxia, hypoxia blocks molidustat-mediated increases in glycolysis. In an effort to identify the mechanism mediating this effect, we performed RNA-seq on lung fibroblasts treated for three days with 0.5% oxygen or molidustat (10 μM) with 21% and DMSO (0.1%) controls. RNA was collected and purified and sent for RNA-seq analysis.

# Analysis

We will use the `DESeq2` package to identify differentially expresssed genes. We will use the model `~ experiment + oxygen * treatment`.

## PCA

```{r}
pca
```

## Differential Expression

In this analysis, will identify those genes where the effects of hypoxia and molidustat are different, which will hopefully point to mechanisms by which hypoxia may regulate cell metabolism differently than PHD2 inhibition.

```{r}
deg %>% 
  dplyr::select(row, symbol, description, log2fc = log2FoldChange, padj) %>% 
  head(20) %>% 
  my_kable()
```

### 2022-06-08 R/1_rnaseq.R

@@ -1,0 +1,518 @@
+# 1_rnaseq.R
+
+count_rnaseq <- function(){
+  dds <- rnaseq.lf.hypoxia.molidustat::lf_hyp_bay_rnaseq
+
+  SummarizedExperiment::colData(dds) <-
+    SummarizedExperiment::colData(dds) |>
+    tibble::as_tibble() |>
+    dplyr::mutate(
+      experiment = factor(experiment),
+      oxygen = factor(oxygen, levels = c("21%", "0.5%"), labels = c("N", "H")),
+      treatment = factor(treatment, labels = c("DMSO", "BAY")),
+      group = stringr::str_c(oxygen, treatment, sep = "."),
+      group = factor(group, levels = c("N.DMSO", "H.DMSO", "N.BAY", "H.BAY"))
+    ) |>
+    as("DataFrame")
+
+  rownames(SummarizedExperiment::colData(dds)) <- SummarizedExperiment::colData(dds)$id
+
+  design <- ~ experiment + group
+
+  dds <-
+    DESeq2::DESeqDataSet(
+      dds,
+      design = design
+    )
+
+  keep <- rowSums(DESeq2::counts(dds)) > 1
+  dds <- dds[keep, ]
+  DESeq2::DESeq(dds)
+}
+
+vst_rnaseq <- function(dds){
+  vsd <- DESeq2::vst(dds, blind = FALSE)
+  SummarizedExperiment::assay(vsd) <-
+    limma::removeBatchEffect(SummarizedExperiment::assay(vsd), vsd$experiment)
+
+  DESeq2::plotPCA(
+    vsd,
+    intgroup = "group",
+    returnData = TRUE
+  ) |>
+    dplyr::mutate(
+      label = dplyr::case_when(
+        group == "N.DMSO" ~ "21%\nDMSO",
+        group == "H.DMSO" ~ "0.5%\nDMSO",
+        group == "N.BAY" ~ "21%\nBAY",
+        group == "H.BAY" ~ "0.5%\nBAY"
+      )
+    )
+}
+
+plot_rnaseq_pca <- function(pca_data){
+  percent_variance <- round(100 * attr(pca_data, "percentVar"))
+
+  pair_clrs <- c(
+    N.DMSO = "#b2df8a",
+    H.DMSO = "#33a02c",
+    N.BAY = "#cab2d6",
+    H.BAY = "#6a3d9a"
+  )
+
+  ggplot2::ggplot(pca_data) +
[…]

### 2022-06-08 _targets.R

@@ -23,7 +23,7 @@ future::plan(future.callr::callr(workers = future::availableCores() - 1))
 
 # target-specific options
 tar_option_set(
-  packages = c("tidyverse", "patchwork"),
+  packages = c("tidyverse", "patchwork", "scales"),
   # packages = c("tidyverse", "patchwork", "xcms"),
   # imports = c("rnaseq.lf.hypoxia.molidustat"),
   format = "qs"
@@ -417,6 +417,70 @@ list(
     nad_annot,
     annot_nad(nad_final)
   ),
+
+  # rnaseq ------------------------------------------------------------------
+
+  tar_target(
+    dds,
+    count_rnaseq()
+  ),
+  tar_target(
+    pca_data,
+    vst_rnaseq(dds)
+  ),
+  tar_target(
+    rnaseq_pca,
+    plot_rnaseq_pca(pca_data)
+  ),
+  tar_target(
+    hallmark_pathways,
+    get_msigdb_pathways(category = "H")
+  ),
+  tar_target(
+    tfea,
+    run_tfea(dds)
+  ),
+  tar_map(
+    values = list(
+      names = c("hyp", "bay", "hyp_bay", "int"),
+      comp = list(
+        "h.dmso - n.dmso",
+        "n.bay - n.dmso",
+        "h.bay - n.bay",
+        "(h.dmso - n.dmso) - (n.bay - n.dmso)"
+      ),
+      xlab = c(
+        "Hypoxia/Normoxia",
+        "BAY/DMSO",
+        "Hypoxia/Normoxia",
+        "ΔHypoxia/ΔBAY"
+      )
+    ),
+    names = names,
+    tar_target(
+      deg,
+      identify_deg(dds, rlang::expr(comp))
+    ),
+    tar_target(
+      rnaseq_vol,
+      plot_rnaseq_volcano(deg, xlab)
+    ),
+    tar_target(
+      gsea,
+      run_gsea(deg, hallmark_pathways)
+    ),
+    tar_target(
+      tfea_res,
+      index_tfea(tfea, names)
+    ),
+    NULL
+  ),
+  tar_render(
+    rnaseq_report,
+    path = path_to_reports("rnaseq.Rmd"),
+    output_dir = system.file("analysis/pdfs", package = "Copeland.2022.hypoxia.flux")
+  ),
+
   # manuscript --------------------------------------------------------------
 
   # tar_target(

### 2022-06-09 R/1_rnaseq.R
split tfea analysis and limma to different functions

@@ -146,54 +146,6 @@ identify_deg <- function(dds, comp){
     dplyr::arrange(padj)
 }
 
-find_rnaseq_overlap <- function(dds){
-  bay <-
-    identify_deg(dds, expr(n.bay - n.dmso)) |>
-    dplyr::filter(padj < 0.05)
-  hyp <- identify_deg(dds, expr(h.dmso - n.dmso)) |>
-    dplyr::filter(padj < 0.05)
-
-  nm <- rownames(SummarizedExperiment::assay(dds))
-  bay_deg <- nm %in% bay$row
-  hyp_deg <- nm %in% hyp$row
-  tibble::tibble(nm, hyp_deg, bay_deg)
-}
-
-find_gsea_overlap <- function(a, b, src = "HALLMARK"){
-  dplyr::full_join(a, b, by = c("source", "pathway")) |>
-    dplyr::select(source, pathway, padj.x, padj.y) |>
-    dplyr::filter(source %in% src) |>
-    dplyr::mutate(
-      hyp_deg = padj.x < 0.05,
-      bay_deg = padj.y < 0.05,
-      dplyr::across(tidyselect::matches("_deg"), tidyr::replace_na, FALSE)
-    )
-}
-
-plot_rnaseq_venn <- function(df, title){
-  ggplot2::ggplot(df) +
-    ggplot2::aes(A = hyp_deg, B = bay_deg) +
-    ggvenn::geom_venn(
-      set_names = c("0.5%", "BAY"),
-      digits = 0,
-      show_percentage = TRUE,
-      fill_color = clrs[c(2, 4)],
-      fill_alpha = 0.5,
-      stroke_size = 0.25,
-      set_name_size = 6/ggplot2::.pt,
-      text_size = 5/ggplot2::.pt
-    ) +
-    ggplot2::labs(title = title) +
-    theme_plots() +
-    ggplot2::coord_fixed(clip = "off") +
-    ggplot2::theme(
-      panel.border = ggplot2::element_blank(),
-      axis.text = ggplot2::element_blank(),
-      axis.ticks = ggplot2::element_blank(),
-      plot.title = ggplot2::element_text(hjust = 0.5)
-    )
-}
-
 plot_rnaseq_volcano <- function(results, gois = NULL, xlab = NULL){
   df <-
     tibble::as_tibble(results)
@@ -441,6 +393,9 @@ plot_gsea <- function(rnaseq_gsea, sources, lbls, vals){
 }
 
 run_tfea <- function(dds){
+  # vst if using scale as method
[…]

### 2022-06-09 R/1_utils.R

@@ -160,9 +160,10 @@ print_plots <- function(
 annot_p <- function(num) dplyr::if_else(num < 0.05, "*", NA_character_)
 
 my_kable <- function(data, ...) {
-  kableExtra::kable(data, booktabs = TRUE, linesep = "", ...) %>%
+  kableExtra::kable(data, booktabs = TRUE, linesep = "", ...) |>
     kableExtra::kable_styling(
       latex_options = c("hold_position"),
       font_size = 9
     )
 }
+

### 2022-06-09 _targets.R

@@ -440,6 +440,10 @@ list(
     tfea,
     run_tfea(dds)
   ),
+  tar_target(
+    tfea_fit,
+    fit_tfea(dds, tfea)
+  ),
   tar_map(
     values = list(
       names = c("hyp", "bay", "hyp_bay", "int"),
@@ -471,7 +475,7 @@ list(
     ),
     tar_target(
       tfea_res,
-      index_tfea(tfea, names)
+      index_tfea(tfea_fit, names)
     ),
     NULL
   ),
@@ -480,6 +484,102 @@ list(
     path = path_to_reports("rnaseq.Rmd"),
     output_dir = system.file("analysis/pdfs", package = "Copeland.2022.hypoxia.flux")
   ),
+
+  # metab -------------------------------------------------------------------
+
+  tar_target(
+    metab_tar_files,
+    path_to_data("lf_05-bay_metabolomics-targeted.xlsx"),
+    format = "file"
+  ),
+  tar_target(
+    metab_tar_raw,
+    format_metab_tar(metab_tar_files)
+  ),
+  tar_target(
+    metab_tar_clean,
+    remove_missing_metab(metab_tar_raw) %>%
+      correct_drift() %>%
+      quality_control() %>%
+      impute_missing() %>%
+      pqn() %>%
+      log_transform()
+  ),
+  tar_target(
+    metab_tar_pca,
+    plot_metab_pca(metab_tar_clean)
+  ),
+  tar_target(
+    metab_tar_limma,
+    fit_metab_limma(metab_tar_clean)
+  ),
+  tar_map(
+    values = list(
+      names = list("hyp", "bay", "hyp_bay", "int"),
+      colors = list(clrs[2:1], clrs[4:3], clrs[2:1], clrs[c(2, 4)]),
+      xlab = list("Hypoxia/Normoxia", "BAY/DMSO", "Hypoxia/Normoxia", "ΔHypoxia/ΔBAY")
+    ),
+    names = names,
+    tar_target(
+      metab_tar_res,
+      index_metab_limma(metab_tar_clean, metab_tar_limma, names)
+    ),
+    tar_target(
+      metab_tar_vol,
+      plot_metab_volcano(metab_tar_res, colors = colors, xlab = xlab)
+    ),
+    NULL
+  ),
+  # tar_target(
+  #   metab_venn,
+  #   plot_metab_venn(metab_hyp, metab_bay)
+  # ),
+  # tar_target(
[…]

### 2022-06-09 analysis/metabolomics-targeted.Rmd
targeted metabolomics analysis

--- 2022-06-08 analysis/rnaseq.Rmd
@@
 ---
-title: "RNA-seq Analysis"
+title: "Targeted Metabolomics"
 subtitle: "`r format(Sys.time(), '%B %d, %Y')`"
@@
 withr::with_dir(here::here(), {
-  pca <- tar_read(rnaseq_pca)
-  deg <- tar_read(deg_int)
-  results <- tar_read(deg_int)
-  gsea_res <- tar_read(gsea_int)
+  raw <- tar_read(metab_tar_raw)
+  clean <- tar_read(metab_tar_clean)
+  metab_pca <- tar_read(metab_tar_pca)
+  metab_limma <- tar_read(metab_tar_limma)
+  tt <- tar_read(metab_tar_res_int)
+  volcano <- tar_read(metab_tar_vol_int)
 })
@@
 Interestingly, when molidustat-treated cells are cultured in hypoxia, hypoxia blocks molidustat-mediated increases in glycolysis.
-In an effort to identify the mechanism mediating this effect, we performed RNA-seq on lung fibroblasts treated for three days with 0.5% oxygen or molidustat (10 μM) with 21% and DMSO (0.1%) controls.
-RNA was collected and purified and sent for RNA-seq analysis.
+In an effort to identify the mechanism mediating this effect, we performed metabolomics on lung fibroblasts treated for three days with 0.5% oxygen or molidustat (10 μM) with 21% and DMSO (0.1%) controls.
 
-# Analysis
-
-We will use the `DESeq2` package to identify differentially expresssed genes. We will use the model `~ experiment + oxygen * treatment`.
+# Data Processing
 
@@
 ```{r}
-pca
+pcaMethods::pca(t(SummarizedExperiment::assay(raw)), scale = "uv", center = TRUE) %>% 
+  pcaMethods::scores() %>% 
+  merge(SummarizedExperiment::colData(raw), by = 0) %>% 
+  # filter(type %nin% c("qc", "blank")) %>% 
+  ggplot() +
+  aes(
+    x = PC1, 
+    y = PC2, 
+    shape = type,
+    color = interaction(oxygen, treatment, sep = "|")
+  ) + 
+  geom_point() +
+  labs(
+    color = "Group", 
+    shape = "Type"
+  )
+```
+## Cleaning
+
+1. drift correction
+1. remove low quality features (QC RSD > 0.2, dispersion ratio > 0.4)
+1. impute missing values with random forest
+1. probabilistic quotient normalization
+1. batch correction
+
+```{r}
+metab_pca
 ```
 
-## Differential Expression
-
-In this analysis, will identify those genes where the effects of hypoxia and molidustat are different, which will hopefully point to mechanisms by which hypoxia may regulate cell metabolism differently than PHD2 inhibition.
+## Differentially regulated metabolites
 
 ```{r}
-deg %>% 
-  dplyr::select(row, symbol, description, log2fc = log2FoldChange, padj) %>% 
-  head(20) %>% 
-  my_kable()
+volcano
 ```

### 2022-06-09 R/1_figures.R

@@ -1,6 +1,8 @@
 # 1_figures.R
 
 # plot setup
+clrs <- c(RColorBrewer::brewer.pal(4, "Set1")[1:4], "#08306b", RColorBrewer::brewer.pal(9, "Set1")[9:8])
+names(clrs) <- c("21%", "0.5%", "DMSO", "BAY", "0.2%", "siCTL", "siMYC")
 
 theme_plots <- function() {
   list(

### 2022-06-09 R/1_metab.R

@@ -1,0 +1,568 @@
+# 1_metab.R
+
+new_tbl_se <- function(
+    tbl,
+    a_data,
+    f_names,
+    f_data = NULL,
+    s_names,
+    s_data = NULL
+){
+  structure(
+    tbl,
+    class = c("tbl_se", class(tbl)),
+    a_data = a_data,
+    f_names = f_names,
+    s_names = s_names,
+    f_data = f_data,
+    s_data = s_data
+  )
+}
+
+tbl_to_se <- function(tbl_se, assay_name){
+  assay_data <-
+    tbl_se |>
+    dplyr::select(
+      attr(tbl_se, "f_names"),
+      attr(tbl_se, "s_names"),
+      attr(tbl_se, "a_data")
+    ) |>
+    tidyr::pivot_wider(
+      names_from = attr(tbl_se, "s_names"),
+      values_from = attr(tbl_se, "a_data")
+    ) |>
+    tibble::column_to_rownames(attr(tbl_se, "f_names"))
+
+  feature_data <-
+    tbl_se |>
+    dplyr::select(
+      attr(tbl_se, "f_names"),
+      attr(tbl_se, "f_data")
+    ) |>
+    dplyr::group_by(!!rlang::sym(attr(tbl_se, "f_names"))) |>
+    dplyr::summarise(
+      metabolite = unique(metabolite),
+      mz = mean(mz, na.rm = TRUE),
+      mz_min = min(mz, na.rm = TRUE),
+      mz_max = max(mz, na.rm = TRUE),
+      rt = mean(rt, na.rm = TRUE),
+      rt_min = min(rt, na.rm = TRUE),
+      rt_max = max(rt, na.rm = TRUE)
+    ) |>
+    tibble::column_to_rownames(attr(tbl_se, "f_names")) %>%
+    .[match(rownames(assay_data), rownames(.)), ]
+
+  sample_data <-
+    tbl_se |>
+    dplyr::select(
+      attr(tbl_se, "s_names"),
+      attr(tbl_se, "s_data")
+    ) |>
+    dplyr::distinct() |>
+    tibble::column_to_rownames(attr(tbl_se, "s_names")) %>%
+    .[match(colnames(assay_data), rownames(.)), ]
+
+  SummarizedExperiment::SummarizedExperiment(
+    assays = assay_data,
+    rowData = feature_data,
+    colData = sample_data
+  )
+}
+
+format_metab_tar <- function(files){
[…]

### 2022-06-09 _targets.R

@@ -498,12 +498,13 @@ list(
   ),
   tar_target(
     metab_tar_clean,
-    remove_missing_metab(metab_tar_raw) %>%
-      correct_drift() %>%
-      quality_control() %>%
-      impute_missing() %>%
-      pqn() %>%
-      log_transform()
+    remove_missing_metab(metab_tar_raw) |>
+      correct_drift() |>
+      quality_control() |>
+      impute_missing() |>
+      pqn() |>
+      log_transform() |>
+      annot_metabs()
   ),
   tar_target(
     metab_tar_pca,
@@ -512,6 +513,15 @@ list(
   tar_target(
     metab_tar_limma,
     fit_metab_limma(metab_tar_clean)
+  ),
+  tar_target(
+    metab_pathways_file,
+    path_to_data("metabolites.tab"),
+    format = "file"
+  ),
+  tar_target(
+    metab_pathways,
+    read_metab_pathways(metab_pathways_file)
   ),
   tar_map(
     values = list(
@@ -528,58 +538,21 @@ list(
       metab_tar_vol,
       plot_metab_volcano(metab_tar_res, colors = colors, xlab = xlab)
     ),
+    tar_target(
+      metab_tar_msea,
+      run_msea(metab_tar_res, metab_pathways)
+    ),
     NULL
   ),
   # tar_target(
   #   metab_venn,
   #   plot_metab_venn(metab_hyp, metab_bay)
   # ),
-  # tar_target(
-  #   metab_moi,
-  #   plot_mois(metab_targeted_clean, c("GAP", "2-hydroxyglutarate", "aconitate", "taurine", "hydroxyproline", "GABA"))
-  # ),
-  # # tar_target(
-  # #   metab_msea,
-  # #   run_msea(metab_different_differences, metab_pathways)
-  # # ),
-  # # tar_target(
-  # #   metab_msea_hyp,
-  # #   run_msea(metab_hyp, metab_pathways)
-  # # ),
-  # # tar_target(
-  # #   metab_msea_bay,
-  # #   run_msea(metab_bay, metab_pathways)
-  # # ),
-  # tar_target(
-  #   metabolite_pathways_file,
-  #   path_to_data("metabolites.tab"),
-  #   format = "file"
-  # ),
-  # tar_target(
-  #   metabolite_pathways,
-  #   read_pathways(metabolite_pathways_file)
-  # ),
-  # # tar_target(
-  # #   msea_plot,
[…]

### 2022-06-10 R/1_myc.R
add MYC analyses

@@ -1,0 +1,127 @@
+# 1_myc.R
+
+# 1_myc.R
+
+combine_fluxes <- function(growth_rates, fluxes, exp) {
+  growth_rates |>
+    dplyr::filter(experiment == exp) |>
+    dplyr::rename(flux = mu) |>
+    dplyr::select(-X0) |>
+    dplyr::mutate(metabolite = "growth") |>
+    dplyr::bind_rows(dplyr::filter(fluxes, experiment == exp)) |>
+    dplyr::filter(treatment %nin% c("siHIF1A", "siHIF2A")) |>
+    dplyr::mutate(
+      flux = ifelse(
+        experiment == "bay-myc" & metabolite == "lactate" & date %in% c("2021-09-21", "2021-11-01"),
+        flux / 3,
+        flux
+      )
+    ) |>
+    dplyr::group_by(oxygen, treatment, virus) |>
+    wmo::remove_nested_outliers(flux, TRUE)
+}
+
+annot_myc_fluxes <- function(df, intervention) {
+  if (intervention == "treatment") {
+    lvls <- c("siCTL", "siMYC")
+    x <- "oxygen"
+  } else if (intervention == "virus") {
+    lvls <- c("YFP", "MYC")
+    x <- "virus"
+  }
+  fo1 <- as.formula(paste("flux ~ ", x, " * treatment + (1 | date)"))
+  fo2 <- as.formula(paste("pairwise ~ ", x, "* treatment"))
+
+  int <- rlang::sym(intervention)
+
+  df |>
+    dplyr::group_by(metabolite) |>
+    tidyr::nest() |>
+    dplyr::mutate(
+      m = purrr::map(data, ~lmerTest::lmer(fo1, data = .x)),
+      res = purrr::map(m, ~emmeans::emmeans(
+        .x,
+        fo2,
+        simple = "each",
+        adjust = "mvt",
+        combine = TRUE
+      )[["contrasts"]]
+      ),
+      out = purrr::map(res, broom::tidy)
+    ) |>
+    tidyr::unnest(c(out)) |>
+    dplyr::filter(!!int != ".") |>
+    dplyr::select(metabolite, {{ intervention }}, adj.p.value) |>
+    dplyr::mutate(
+      group = factor(!!int, levels = lvls),
+      y_pos = Inf,
+      vjust = 1.5,
+      lab = annot_p(adj.p.value)
+    ) |>
+    dplyr::select(-!!int) |>
+    dplyr::rename(!!int := group)
+}
+
[…]

### 2022-06-10 R/1_cosmos.R
add comsos analyses

@@ -1,0 +1,104 @@
+# 1_cosmos.R
+
+set_carnival_options <- function(){
+  path <- "analysis/carnival"
+  carnival_options <- cosmosR::default_CARNIVAL_options(solver = "cplex")
+  carnival_options$solverPath <- "/Applications/CPLEX_Studio201/cplex/bin/x86-64_osx/cplex"
+  carnival_options$outputFolder <- path
+  carnival_options$workdir <- path
+  carnival_options$threads <- 0
+  carnival_options$clonelog <- -1
+  carnival_options$poolIntensity <- 2
+  carnival_options$mipGap <- 0.2
+  carnival_options
+}
+
+get_cosmos_network <- function(){
+  data(meta_network, package = "cosmosR")
+  meta_network
+}
+
+format_tf <- function(tfea) {
+  tfea |>
+    dplyr::mutate(activity = ifelse(adj.P.Val >= 0.05, 0, sign(t))) |>
+    dplyr::select(tf, activity) |>
+    tibble::deframe()
+}
+
+format_metab <- function(metab) {
+  metab |>
+    dplyr::select(hmdb, t, P.Value) |>
+    dplyr::mutate(activity = dplyr::if_else(P.Value < 0.05, sign(t), 0)) |>
+    dplyr::select(hmdb, activity) |>
+    tibble::deframe() |>
+    cosmosR::prepare_metab_inputs(compartment_codes = c("c", "m"))
+}
+
+format_deg <- function(deg) {
+  deg |>
+    dplyr::filter(!is.na(symbol) & symbol != "") |>
+    dplyr::filter(!is.na(padj)) |>
+    dplyr::select(symbol, stat, padj) |>
+    dplyr::group_by(symbol) |>
+    dplyr::arrange(abs(stat), .by_group = TRUE) |>
+    dplyr::slice_tail(n = 1) |>
+    dplyr::mutate(activity = dplyr::if_else(padj < 0.05, sign(stat), 0)) |>
+    dplyr::select(symbol, activity) |>
+    tibble::deframe()
+}
+
+preprocess <- function(
+    route = c("forward", "reverse"),
+    network,
+    signaling,
+    metabolites,
+    transcripts,
+    carnival_options
+) {
+  carnival_options$timelimit <- 3600
+
+  network_members <- unique(c(network$source, network$target))
+  signaling <- signaling[names(signaling) %in% network_members]
+  metabolites <- metabolites[names(metabolites) %in% network_members]
+
[…]

### 2022-06-10 _targets.R

@@ -554,6 +554,75 @@ list(
     output_dir = system.file("analysis/pdfs", package = "Copeland.2022.hypoxia.flux")
   ),
 
+  # myc ---------------------------------------------------------------------
+
+  tar_map(
+    values = list(
+      names = c("simyc", "oemyc"),
+      exp = list("05-simyc", "bay-myc"),
+      intervention = list("treatment", "virus"),
+      x = rlang::syms(c("treatment", "virus")),
+      y = rlang::syms(c("oxygen", "treatment"))
+    ),
+    names = names,
+    tar_target(
+      myc_fluxes,
+      combine_fluxes(growth_rates, fluxes, exp = exp)
+    ),
+    tar_target(
+      myc_fluxes_annot,
+      annot_myc_fluxes(myc_fluxes, intervention)
+    ),
+    tar_target(
+      myc_growth_plot,
+      plot_myc(myc_fluxes, myc_fluxes_annot, "growth", "Growth rate (/h)", x = x, fill = y)
+    ),
+    tar_target(
+      myc_lactate_plot,
+      plot_myc(myc_fluxes, myc_fluxes_annot, "lactate", "Lactate\n(fmol/cell/h)", x = x, fill = y)
+    ),
+    NULL
+  ),
+
+  # cosmos ------------------------------------------------------------------
+
+  tar_target(
+    carnival_options,
+    set_carnival_options()
+  ),
+  tar_target(
+    cosmos_network,
+    get_cosmos_network()
+  ),
+  tar_target(
+    cosmos_tf,
+    format_tf(tfea_res_int)
+  ),
+  tar_target(
+    cosmos_metab,
+    format_metab(metab_tar_res_int)
+  ),
+  tar_target(
+    cosmos_deg,
+    format_deg(deg_int)
+  ),
+  tar_target(
+    cosmos_prep_forward,
+    preprocess("forward", cosmos_network, cosmos_tf, cosmos_metab, cosmos_deg, carnival_options)
+  ),
+  tar_target(
+    cosmos_prep_reverse,
+    preprocess("reverse", cosmos_network, cosmos_tf, cosmos_metab, cosmos_deg, carnival_options)
+  ),
+  tar_target(
+    cosmos_forward,
+    run_cosmos("forward", cosmos_prep_forward, carnival_options)
+  ),
+  tar_target(
+    cosmos_reverse,
+    run_cosmos("reverse", cosmos_prep_reverse, carnival_options)
+  ),
+
[…]

### 2022-06-19 manuscript/manuscript.Rmd
start working on manuscript

---
title: "MYC overrides HIF to sustain proliferating primary cell metabolism in hypoxia"

author:
- Courtney A. Copeland: 
    institute: bwh, hms
- Benjamin A. Olenchock: 
    institute: bwh, hms
- David R. Ziehr:
    institute: bwh, mgh, hms
- Sarah McGarrity: 
    institute: bwh, hms, ui
- Kevin Leahy: 
    institute: bwh, hms
- Jamey D. Young: 
    institute: vu
- Joseph Loscalzo: 
    institute: bwh, hms
- William M. Oldham:
    institute: bwh, hms
    correspondence: true
    email: woldham@bwh.harvard.edu

institute:
- bwh: Department of Medicine, Brigham and Women's Hospital, Boston, MA 02115, U.S.A.
- hms: Department of Medicine, Harvard Medical School, Boston, MA 02115, U.S.A.
- mgh: Department of Medicine, Massachusetts General Hospital, Boston, MA, 02114 U.S.A.
- ui: Center for Systems Biology, School of Health Sciences, University of Iceland, Reykjavik, Iceland
- vu: Departments of Chemical & Biomolecular Engineering and Molecular Physiology & Biophysics, Vanderbilt University, Nashville, TN 37240, U.S.A.

bibliography: 
- library.json
- packages.bib
csl: cell-metabolism.csl
link-citations: yes

output: 
  bookdown::word_document2:
    reference_docx: !expr system.file("manuscript", "template.docx", package = "Copeland.2022.hypoxia.flux")
    df_print: kable
    fig_caption: true
    number_sections: false
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
      - '--lua-filter=pagebreak.lua'
      # - '--lua-filter=multiple-bibliographies.lua'

compact-title: true
geometry: margin = 0.75in
fontsize: 11pt
mainfont: Calibri
graphics: yes
---

```{r setup, include = FALSE}
# library(Copeland.2022.hypoxia.flux)
devtools::load_all()

knitr::opts_chunk$set(
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE, 
  out.width = "100%", 
  out.extra = ""
)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

knitr::opts_knit$set(kable.force.latex = FALSE)

options(
  tinytex.verbose = TRUE, 
  knitr.graphics.rel_path = FALSE
)

# targets::tar_load(
#   c(
#     m1_figure, 
#   )
# )
```

# Abstract

# Keywords

/ /

\newpage

# Introduction

Cellular responses to hypoxia drive fundamental physiologic and pathologic processes from wound healing and angiogenesis to vascular remodeling and fibrosis [@semenza2012b; @lee2019]. These responses require cells to maintain energetically demanding activities, including proliferation, migration, and macromolecular biosynthesis, even in the face of a limited oxygen supply. Since metazoan cells depend primarily on respiration to meet their energy needs,

Since metazoan cells depend on respiration to meet their energy needs, they must reduce energy consumption and shift energy production away from oxidative phosphorylation when challenged by hypoxia. Cells accomplish this goal through stabilization of the hypoxia-inducible transcription factor 1α (HIF-1α), which activates the transcription of glucose transporters, glycolytic enzymes, lactate dehydrogenase, and pyruvate dehydrogenase kinase, while decreasing the expression of enzymes in the tricarboxylic acid (TCA) cycle and electron transport chain [@semenza2012b; @lee2020b]. Although HIF-1α is constitutively expressed, it is hydroxylated by prolyl hydroxylase enzymes (PHDs) in normoxia and targeted for proteasomal degradation. PHDs are the principal oxygen sensors in metazoan cells [@kaelin2008a]. PHDs are α-ketoglutarate-dependent dioxygenase enzymes that require molecular oxygen for their enzymatic activity. When oxygen tension falls, PHD activity decreases, leading to HIF-1α stabilization and activation of its associated transcriptional program. Overall, the HIF-1 transcriptional program should increase glycolytic capacity and divert glucose-derived pyruvate from oxidative phosphorylation toward lactate fermentation to maintain ATP production.

\newpage

# Results

\newpage

# Discussion

\newpage

# Materials and Methods

## Data Availability

\newpage

# Acknowledgements

# Author Contributions

# Disclosures and competing interests

\newpage

# References

::: {#refs} 
:::

\newpage

# Tables

\newpage

# Figure Legends

(ref:m1) **Caption.** Info.

```{r}
figs <- 
  list.files(
    system.file(
      "manuscript/figures", 
      package = "package"
    ), 
    pattern = "m\\d{1}\\.png"
  ) %>% 
  rlang::set_names(stringr::str_extract(., "^.*(?=\\.png)"))
```

```{r results = "asis"}
out <- 
  purrr::imap(
    figs, 
    ~ cat("Figure ", stringr::str_extract(.y, "\\d"), ": (ref:", .y, ")\n\n", sep = "")
  )
```

\newpage

# Figures

```{r m1, fig.cap = "(ref:m1)"}
knitr::include_graphics(
  system.file(
    "manuscript/figures/m1.png",
    package = "package"
  )
)
```

\newpage

### 2022-06-19 R/1_metab.R
separate pca calculations from plot

@@ -7,7 +7,7 @@ new_tbl_se <- function(
     f_data = NULL,
     s_names,
     s_data = NULL
-){
+) {
   structure(
     tbl,
     class = c("tbl_se", class(tbl)),
@@ -242,7 +242,7 @@ annot_metabs <- function(se){
   se
 }
 
-plot_metab_pca <- function(clean) {
+calc_metab_pca <- function(clean) {
   input <-
     SummarizedExperiment::assay(clean) |>
     t() |>
@@ -254,11 +254,12 @@ plot_metab_pca <- function(clean) {
   design <- model.matrix(~ 0 + group, data = pheno)
   colnames(design) <- stringr::str_extract(colnames(design), "(?<=group).*")
 
-  df <-
-    limma::removeBatchEffect(input, batch = pheno$replicate, design = design) |>
+  limma::removeBatchEffect(input, batch = pheno$replicate, design = design) |>
     t() |>
     pcaMethods::pca(scale = "none", center = TRUE)
+}
 
+plot_metab_pca <- function(clean, df) {
   percent_variance <- round(100 * c(df@R2[[1]], df@R2[[2]]))
 
   pair_clrs <- c(
@@ -557,8 +558,7 @@ run_msea <- function(tt, pathways){
   fgsea::fgsea(
     pathways = pathways,
     stats = stats,
-    sampleSize = 4,
-    minSize = 1,
+    minSize = 3,
     BPPARAM = BiocParallel::bpparam()
   ) %>%
     tibble::as_tibble() %>%

### 2022-06-19 _targets.R

@@ -508,7 +508,11 @@ list(
   ),
   tar_target(
     metab_tar_pca,
-    plot_metab_pca(metab_tar_clean)
+    calc_metab_pca(metab_tar_clean)
+  ),
+  tar_target(
+    metab_tar_pca_plot,
+    plot_metab_pca(metab_tar_pca, metab_tar_clean)
   ),
   tar_target(
     metab_tar_limma,
@@ -627,22 +631,29 @@ list(
 
   # tar_target(
   #   template,
-  #   system.file("manuscript/template.docx", package = "McGarrity.2022.hypoxia.omics"),
+  #   system.file("manuscript/template.docx", package = "Copeland.2022.hypoxia.flux"),
   #   format = "file"
   # ),
   # tar_target(
   #   pkgs,
-  #   system.file("manuscript/packages.bib", package = "McGarrity.2022.hypoxia.omics"),
+  #   system.file("manuscript/packages.bib", package = "Copeland.2022.hypoxia.flux"),
   #   format = "file"
   # ),
   # tar_target(
+  #   bib_file,
+  #   rbbt::bbt_update_bib(
+  #     path = "manuscript/manuscript.Rmd",
+  #     path_bib = "manuscript/library.json"
+  #   )
+  # ),
+  # tar_target(
   #   bib,
-  #   system.file("manuscript/library.json", package = "McGarrity.2022.hypoxia.omics"),
+  #   system.file("manuscript/library.json", package = "Copeland.2022.hypoxia.flux"),
   #   format = "file"
   # ),
   # tar_target(
   #   csl,
-  #   system.file("manuscript/cell-metabolism.csl", package = "McGarrity.2022.hypoxia.omics"),
+  #   system.file("manuscript/cell-metabolism.csl", package = "Copeland.2022.hypoxia.flux"),
   #   format = "file"
   # ),
   # tar_render(

### 2022-07-11 manuscript/manuscript.Rmd
revising manuscript through figure 1

@@
 ---
-title: "MYC overrides HIF to sustain proliferating primary cell metabolism in hypoxia"
+title: "MYC overrides HIF to regulate proliferating primary cell metabolism in hypoxia"
 
@@
 
-knitr::opts_chunk$set(
-  collapse = TRUE,
-  comment = "#>"
+rbbt::bbt_update_bib(
+  path = "manuscript/manuscript.Rmd",
+  path_bib = "manuscript/library.json"
 )
-
-knitr::opts_knit$set(kable.force.latex = FALSE)
 
@@
 
-Cellular responses to hypoxia drive fundamental physiologic and pathologic processes from wound healing and angiogenesis to vascular remodeling and fibrosis [@semenza2012b; @lee2019].
-These responses require cells to maintain energetically demanding activities, including proliferation, migration, and macromolecular biosynthesis, even in the face of a limited oxygen supply.
-Since metazoan cells depend primarily on respiration to meet their energy needs,
+Cellular responses to hypoxia propel many physiologic and pathologic activities from wound healing and angiogenesis to vascular remodeling and fibrosis [@semenza2012b; @lee2019].
+These activities require cells to continue energetically demanding tasks, such as macromolecular biosynthesis and proliferation, despite limited oxygen availability.
+Since respiration is the most efficient way for cells to produce energy, cell metabolism must adapt to meet energetic needs when oxygen supply is limiting.
+Understanding how these metabolic adaptations sustain critical cellular processes in hypoxia is fundamentally important to our understanding of human health and disease.
 
-Since metazoan cells depend on respiration to meet their energy needs, they must reduce energy consumption and shift energy production away from oxidative phosphorylation when challenged by hypoxia.
-Cells accomplish this goal through stabilization of the hypoxia-inducible transcription factor 1α (HIF-1α), which activates the transcription of glucose transporters, glycolytic enzymes, lactate dehydrogenase, and pyruvate dehydrogenase kinase, while decreasing the expression of enzymes in the tricarboxylic acid (TCA) cycle and electron transport chain [@semenza2012b; @lee2020b].
+Cells typically respond to hypoxia by shifting energy production away from respiration and toward glycolysis.
+This response is mediated primarily by stabilization of the hypoxia-inducible transcription factor 1α (HIF-1α).
+HIF-1α activates the transcription of glucose transporters, glycolytic enzymes, lactate dehydrogenase, and pyruvate dehydrogenase kinase, while decreasing the expression of tricarboxylic acid (TCA) cycle and electron transport chain enzymes [@semenza2012b; @lee2020b].
 Although HIF-1α is constitutively expressed, it is hydroxylated by prolyl hydroxylase enzymes (PHDs) in normoxia and targeted for proteasomal degradation.
@@
 When oxygen tension falls, PHD activity decreases, leading to HIF-1α stabilization and activation of its associated transcriptional program.
-Overall, the HIF-1 transcriptional program should increase glycolytic capacity and divert glucose-derived pyruvate from oxidative phosphorylation toward lactate fermentation to maintain ATP production.
+Overall, this transcriptional program should increase glycolytic capacity and divert glucose-derived pyruvate from oxidative phosphorylation toward lactate fermentation to maintain glycolytic ATP production.
+
+In addition to metabolic changes designed to maintain energy supply, hypoxic cells also reduce energy demands through down-regulation of Na^+^/K^+^-ATPase, slowing protein translation, and attenuating cell proliferation [@wheaton2011; @hubbi2015].
+In particular, HIF-1α decreases cell proliferation by activating cyclin-dependent kinase inhibitor expression, inhibiting cell cycle checkpoint progression [@gardner2001], and antagonizing pro-proliferative MYC signaling [@koshiji2004].
+Despite these canonical effects of HIF-1α activation, there are many examples where cells continue to proliferate despite hypoxic stress, including cancer cells, stem cells, and lung vascular cells [@hubbi2015].
+How these cells meet the metabolic needs of sustained proliferation in hypoxia and how these adaptations are regulated are an active area of investigation [@jain2020; @oldham2015; @lee2020b].
+Since hypoxia is a prominent feature of cancer biology as tumor growth outstrips blood supply, most detailed metabolic studies of hypoxic cell metabolism have used tumor cell models, yielding important insights into the metabolic pathobiology of cancer [@wise2011; @metallo2011; @melendez-rodriguez2019; @garcia-bermudez2018; @lee2019a; @jiang2016].
+For example, stable isotope tracing and metabolic flux analyses identified a critical role for reductive carboxylation of glutamine-derived α-ketoglutarate for lipid biosynthesis in supporting tumor growth [@metallo2011; @gameiro2013a; @scott2011; @wise2011], and metabolomic studies identified aspartate as a limiting metabolite for cancer cell proliferation under hypoxia [@garcia-bermudez2018].
+By contrast, comparatively little is known about metabolic adaptations of primary cells to hypoxia.
+Indeed, the importance of reductive carboxylation or aspartate biosynthesis remains to be elucidated in these cells.
+A more complete understanding of primary cell metabolic adaptation to hypoxia would provide an important context for understanding how metabolic reprogramming supports normal cellular responses to hypoxia, how these responses may be (mal)adaptive in a variety of disease contexts, and how the hypoxia metabolic program in primary cells differs from that observed in cancer cells.
+
+To address these questions, we have developed models of bioenergetic carbon flux in human lung fibroblasts (LFs) and pulmonary artery smooth muscle cells (PASMCs) cultured in 21% or 0.5% oxygen.
+These cells may be exposed to a wide range of oxygen concentrations *in vivo*, continue to proliferate despite hypoxic culture conditions *in vitro*, and play important roles in the pathology of non-cancerous diseases in which tissue hypoxia is a prominent feature.
+We found that hypoxia fails to increase glycolysis in these primary cells despite robust up-regulation of the HIF-1α transcriptional program.
+In normoxia, HIF-1α stabilization by the PHD inhibitor molidustat (BAY-85-3934, "BAY") [@flamme2014] did increase glycolysis and lactate efflux; however, hypoxia blocked this response.
+These findings suggested the existence of important hypoxia-dependent regulatory mechanisms that override the metabolic consequences of HIF-1α-dependent up-regulation of glycolytic gene expression in human primary cells.
+Transcriptomic profiling suggested a critical role for the transcription factor MYC in the adaptive response to hypoxia.
+Using knockdown and overexpression approaches, we demonstrate that MYC attenuates HIF-driven glycolysis in hypoxia and following BAY treatment.
 
@@
 # Results
+
+## Hypoxia uncouples HIF target gene expression and glycolytic flux
+
+The goal of this study was to characterize hypoxia-induced metabolic changes in proliferating primary LFs and PASMCs.
+To accomplish this goal, we adopted a metabolic flux analysis (MFA) technique that enabled us to link intracellular metabolic fluxes to cell proliferation rates.
+MFA fits cell proliferation rate, extracellular flux measurements, and ^13^C intracellular isotope labeling patterns to a model of cell metabolism [@antoniewicz2018].
+This analysis reconstructs comprehensive flux maps that depict the flow of carbon from extracellular substrates, throughout intracellular metabolism, and into cell biomass and metabolic by-products [@young2014].
+These models assume that cells are at a metabolic pseudo-steady state over the experimental time course [@buescher2015].
+Exponential growth phase is throught to reflect metabolic pseudo-steady state, because cells in the culture steadily divide at their maximal condition-specific rate, as long as nutrient supply does not become limiting [@buescher2015; @ahn2011].
+Thus, we first set out to define experimental conditions to capture exponential growth phase in normoxic and hypoxic cultures.
+
+Cells were seeded and placed into hypoxia for 24 h prior to sample collection to provide adequate time for activation of the hypoxia-dependent transcriptional program (**Fig \@ref(fig:m1)A**).
+We selected 0.5% oxygen for hypoxia as this level yielded the most reproducible phenotypic differences compared to 21% oxygen culture while being physiologically relevant and above the K~M~ of cytochrome *c* oxidase (electron transport chain complex IV) for oxygen [@lee2020b; @wenger2015].
+From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (**Fig \@ref(fig:m1)B**).
+LFs cultured in 0.5% oxygen grew more slowly than LFs cultured in 21% oxygen (**Fig \@ref(fig:m1)C**), but slower growth was not associated with decreased cell viability (**Fig S1A**).
+As anticipated, hypoxic cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1 (GLUT1) and lactate dehydrogenase A (LDHA) (**Figs \@ref(fig:m1)D-H**).
+These changes persisted for the duration of the experimental time course.
+
+Having identified experimental conditions for exponential growth, we next determined extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (**Figs \@ref(fig:m1)I-J**).
+Flux calculations incorporated changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@murphy2013a] (**Figure S1**).
+
+Surprisingly, we observed no significant differences in glucose uptake or lactate efflux rates in 0.5% oxygen cultures (**Figure \@ref(fig:m1)I**) despite activation of the HIF-1 transcriptional program as reflected by increased expression of GLUT1 and LDHA.
+No change in glucose or lactate fluxes were observed when the ambient oxygen level was decreased further to 0.2% (**Figure S2**).
+Similar results were also observed in PASMCs where lactate efflux was significantly *decreased* by 0.5% ambient oxygen (**Figure S3**).
 
@@
 ```{r m1, fig.cap = "(ref:m1)"}
-knitr::include_graphics(
-  system.file(
-    "manuscript/figures/m1.png",
-    package = "package"
-  )
-)
+# knitr::include_graphics(
+#   system.file(
+#     "manuscript/figures/m1.png",
+#     package = "package"
+#   )
+# )
 ```

### 2022-07-11 R/1_dna-per-cell.R

@@ -10,16 +10,33 @@ clean_dna_per_cell <- function(filename) {
 }
 
 calculate_cells_per_dna <- function(tbl) {
-  tbl %>%
-    dplyr::filter(!(cell_type == "lf" & volume == "100" & cells == 300000)) %>%
-    dplyr::filter(!(cell_type == "pasmc" & volume == "200" & cells == 400000)) %>%
-    dplyr::group_by(cell_type, volume) %>%
-    tidyr::nest() %>%
+  tbl |>
+    dplyr::filter(!(cell_type == "lf" & volume == "100" & cells == 300000)) |>
+    dplyr::filter(!(cell_type == "pasmc" & volume == "200" & cells == 400000)) |>
+    dplyr::group_by(cell_type, volume) |>
+    tidyr::nest() |>
     dplyr::mutate(
       model = map(data, ~lm(conc ~ 0 + cells, data = .x, na.action = modelr::na.warn)),
       glance = map(model, broom::tidy)
-    ) %>%
-    tidyr::unnest(c(glance)) %>%
-    dplyr::select(cell_type, volume, slope = estimate) %>%
+    ) |>
+    tidyr::unnest(c(glance)) |>
+    dplyr::select(cell_type, volume, slope = estimate) |>
     dplyr::mutate(slope = 1/slope)
+}
+
+clean_dna_count_hypoxia <- function(dna_count_hypoxia_file) {
+  df <-
+    readr::read_csv(dna_count_hypoxia_file) |>
+    dplyr::mutate(oxygen = factor(oxygen, levels = c("21%", "0.5%"))) |>
+    clean_technical_replicates()
+
+  std <-
+    df |>
+    dplyr::filter(!is.na(conc)) |>
+    lm(value ~ conc, data = _)
+
+  df |>
+    dplyr::filter(is.na(conc)) |>
+    {\(x) dplyr::mutate(x, conc = wmo::interpolate(x, std))}() |>
+    dplyr::select(-value)
 }

### 2022-07-11 R/1_figures.R

@@ -11,11 +11,489 @@ theme_plots <- function() {
       base_size = 8
     ),
     ggplot2::theme(
-      panel.border = ggplot2::element_rect(size = 0.25),
+      # panel.border = ggplot2::element_rect(size = 0.1),
+      axis.line = ggplot2::element_line(colour = "black", size = 0.25, lineend = "square"),
+      panel.border = ggplot2::element_blank(),
       plot.margin = ggplot2::margin(5, 5, 5, 5),
       plot.tag = ggplot2::element_text(face = "bold"),
       axis.title.y.left = ggplot2::element_text(margin = ggplot2::margin(r = 3))
     ),
     ggplot2::coord_cartesian(clip = "off")
   )
+}
+
+theme_patchwork <- function(design = NULL, widths = NULL, heights = NULL, tags = "A", ...) {
+  list(
+    patchwork::plot_layout(
+      design = design,
+      widths = widths,
+      heights = heights,
+      ...
+    ),
+    patchwork::plot_annotation(
+      tag_levels = tags,
+      theme = ggplot2::theme(plot.margin = ggplot2::margin(0, 0, 0, 0))
+    )
+  )
+}
+
+write_figures <- function(plot, filename, path = "manuscript/figures") {
+  # path <- "manuscript/figures"
+
+  gtab <- patchwork::patchworkGrob(plot)
+
+  overall_width <-
+    grid::convertWidth(
+      sum(gtab$widths),
+      unitTo = "in",
+      valueOnly = TRUE
+    )
+
+  overall_height <-
+    grid::convertHeight(
+      sum(gtab$heights),
+      unitTo = "in",
+      valueOnly = TRUE
+    )
+
+  ggplot2::ggsave(
+    filename = filename,
+    plot = plot,
+    device = ragg::agg_png,
+    path = path,
+    width = overall_width,
+    height = overall_height,
+    units = "in",
+    res = 300
+  )
+
+  if (file.exists("Rplots.pdf")) unlink("Rplots.pdf")
+
+  stringr::str_c(path, "/", filename)
+}
+
+plot_time_lines <- function(
+    df,
+    y,
+    ylab = prot,
+    clr = c("oxygen", "treatment", "group")
+) {
+  ggplot2::ggplot(df) +
+    ggplot2::aes(
+      x = time,
+      y = {{y}},
[…]

### 2022-07-11 R/1_utils.R

@@ -167,3 +167,20 @@ my_kable <- function(data, ...) {
     )
 }
 
+plot_image <- function(img, scale = 1, hjust = 0, vjust = 0) {
+  # blot_image <- magick::image_read_pdf(blot_image)
+  img <- magick::image_read(img)
+
+  cowplot::ggdraw() +
+    cowplot::draw_image(
+      img,
+      scale = scale,
+      hjust = hjust,
+      vjust = vjust
+    ) +
+    theme_plots() +
+    ggplot2::theme(
+      panel.border = ggplot2::element_blank(),
+      axis.line = ggplot2::element_blank()
+    )
+}

### 2022-07-11 _targets.R

@@ -631,55 +631,158 @@ list(
     format_cosmos(cosmos_forward, cosmos_reverse)
   ),
 
+  # m1 ----------------------------------------------------------------------
+
+  tar_target(
+    lf_hyp_05_timeline_png,
+    system.file("manuscript/ai/lf-hypoxia-timeline.png", package = "Copeland.2022.hypoxia.flux"),
+    format = "file"
+  ),
+  tar_target(
+    lf_hyp_05_timeline,
+    plot_image(lf_hyp_05_timeline_png, scale = 1.4, hjust = 0.2, vjust = 0.1)
+  ),
+  tar_target(
+    lf_hyp_05_growth_curve,
+    plot_growth_curve(flux_measurements, cell = "lf", exp = "05")
+  ),
+  tar_target(
+    lf_hyp_05_growth_rate,
+    plot_growth_rates(growth_rates, cell = "lf", exp = "05")
+  ),
+  tar_target(
+    lf_hyp_05_blot_png,
+    path_to_manuscript("ai/lf_05_hif1a-ldha-blots.png"),
+    format = "file"
+  ),
+  tar_target(
+    lf_hyp_05_blot,
+    plot_image(lf_hyp_05_blot_png, scale = 1.3, hjust = 0.2, vjust = 0)
+  ),
+  tar_target(
+    lf_hyp_05_hif1a_prot,
+    plot_expression(blot_norm, "lf_05", "hif1a", "HIF-1α protein\n(normalized)")
+  ),
+  tar_target(
+    lf_hyp_05_ldha_prot,
+    plot_expression(blot_norm, "lf_05", "ldha", "LDHA protein\n(normalized)")
+  ),
+  tar_target(
+    lf_hyp_05_glut1_rna,
+    plot_expression(mrna_norm, "lf_05", "glut1", "GLUT1 mRNA\n(normalized)")
+  ),
+  tar_target(
+    lf_hyp_05_ldha_rna,
+    plot_expression(mrna_norm, "lf_05", "ldha", "LDHA mRNA\n(normalized)")
+  ),
+  tar_target(
+    m1,
+    arrange_fluxes(
+      lf_hyp_05_timeline,
+      lf_hyp_05_growth_curve,
+      lf_hyp_05_growth_rate,
+      lf_hyp_05_blot,
+      lf_hyp_05_hif1a_prot,
+      lf_hyp_05_glut1_rna,
+      lf_hyp_05_ldha_rna,
[…]

### 2022-07-12 manuscript/manuscript.Rmd
more writing of extracellular flux results

@@
 
-## Hypoxia uncouples HIF target gene expression and glycolytic flux
+## Hypoxia uncouples glycolytic gene expression from glycolytic metabolic flux
 
@@
 These models assume that cells are at a metabolic pseudo-steady state over the experimental time course [@buescher2015].
-Exponential growth phase is throught to reflect metabolic pseudo-steady state, because cells in the culture steadily divide at their maximal condition-specific rate, as long as nutrient supply does not become limiting [@buescher2015; @ahn2011].
+Exponential growth phase is thought to reflect metabolic pseudo-steady state, because cells in the culture steadily divide at their maximal condition-specific rate, as long as nutrient supply does not become limiting [@buescher2015; @ahn2011].
 Thus, we first set out to define experimental conditions to capture exponential growth phase in normoxic and hypoxic cultures.
@@
 Having identified experimental conditions for exponential growth, we next determined extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (**Figs \@ref(fig:m1)I-J**).
-Flux calculations incorporated changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@murphy2013a] (**Figure S1**).
+Flux calculations incorporated changes in cell number, extracellular metabolite concentrations, metabolite degradation rates, and medium evaporation over time [@murphy2013a] (**Figure S1**).
+Interestingly, while we observed a modest increase in glucose uptake, we found that hypoxia actually decreased lactate efflux (**Figs \@ref(fig:m1)I**).
+This decrease in lactate efflux occurred despite activation of the HIF-1α transcriptional program as reflected by increased expression of GLUT1 and LDHA.
+To test if more severe hypoxia would augment glycolysis, we culture cells in 0.2% ambient oxygen (**Fig S2**).
+These results were similar to 0.5% oxygen culture.
+We observed no change in glucose or lactate fluxes.
+To test if this unexpected response was unique to LFs, we next studied PASMCs under 0.5% oxygen conditions (**Fig S3**).
+PASMCs grew faster than LFs, and so samples were collected every 12 h for 48 h for these cells.
+Again, similar to LFs at 0.5% and 0.2% oxygen, we observed no change in glucose uptake and reduced lactate efflux in PASMCs regardless of HIF-1α stabilization.
+Together, these data suggest that hypoxia uncouples HIF target gene expression and glycolytic flux in proliferating primary cells.
 
-Surprisingly, we observed no significant differences in glucose uptake or lactate efflux rates in 0.5% oxygen cultures (**Figure \@ref(fig:m1)I**) despite activation of the HIF-1 transcriptional program as reflected by increased expression of GLUT1 and LDHA.
-No change in glucose or lactate fluxes were observed when the ambient oxygen level was decreased further to 0.2% (**Figure S2**).
-Similar results were also observed in PASMCs where lactate efflux was significantly *decreased* by 0.5% ambient oxygen (**Figure S3**).
+## HIF stabilization in normoxia does increase glycolysis
+
+Given that hypoxia exposure did not increase glycolysis in LFs, we next wanted to determine how these cells responded to HIF-1α stabilization in normoxia.
+To accomplish this, LFs were treated with the PHD inhibitor molidustat (BAY) using a similar time course as our hypoxia experiments.
+Cells were treated with BAY for 24 h to activate the HIF transcriptional program prior to sample collection (**Fig \@ref(fig:m2)A**).
+Similar to hypoxia treatment, BAY decreased cell growth rate (**Figs \@ref(fig:m2)B-C**) and activated the HIF-1α transcriptional program (**Figs \@ref(fig:m1)D-H**).
+Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1α target gene transcription and protein expression.
+Unlike hypoxia, HIF-1α stabilization in normoxia markedly increased glucose uptake and lactate efflux (**Fig \@ref(fig:m1)I**), as expected based on the pattern of glycolytic gene expression that we observed.
+Interestingly, although hypoxia and BAY treatments resulted in similar increases in HIF-1α, GLUT1, and LDHA, the glycolytic response was markedly different.
+
+## Extracellular flux measurements highlight subtle differences between hypoxia and BAY
+
+In addition to glucose and lactate, we also determined the extracellular fluxes of pyruvate and amino acids (**Figs \@ref(fig:m1)J, \@ref(fig:m2)J**).
+Overall, changes were modest, with hypoxia generally decreasing the fluxes of all measured metabolites.
+Notably, a substantial decrease in glutamine consumption was observed in LFs cultured in 0.5% oxygen.
+This observation is in contrast to previous studies in cancer cell metabolism demonstrating increased glutamine uptake as a key feature of the metabolic response to hypoxia [@metallo2011; @wise2011; @gameiro2013].
+Similar patterns were observed in 0.2% oxygen (**Figure S2**).
+In PASMCs, glutamine uptake did increase, as did the uptake of branched-chain amino acids and arginine (**Figure S3**), highlighting differential responses of these mesenchymal cells to hypoxia.
+Comparatively modest effects of BAY on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (**Figure \@ref(fig:m1)J**).
+BAY treatment elicited different responses to hypoxia with relative preservation of glutamine and asparagine uptake and alanine and glutamate efflux compared to hypoxia.
+
+## Stable isotope labeling
+
+<!-- discuss LF v. PASMC -->
 

### 2022-07-12 R/1_figures.R

@@ -1,6 +1,8 @@
 # 1_figures.R
 
-# plot setup
+
+# setup -------------------------------------------------------------------
+
 clrs <- c(RColorBrewer::brewer.pal(4, "Set1")[1:4], "#08306b", RColorBrewer::brewer.pal(9, "Set1")[9:8])
 names(clrs) <- c("21%", "0.5%", "DMSO", "BAY", "0.2%", "siCTL", "siMYC")
 
@@ -91,10 +93,18 @@ plot_time_lines <- function(
     #   alpha = 0.25,
     #   show.legend = FALSE
     # ) +
+    # ggplot2::stat_summary(
+    #   geom = "linerange",
+    #   fun.data = ggplot2::mean_se,
+    #   size = 0.5,
+    #   show.legend = FALSE
+    # ) +
     ggplot2::stat_summary(
-      geom = "linerange",
+      geom = "errorbar",
       fun.data = ggplot2::mean_se,
-      size = 0.5,
+      color = "black",
+      width = 2,
+      size = 0.25,
       show.legend = FALSE
     ) +
     ggplot2::stat_summary(
@@ -171,6 +181,7 @@ plot_growth_rates <- function(
     broom::tidy() |>
     dplyr::mutate(
       group = stringr::str_extract(contrast, "(?<= - ).*"),
+      x = 1.5,
       y = Inf,
       vjust = 1,
       label = annot_p(p.value)
@@ -197,13 +208,6 @@ plot_growth_rates <- function(
       show.legend = FALSE,
       alpha = 0.5
     ) +
-    ggplot2::stat_summary(
-      geom = "errorbar",
-      fun.data = ggplot2::mean_se,
-      width = 0.2,
-      size = 0.25,
-      show.legend = FALSE
-    ) +
     ggbeeswarm::geom_beeswarm(
       ggplot2::aes(fill = group),
       pch = 21,
@@ -213,9 +217,17 @@ plot_growth_rates <- function(
       color = "white",
       show.legend = FALSE
     ) +
+    ggplot2::stat_summary(
+      geom = "errorbar",
+      fun.data = ggplot2::mean_se,
+      width = 0.2,
+      size = 0.25,
+      show.legend = FALSE
+    ) +
     ggplot2::geom_text(
       data = annot,
       ggplot2::aes(
+        x = x,
         label = label,
         y = y,
         vjust = vjust
[…]

### 2022-07-12 _targets.R

@@ -635,12 +635,12 @@ list(
 
   tar_target(
     lf_hyp_05_timeline_png,
-    system.file("manuscript/ai/lf-hypoxia-timeline.png", package = "Copeland.2022.hypoxia.flux"),
+    system.file("manuscript/ai/lf_hyp_05_timeline.png", package = "Copeland.2022.hypoxia.flux"),
     format = "file"
   ),
   tar_target(
     lf_hyp_05_timeline,
-    plot_image(lf_hyp_05_timeline_png, scale = 1.4, hjust = 0.2, vjust = 0.1)
+    plot_image(lf_hyp_05_timeline_png, scale = 1.6, hjust = 0.2, vjust = 0.1)
   ),
   tar_target(
     lf_hyp_05_growth_curve,
@@ -676,6 +676,14 @@ list(
     plot_expression(mrna_norm, "lf_05", "ldha", "LDHA mRNA\n(normalized)")
   ),
   tar_target(
+    lf_hyp_05_high,
+    plot_high_fluxes(fluxes, "lf", "05")
+  ),
+  tar_target(
+    lf_hyp_05_low,
+    plot_low_fluxes(fluxes, "lf", "05")
+  ),
+  tar_target(
     m1,
     arrange_fluxes(
       lf_hyp_05_timeline,
@@ -685,7 +693,9 @@ list(
       lf_hyp_05_hif1a_prot,
       lf_hyp_05_glut1_rna,
       lf_hyp_05_ldha_rna,
-      lf_hyp_05_ldha_prot
+      lf_hyp_05_ldha_prot,
+      lf_hyp_05_high,
+      lf_hyp_05_low
     )
   ),
   tar_target(
@@ -725,10 +735,10 @@ list(
     evap_plot,
     plot_evap_data(evap_clean)
   ),
-  # tar_target(
-  #   s1e,
-  #   plot_k(degradation_rates, k)
-  # ),
+  tar_target(
+    k_plot,
+    plot_k(degradation_rates, k)
+  ),
   tar_target(
     s1,
     arrange_s1(
@@ -736,13 +746,232 @@ list(
       lf_dna_curve,
       pasmc_dna_curve,
       dna_count_hypoxia_plot,
-      evap_plot
-      # k_plot
+      evap_plot,
+      k_plot
     )
   ),
   tar_target(
     s1_figure,
     write_figures(s1, "s1.png"),
+    format = "file"
+  ),
+
[…]

### 2022-07-12 _targets.R
rename original reactions file

@@ -286,7 +286,7 @@ list(
 
   tar_target(
     reactions_file,
-    path_to_reports("modeling/matlab-input/reactions.csv"),
+    path_to_reports("modeling/matlab-input/reactions_01.csv"),
     format = "file"
   ),
   tar_target(

### 2022-07-12 R/1_biomass.R
trial to refit models with split ASP and MAL pools

@@ -102,7 +102,7 @@ calculate_biomass_equations <- function(biomass) {
       dplyr::select(.data$metabolite, .data$coefficient) |>
       dplyr::mutate(
         metabolite = dplyr::case_when(
-          .data$metabolite %in% c("AcCoA") ~ stringr::str_c(metabolite, ".c"),
+          .data$metabolite %in% c("AcCoA", "ASP") ~ stringr::str_c(metabolite, ".c"),
           TRUE ~ metabolite
         ))
   }

### 2022-07-12 _targets.R

@@ -26,7 +26,8 @@ tar_option_set(
   packages = c("tidyverse", "patchwork", "scales"),
   # packages = c("tidyverse", "patchwork", "xcms"),
   # imports = c("rnaseq.lf.hypoxia.molidustat"),
-  format = "qs"
+  format = "qs",
+  error = "continue"
 )
 
 # targets -----------------------------------------------------------------
@@ -286,7 +287,7 @@ list(
 
   tar_target(
     reactions_file,
-    path_to_reports("modeling/matlab-input/reactions_01.csv"),
+    path_to_reports("modeling/matlab-input/reactions_02.csv"),
     format = "file"
   ),
   tar_target(

### 2022-07-12 analysis/modeling/lf_21.m

@@ -3,8 +3,8 @@ function m = normoxia()
 % setup
 oxygen = '21%';
 treatment = 'None';
-in_path = '/Users/will/Dropbox (Partners HealthCare)/Copeland.2021.hypoxia.flux/inst/analysis/modeling/matlab-input/';
-out_path = '/Users/will/Dropbox (Partners HealthCare)/Copeland.2021.hypoxia.flux/inst/analysis/modeling/';
+in_path = '/Users/will/Dropbox (Partners HealthCare)/_data/Copeland.2022.hypoxia.flux/analysis/modeling/matlab-input/';
+out_path = '/Users/will/Dropbox (Partners HealthCare)/_data/Copeland.2022.hypoxia.flux/analysis/modeling/';
 cell_type = 'lf';
 addpath(genpath(in_path));
 output = [out_path cell_type '_21/' cell_type '_21_' datestr(now,'yyyy-mm-dd-HH-MM-SS') '.mat'];
@@ -19,7 +19,7 @@ biomasseqn = [biomasseqn num2str(biomass{height(biomass), 2}) ' ' char(biomass{h
 clear biomass;
 
 % reaction equations
-reactions = readtable('reactions.csv');
+reactions = readtable('reactions_02.csv');
 % reactions = rmmissing(reactions);
 reaction_list = [biomasseqn; reactions.equation]; 
 reaction_ids = ['BIOMASS'; reactions.name];
@@ -85,8 +85,8 @@ for label = {'glc2' 'glc6' 'q5'}
             y = [y c];
             h = [h time];
         end
-        d{n}{char(metabolite)}.mdvs = mdv(y);
-        d{n}{char(metabolite)}.mdvs.time = h;
+        d{n}{char(metabolite)}.idvs = idv(y);
+        d{n}{char(metabolite)}.idvs.time = h;
     end
     n = 1 + n;
 end
@@ -114,7 +114,7 @@ clear a b c y n x z time compound label;
 sem_value = 0.013;
 for i=1:length(d)
     for j=1:length(d{i})
-        d{i}{d{i}.id{j}}.mdvs.std = sem_value * ones(length(d{i}{d{i}.id{j}}.mdvs.val),1);
+        d{i}{d{i}.id{j}}.idvs.std = sem_value * ones(length(d{i}{d{i}.id{j}}.idvs.val),1);
     end
[…]

### 2022-07-13 analysis/modeling/lf_21.m
trial to refit models by adding split CIT pool

@@ -19,7 +19,7 @@ biomasseqn = [biomasseqn num2str(biomass{height(biomass), 2}) ' ' char(biomass{h
 clear biomass;
 
 % reaction equations
-reactions = readtable('reactions_02.csv');
+reactions = readtable('reactions_03.csv');
 % reactions = rmmissing(reactions);
 reaction_list = [biomasseqn; reactions.equation]; 
 reaction_ids = ['BIOMASS'; reactions.name];
@@ -129,7 +129,7 @@ f.std = fluxes.se';
 % f.std = 0.1 * f.val;
 
 % set fixed rates
-fixed_flux_equations = {'sPYR' 'sMAL' 'sASP'}';
+fixed_flux_equations = {'sPYR' 'sMAL' 'sASP' 'sCIT'}';
 
 for i=1:length(fixed_flux_equations)
     for j=1:length(r)
@@ -170,7 +170,7 @@ m.states{'CO2','c'}.bal = false;
 for metabolite = {'LAC.x' 'PYR.x' 'ALA.x' 'GLN.x' 'GLU.x' 'ASP.x'}
     m.states{char(metabolite)}.bal = false;
 end
-for pool = {'PYR.ms' 'ASP.ms' 'MAL.ms'}
+for pool = {'PYR.ms' 'ASP.ms' 'MAL.ms' 'CIT.ms'}
     m.states{char(pool)}.val = 0;
     m.states{char(pool)}.fix = true;
 end

### 2022-07-14 manuscript/manuscript.Rmd
start writing up isotope tracing studies

@@
 
-## Hypoxia uncouples glycolytic gene expression from glycolytic metabolic flux
+## Hypoxia uncouples HIF-dependent glycolytic gene expression from glycolytic metabolic flux
 
@@
 To accomplish this goal, we adopted a metabolic flux analysis (MFA) technique that enabled us to link intracellular metabolic fluxes to cell proliferation rates.
-MFA fits cell proliferation rate, extracellular flux measurements, and ^13^C intracellular isotope labeling patterns to a model of cell metabolism [@antoniewicz2018].
-This analysis reconstructs comprehensive flux maps that depict the flow of carbon from extracellular substrates, throughout intracellular metabolism, and into cell biomass and metabolic by-products [@young2014].
+MFA fits cell proliferation rate, extracellular flux measurements, and ^13^C intracellular isotope labeling patterns to a computational model of cell metabolism [@antoniewicz2018].
+This analysis reconstructs comprehensive flux maps that depict the flow of carbon from extracellular substrates, through intracellular metabolic pathways, and into cell biomass and metabolic by-products [@young2014].
 These models assume that cells are at a metabolic pseudo-steady state over the experimental time course [@buescher2015].
-Exponential growth phase is thought to reflect metabolic pseudo-steady state, because cells in the culture steadily divide at their maximal condition-specific rate, as long as nutrient supply does not become limiting [@buescher2015; @ahn2011].
+Exponential growth phase is thought to reflect metabolic pseudo-steady state as cells in the culture steadily divide at their maximal condition-specific rate, provided nutrient supply does not become limiting [@buescher2015; @ahn2011].
 Thus, we first set out to define experimental conditions to capture exponential growth phase in normoxic and hypoxic cultures.
@@
 
-## HIF stabilization in normoxia does increase glycolysis
-
 Given that hypoxia exposure did not increase glycolysis in LFs, we next wanted to determine how these cells responded to HIF-1α stabilization in normoxia.
-To accomplish this, LFs were treated with the PHD inhibitor molidustat (BAY) using a similar time course as our hypoxia experiments.
+To accomplish this, LFs were treated with the PHD inhibitor molidustat (BAY, 10 μM) using a similar time course as our hypoxia experiments.
 Cells were treated with BAY for 24 h to activate the HIF transcriptional program prior to sample collection (**Fig \@ref(fig:m2)A**).
-Similar to hypoxia treatment, BAY decreased cell growth rate (**Figs \@ref(fig:m2)B-C**) and activated the HIF-1α transcriptional program (**Figs \@ref(fig:m1)D-H**).
+As with hypoxia, BAY decreased cell growth rate (**Figs \@ref(fig:m2)B-C**) and activated the HIF-1α transcriptional program (**Figs \@ref(fig:m1)D-H**).
 Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1α target gene transcription and protein expression.
-Unlike hypoxia, HIF-1α stabilization in normoxia markedly increased glucose uptake and lactate efflux (**Fig \@ref(fig:m1)I**), as expected based on the pattern of glycolytic gene expression that we observed.
+Unlike hypoxia, however, HIF-1α stabilization in normoxia markedly increased glucose uptake and lactate efflux (**Fig \@ref(fig:m1)I**), as expected based on the pattern of glycolytic gene expression that we observed.
 Interestingly, although hypoxia and BAY treatments resulted in similar increases in HIF-1α, GLUT1, and LDHA, the glycolytic response was markedly different.
 
-## Extracellular flux measurements highlight subtle differences between hypoxia and BAY
+## Extracellular fluxes are treatment and cell-type dependent
 
 In addition to glucose and lactate, we also determined the extracellular fluxes of pyruvate and amino acids (**Figs \@ref(fig:m1)J, \@ref(fig:m2)J**).
-Overall, changes were modest, with hypoxia generally decreasing the fluxes of all measured metabolites.
-Notably, a substantial decrease in glutamine consumption was observed in LFs cultured in 0.5% oxygen.
-This observation is in contrast to previous studies in cancer cell metabolism demonstrating increased glutamine uptake as a key feature of the metabolic response to hypoxia [@metallo2011; @wise2011; @gameiro2013].
-Similar patterns were observed in 0.2% oxygen (**Figure S2**).
-In PASMCs, glutamine uptake did increase, as did the uptake of branched-chain amino acids and arginine (**Figure S3**), highlighting differential responses of these mesenchymal cells to hypoxia.
-Comparatively modest effects of BAY on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (**Figure \@ref(fig:m1)J**).
-BAY treatment elicited different responses to hypoxia with relative preservation of glutamine and asparagine uptake and alanine and glutamate efflux compared to hypoxia.
+To our knowledge, this is the first comprehensive extracellular flux profiling of key metabolic substrates in these primary cells.
+In LFs, overall, changes in these fluxes were modest, with hypoxia generally decreasing the fluxes of all measured metabolites.
+These findings were similar with 0.2% oxygen exposure (**Fig S2J**).
 
-## Stable isotope labeling
+Notably, we observed a significant decrease in glutamine consumption in hypoxic LFs.
+This finding contrasts with previous studies of cancer cell metabolism demonstrating increased glutamine uptake as a key feature of the metabolic response to hypoxia [@metallo2011; @wise2011; @gameiro2013].
+In these systems, glutamine-derived 2-oxoglutarate was reductively carboxylated by isocitrate dehydrogenase enzymes to generate citrate for lipogenesis.
+In addition, glutamine has been shown to support TCA cycling in hypoxia in a Burkitt lymphoma model [@le2012].
+Unlike LFs, PASMCs did exhibit a trend toward increased glutamine uptake (**Figure S3J**), suggesting a greater reliance on these metabolic pathways in their adaptive response to hypoxia.
+
+In LFs, among all of the measured amino acid fluxes, proline consumption uniquely increased (**Fig \@ref(fig:m1)J**).
+Hypoxia increases collagen expression in these cells [@liu2013a] and proline constitutes ~ 10% of the total amino acid content of collagens.
+Together, these data suggest an important contribution of extracellular proline to collagen production in hypoxic LFs as has been observed in other fibroblast cell lineages [@szoka2017].
+
+In PASMCs, we observed increased consumption of the branched-chain amino acids (BCAAs) leucine and valine as well as arginine (**Figure S3J**), which was not observed in LFs.
+BCAAs are transaminated by branch chain amino transferase enzymes to branched chain α-keto acids (BCKAs).
+BCKAs are further metabolized to yield acyl-CoA derivatives for lipogenesis or oxidation [@mann2021; @crown2015].
+Previous studies have shown that hypoxia up-regulates arginase expression in hypoxic PASMCs [@chen2009; @xue2017] to support polyamine and proline synthesis required for cell proliferation [@li2001].
+Interestingly, activation of these metabolic pathways in hypoxia was not observed in LFs and suggests distinct metabolic vulnerabilities of these different cell types.
+
+Compared to hypoxia treatment, BAY demonstrated more modest effects on amino acid fluxes generally (**Figure \@ref(fig:m2)J**).
+In particular, glutamate efflux was not affected by BAY treatment, while it was reduced by hypoxia.
+Alanine efflux was increased by BAY treatment, but decreased by hypoxia.
+In addition to the glucose and lactate fluxes noted above, these findings further highlight fundamental differences in the metabolic consequences of HIF activation in normoxia and hypoxia.
+
+## Isotope tracing demonstrates less incorporation of glucose into downstream pathways
+
+To investigate intracellular metabolic reprogramming in hypoxic cells, we performed ^13^C stable isotope tracing with [U-^13^C~6~]-glucose, [1,2-^13^C~2~]-glucose, and [U-^13^C~5~]-glutamine.
+Isotopic enrichment of downstream metabolites in glycolysis and the TCA cycle were determined by LC-MS.
 

### 2022-07-14 R/1_figures.R

@@ -98,15 +98,15 @@ plot_time_lines <- function(
     #   fun.data = ggplot2::mean_se,
     #   size = 0.5,
     #   show.legend = FALSE
-    # ) +
-    ggplot2::stat_summary(
-      geom = "errorbar",
-      fun.data = ggplot2::mean_se,
-      color = "black",
-      width = 2,
-      size = 0.25,
-      show.legend = FALSE
-    ) +
+  # ) +
+  ggplot2::stat_summary(
+    geom = "errorbar",
+    fun.data = ggplot2::mean_se,
+    color = "black",
+    width = 2,
+    size = 0.25,
+    show.legend = FALSE
+  ) +
     ggplot2::stat_summary(
       geom = "line",
       fun.data = ggplot2::mean_se,
@@ -621,6 +621,9 @@ plot_low_fluxes <- function(
       cell_type %in% cell &
         experiment %in% exper &
         metabolite %nin% c("lactate", "glucose")
+    ) |>
+    dplyr::filter(
+      !(metabolite == "glutamine" & flux > 0)
     )
 
   annot <-
@@ -717,6 +720,140 @@ plot_low_fluxes <- function(
     )
 }
 
+annot_mids_main <- function(a, formula) {
+  lmerTest::lmer(
+    mid ~ group * isotope + (1 | date),
+    data = a
+  ) |>
+    emmeans::emmeans(~ group * isotope) |>
+    emmeans::mvcontrast("pairwise", mult.name = "isotope") |>
+    tibble::as_tibble() |>
+    dplyr::select(contrast, tidyselect::contains("p.value")) |>
+    dplyr::rename_with(~ "pval", .cols = tidyselect::contains("p.value")) |>
+    dplyr::mutate(
+      label = dplyr::case_when(
+        pval < 0.05 & contrast == "21% - 0.5%" ~ "*",
+        pval < 0.05 & contrast == "21% - 0.2%" ~ "*",
+        pval < 0.05 & contrast == "DMSO - BAY" ~ "†",
+        pval < 0.05 & contrast == "0.5% - BAY" ~ "‡"
+      ),
+      order = dplyr::case_when(
+        contrast == "21% - 0.5%" ~ 1,
+        contrast == "21% - 0.2%" ~ 1,
+        contrast == "DMSO - BAY" ~ 2,
+        contrast == "0.5% - BAY" ~ 3
+      )
+    ) |>
+    dplyr::filter(!is.na(label)) |>
[…]

### 2022-07-14 _targets.R

@@ -4,6 +4,8 @@ devtools::load_all()
 library(targets)
 library(tarchetypes)
 
+extrafont::loadfonts(quiet = TRUE)
+
 invisible(
   lapply(
     list.files(path = "R", pattern = "\\.R$", full.names = TRUE),
@@ -973,6 +975,34 @@ list(
   tar_target(
     m2_figure,
     write_figures(m2, "m2.png"),
+    format = "file"
+  ),
+
+  # m3 ----------------------------------------------------------------------
+
+  tar_target(
+    mid_glc6_pyr,
+    plot_mids(pruned_mids, "lf", "PYR", track = "glc6")
+  ),
+  tar_target(
+    mid_glc6_cit,
+    plot_mids(pruned_mids, "lf", "CIT", track = "glc6")
+  ),
+  tar_target(
+    mid_q5_cit,
+    plot_mids(pruned_mids, "lf", "CIT", track = "q5")
+  ),
+  tar_target(
+    m3,
+    arrange_m3(
+      mid_glc6_pyr,
+      mid_glc6_cit,
+      mid_q5_cit
+    )
+  ),
+  tar_target(
+    m3_figure,
+    write_figures(m3, "m3.png"),
     format = "file"
   ),
 

### 2022-07-19 manuscript/manuscript.Rmd
work on stable isotope tracing results and figure

@@
 
-## Isotope tracing demonstrates less incorporation of glucose into downstream pathways
+## Isotope tracing reveals altered substrate utilization in hypoxia
 
 To investigate intracellular metabolic reprogramming in hypoxic cells, we performed ^13^C stable isotope tracing with [U-^13^C~6~]-glucose, [1,2-^13^C~2~]-glucose, and [U-^13^C~5~]-glutamine.
-Isotopic enrichment of downstream metabolites in glycolysis and the TCA cycle were determined by LC-MS.
+Isotopic enrichment of downstream metabolites in glycolysis and the TCA cycle were determined by LC-MS (**Figs S3, S4**).
+Overall, relatively small changes in the patterns of isotope incorporation were observed following hypoxia or BAY treatment.
+The most substantial differences were observed in pyruvate (PYR), the terminal product of glycolysis, and citrate (CIT), a central metabolic node in TCA and fatty acid metabolism (**Figs \@ref(fig:m3)A-C**).
+Both hypoxia and BAY treatments decreased incorporation of glucose-derived carbon into pyruvate (**Fig \@ref(fig:m3)A**) (*i.e.,* the unlabeled, or M0, fraction was greater).
+This suggests a greater contribution from an unlabeled carbon source, such as extracellular pyruvate, lactate, or alanine, than from glucose to the intracellular pyruvate pool following PHD inhibition.
 
-<!-- discuss LF v. PASMC -->
+Total citrate labeling from [U-^13^C~6~]-glucose was unchanged across the treatment conditions (**Fig \@ref(fig:m3)B**).
+As expected, we observed decreased M2 and M4 citrate isotopes, consistent with decreased pyruvate dehydrogenase activity in hypoxia.
+Interestingly, we observed increased M3 and M5 citrate isotopes.
+Pyruvate carboxylase catalyzes the carboxylation of pyruvate to oxaloacetate after which all three pyruvate carbons are incorporated into citrate by citrate synthase.
+Thus, this labeling pattern suggests a more prominent contribution of pyruvate carboxylase to sustain TCA cycle anaplerosis despite pyruvate dehydrogenase inhibition following HIF-1α activation.
+By contrast to glucose labeling, much less citrate was labeled by glutamine with hypoxia or BAY with a more pronounced effect of BAY treatment (**Fig \@ref(fig:m3)C**), suggesting a less important contribution of glutamine to TCA anaplerosis under these conditions.
+In addition, the overall fraction of M5 citrate resulting from reductive carboxylation of glutamine-derived α-ketoglutarate was low (< 7%) (**Fig \@ref(fig:m3)D**).
+Although a hypoxia-mediated increase in M5 citrate was observed, the overall fraction was much less than the 10-20% levels previously reported in cancer cells [@wise2011; @metallo2011].
+
+The stable isotope labeling patterns in PASMCs were generally similar to LFs (**Figure S5**).
+The most notable differences between LF and PASMC labeling were observed in citrate.
+Compared with LFs, a much lower fraction of total citrate was labeled by glucose in PASMCs.
+Less activity of pyruvate carboxylase in these cells was suggested by decreased M3 and M5 citrate isotopes after glucose labeling.
+Interestingly, the M5 citrate fraction in PASMCs was more consistent with previous reports from the cancer literature (**Fig \@ref(fig:m3)D**), consistent with a more important role for glutamine metabolism for biomass synthesis in these cells.
+
+## Glycolytic flux in hypoxia is closely coupled to cell growth rate
 

### 2022-07-19 R/1_figures.R

@@ -712,7 +712,7 @@ plot_low_fluxes <- function(
     ) +
     theme_plots() +
     ggplot2::theme(
-      panel.grid.major.x = ggplot2::element_line(color = "gray90"),
+      panel.grid.major.x = ggplot2::element_line(color = "gray80", size = 0.1),
       legend.key.width = ggplot2::unit(0.5, "lines"),
       legend.key.height = ggplot2::unit(0.5, "lines"),
       legend.position = "bottom",
@@ -753,7 +753,6 @@ annot_mids_main <- function(a, formula) {
 plot_mids <- function(df, cell, metab, t = 72, track) {
   tracer_labels <- c(
     glc2 = expression(paste("[1,2-"^13, "C"[2], "]-GLC")),
-    # glc6 = expression(paste("[U-"^13, "C"[6], "]-GLC")),
     glc6 = bquote("[U-"^13 * "C"[6] * "]-GLC →" ~ .(metab)),
     q5 = bquote("[U-"^13 * "C"[5] * "]-GLN →" ~ .(metab)),
     lac3 = expression(paste("[U-"^13, "C"[3], "]-LAC"))
@@ -764,9 +763,9 @@ plot_mids <- function(df, cell, metab, t = 72, track) {
     df |>
     dplyr::filter(
       cell_type == cell &
-        metabolite %in% metab &
+        metabolite == metab &
         time == t &
-        tracer %in% track
+        tracer == track
     ) |>
     dplyr::filter(
       !(metabolite == "CIT" & isotope == "M6")
@@ -854,6 +853,242 @@ plot_mids <- function(df, cell, metab, t = 72, track) {
     )
 }
 
+plot_all_mids <- function(df, cell, t = 72, track) {
+  tracer_labels <- c(
+    expression(paste("[1,2-"^13, "C"[2], "] glucose")),
+    expression(paste("[U-"^13, "C"[6], "] glucose")),
+    expression(paste("[U-"^13, "C"[5], "] glutamine")),
+    expression(paste("[U-"^13, "C"[3], "] lactate"))
+  )
+  tracer_levels <- c("glc2", "glc6", "q5", "lac3")
+  metab <- c(
+    "FBP",
+    "3PG",
+    "PYR",
+    # "ALA",
+    # "SER",
+    "LAC",
+    "CIT",
+    "AKG",
+    # "GLN",
+    "GLU",
+    "MAL"
+    # "ASP"
+  )
+
+  x <-
+    df |>
+    dplyr::filter(
[…]

### 2022-07-19 _targets.R

@@ -993,17 +993,45 @@ list(
     plot_mids(pruned_mids, "lf", "CIT", track = "q5")
   ),
   tar_target(
+    mid_q5_m5_cit,
+    plot_m5_citrate(pruned_mids)
+  ),
+  tar_target(
     m3,
     arrange_m3(
       mid_glc6_pyr,
       mid_glc6_cit,
-      mid_q5_cit
+      mid_q5_cit,
+      mid_q5_m5_cit
     )
   ),
   tar_target(
     m3_figure,
     write_figures(m3, "m3.png"),
     format = "file"
+  ),
+
+  # s4 ----------------------------------------------------------------------
+
+  tar_target(
+    lf_mids,
+    plot_all_mids(pruned_mids, "lf")
+  ),
+  tar_target(
+    s4_figure,
+    write_figures(lf_mids, "s4.png")
+  ),
+
+
+  # s5 ----------------------------------------------------------------------
+
+  tar_target(
+    pasmc_mids,
+    plot_all_mids(pruned_mids, "pasmc", t = 36)
+  ),
+  tar_target(
+    s5_figure,
+    write_figures(pasmc_mids, "s5.png")
   ),
 
   # manuscript --------------------------------------------------------------

### 2022-07-25 manuscript/manuscript.Rmd
draft sections describing MFA results

@@
 The goal of this study was to characterize hypoxia-induced metabolic changes in proliferating primary LFs and PASMCs.
-To accomplish this goal, we adopted a metabolic flux analysis (MFA) technique that enabled us to link intracellular metabolic fluxes to cell proliferation rates.
-MFA fits cell proliferation rate, extracellular flux measurements, and ^13^C intracellular isotope labeling patterns to a computational model of cell metabolism [@antoniewicz2018].
+To accomplish this goal, we adopted a metabolic flux analysis technique that enabled us to link intracellular metabolic fluxes to cell proliferation rates.
+Metabolic flux analysis fits cell proliferation rate, extracellular flux measurements, and ^13^C intracellular isotope labeling patterns to a computational model of cell metabolism [@antoniewicz2018].
 This analysis reconstructs comprehensive flux maps that depict the flow of carbon from extracellular substrates, through intracellular metabolic pathways, and into cell biomass and metabolic by-products [@young2014].
@@
 This finding contrasts with previous studies of cancer cell metabolism demonstrating increased glutamine uptake as a key feature of the metabolic response to hypoxia [@metallo2011; @wise2011; @gameiro2013].
-In these systems, glutamine-derived 2-oxoglutarate was reductively carboxylated by isocitrate dehydrogenase enzymes to generate citrate for lipogenesis.
+In these systems, glutamine-derived α-ketoglutarate was reductively carboxylated by isocitrate dehydrogenase enzymes to generate citrate for lipogenesis.
 In addition, glutamine has been shown to support TCA cycling in hypoxia in a Burkitt lymphoma model [@le2012].
@@
 Less activity of pyruvate carboxylase in these cells was suggested by decreased M3 and M5 citrate isotopes after glucose labeling.
-Interestingly, the M5 citrate fraction in PASMCs was more consistent with previous reports from the cancer literature (**Fig \@ref(fig:m3)D**), consistent with a more important role for glutamine metabolism for biomass synthesis in these cells.
+Interestingly, the M5 citrate fraction in PASMCs was more consistent with previous reports from the cancer literature (**Fig \@ref(fig:m3)D**), suggesting a more important role for glutamine metabolism for biomass synthesis in these cells.
 
 ## Glycolytic flux in hypoxia is closely coupled to cell growth rate
+
+The mass isotopomer distribution for a given metabolite is determined by the complex relationship among the rate of isotope incorporation into the metabolic network, the contributions of unlabeled substrates, and fluxes through related pathways.
+To clarify how these labeling patterns reflect changes in intracellular metabolite fluxes, we next generated metabolic flux models incorporating the extracellular flux measurements and stable isotope tracing data described above.
+Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (**Fig S6**).
+Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by Isotopomer Network Compartment Analysis (INCA) [@murphy2013a; @young2014a; @jazmin2013] (**Figs \@ref(fig:m3)E-F, S7, Tables S1-S3**).
+
+Overall, LF and PASMC metabolic fluxes were dominated by high rates of glucose uptake and glycolysis (**Figs S7A-B**).
+Approximately 10% of cytoplasmic pyruvate enters the TCA cycle with the balance converted to lactate.
+Consistent with extracellular flux measurements and isotope labeling patterns described above, significant reductions in glycolysis, the TCA cycle, and amino acid metabolism were observed in the metabolic flux models of LFs cultured in hypoxia (**Fig \@ref(fig:m3)E**).
+A significant increase in pentose phosphate pathway flux was also observed, although the absolute flux through this pathway is low.
+By contrast, HIF-1α activation by BAY in 21% oxygen increased glycolysis and lactate fermentation by nearly 50% (**Figure \@ref(fig:m3)F**), but had a similar effect on decreasing serine and glutamine uptake as hypoxia.
+Metabolite fluxes in DMSO-treated cells were similar to 21% oxygen controls (**Table S1-S2**).
+
+In normoxia, the magnitude of intracellular metabolite fluxes was generally similar in LFs and PASMCs (**Figures S7A-C, Tables S1, S3**).
+Compared to LFs, PASMCs had slower rates of glycolysis and faster rates of TCA metabolism driven, in part, by increased glutamine uptake.
+In hypoxia, PASMCs exhibited similar decreases in glycolytic flux as LFs but also a marked, and unexpected, increase in TCA flux (**Figure S7D**).
+The increased TCA flux in PASMCs was driven by increased glutamine consumption.
+This finding is similar to a prior report of glutamine-driven oxidative phosphorylation in hypoxic cancer cells [@fan2013], where oxidative phosphorylation continued to provide the majority of cellular ATP even in 1% oxygen.
+
+Given the global decrease in bioenergetic metabolic flux in hypoxic LFs, we hypothesized that these differences may be a consequence of decreased growth rate.
+After normalizing metabolite fluxes in normoxia and hypoxia to the cell growth rate, modest increases (~10%) in glycolytic flux were observed (**Figure S7E**).
+This finding suggests that, while glycolysis increases relative to growth rate in hypoxic cells, the regulators of cell proliferation rate override the consequences of the HIF-1α transcriptional program.
+Indeed, even after adjusting for cell growth rate, the relative increase in glycolytic flux is modest compared to the marked up-regulation of glycolytic protein levels.
+BAY treatment decreased cell proliferation rate (**Figure \@ref(fig:m2)B**), indicating that, unlike hypoxia, BAY treatment in normoxia uncouples cell proliferation and metabolic flux.
+
+## Hypoxia and BAY treatment increase lactate oxidation
+
+Although the metabolite exchange fluxes for bidirectional reactions tend to be poorly resolved by metabolic flux analysis, two observations are worth highlighting (**Tables S1-S3**).
+First, consistent with the stable isotope tracing results, the modeled rate of reductive carboxylation through reverse flux by isocitrate dehydrogenase in LFs is low (~4 fmol/cell/h), unchanged by hypoxia, and modestly increased by BAY treatment.
+By contrast, the rate of reductive carboxylation increases 6-fold in PASMCs in hypoxia, highlighting an important role for this pathway in the metabolic response of PASMCs to hypoxia (**Fig \@ref(fig:m4)A**).
+
+Second, PHD inhibition is associated with a marked increase in the lactate transport exchange flux in LFs from ~ 0 to 1,500 and 700 fmol/cell/h in 0.5% oxygen and BAY treatment conditions, respectively, with similar results in PASMCs (**Fig \@ref(fig:m4)B**).
+Since the net lactate transport flux is secretion, this observation suggests increased lactate uptake with hypoxia or BAY treatment, a finding that may be consistent with the HIF-driven increased expression of the reversible lactate transporter MCT4 [@contreras-baeza2019].
+To investigate this hypothesis, LFs and PASMCs were treated with [U-^13^C~3~] lactate (2 mM) and ^13^C incorporation into intracellular metabolites was analyzed by LC-MS (**Figs \@ref(fig:m4), S4, S5**).
+Lactate labeled ~50% of citrate and ~20% of downstream TCA cycle metabolites in both LFs and PASMCs, indicating that lactate may be an important respiratory fuel source in these cells even though lactate efflux is high.
+Although lactate has been used less commonly than glucose and glutamine in stable isotope tracing studies, Faubert and colleagues [-@faubert2017] demonstrated lactate incorporation in human lung adenocarcinoma *in vivo*.
+In this study, lactate incorporation corresponded to regions of high glucose uptake as determined by [ ^18^F ]-fluorodeoxyglucose positron emission tomography, suggesting that lactate consumption can occur even in areas of high glucose utilization.
+Subsequently, investigators have demonstrated the importance of lactate as a metabolic fuel *in vivo* [@hui2017; @hui2020].
+As predicated from our metabolic flux analysis, with hypoxia or BAY treatment, we observed increased labeling of the TCA metabolites citrate (CIT), α-ketoglutarate (AKG), malate (MAL), and aspartate (ASP) in LFs.
+Interestingly, although increased labeling of pyruvate was observed in hypoxic PASMCs, the label was not incorporated into the TCA cycle as observed in LFs (**Fig S5**).
+
+In addition to downstream metabolites, we also observed hypoxia- and BAY-dependent increases in lactate incorporation in fructose bisphosphate (FBP) and 3-phosphoglycerate (3PG).
+This observation is consistent with prior reports describing hypoxia-mediated increases in gluconeogenesis and glycogen synthesis [@owczarek2020; @pelletier2012; @favaro2012].
+These data suggest that lactate also makes a small (~5% carbon) contribution to glycogen precursors.
+Together, these findings from stable isotope tracing of lactate reveal its important contribution to primary cell metabolism under standard culture conditions, but also reveal increased utilization of this substrate in hypoxia.
+
+## Hypoxia abrogates the effects of BAY on increasing glycolysis
+
+To reconcile the differential effects of prolyl hydroxylase inhibition by hypoxia and BAY, we next addressed whether hypoxia could suppress the effects of BAY on glucose and lactate fluxes
+
+(**Figures \@ref(fig:m5)A-C**).
+LFs cultured in standard growth medium were treated with BAY and placed in either 21% or 0.5% oxygen.
+Similar to previous experiments, BAY treatment decreased cell growth rate, increased glucose uptake, and increased lactate efflux in 21% oxygen.
+Interestingly, when combined with 0.5% oxygen, BAY treatment was unable to enhance lactate efflux.
+These data suggest that hypoxia antagonizes the effects of HIF-1 activation on glycolytic flux in these primary cells.
+
+To investigate these metabolic differences further, we performed metabolomic profiling of LFs treated for 72 h with hypoxia or BAY separately or in combination (**Figures \@ref(fig:m5)D-F**).
+Both 0.5% oxygen and BAY treatment induced marked changes in intracellular metabolite levels (**Figure S8**).
+Of 133 metabolites, 98 were differentially regulated by hypoxia and 54 were differentially regulated by BAY.
+Of the differentially regulated metabolites, 44 were affected by both 0.5% oxygen and BAY treatment (**Figure S8C**).
+Metabolite set enrichment analysis of KEGG biochemical pathways was consistent with the results of the metabolic models demonstrating significant enrichment of the pentose phosphate pathway with 0.5% oxygen or BAY treatment; decreased alanine, aspartate, and glutamate metabolism with hypoxia; and increased glycolysis/gluconeogenesis with BAY (**Figures S8D, S8E**).
+Indeed, aspartate was the most significantly decreased metabolite with both treatments, consistent with prior reports demonstrating an important role for HIF-1 regulation of aspartate biosynthesis in cancer cells [@RN634; @RN1438].
+
+Principal component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen than among the BAY-treatment groups (**Figure \@ref(fig:m5)D**).
+Moreover, these hypoxia-treated cells were well-segregated from BAY-treated cells.
+These observations are, again, consistent with the results of the metabolic flux models demonstrating an overriding effect of hypoxia *per se* on metabolic flux and highlighting important differences between hypoxic and pharmacologic PHD inhibition.
+To identify the key differences among these treatments, we performed linear modeling focusing on those metabolites with a significant interaction term in the model described by treatment × oxygen (**Figure \@ref(fig:m5)E**) (*i.e.*, metabolites where the change with hypoxia treatment was different from the change with BAY treatment).
+Of 133 metabolites, 77 were significantly differentially regulated by hypoxia and BAY treatments.
+Several patterns emerged from this analysis (**Figure \@ref(fig:m5)F**).
+Interestingly, the hypoxia-regulated metabolite, 2-hydroxyglutarate [@RN1603], increased following BAY treatment, but was unaffected by hypoxia treatment in proliferating LFs.
+The tricarboxylic acid (TCA) cycle metabolite aconitate was decreased in hypoxia, but increased by BAY treatment, while the opposite effect was observed for hydroxyproline and taurine.
+The glycolytic intermediate glyceraldehyde 3-phosphate (GAP) was decreased by both hypoxia and BAY, while hypoxia reversed the BAY-mediated decrease in γ-aminobutyric acid (GABA).
+A metabolite set enrichment analysis of these differentially regulated metabolites revealed the TCA cycle to be the most enriched KEGG metabolite set (**Figure \@ref(fig:m5)G**).
+Leading edge analysis demonstrated negative enrichment scores associated with all of the TCA metabolites detected by our platform (**Figure \@ref(fig:m5)H**).
+This result indicates a more modest impact of BAY treatment on the TCA cycle than hypoxia, as suggested by our metabolic flux models where hypoxia resulted in a 1.5-2-fold reduction of TCA flux compared to a 1.1-1.5-fold reduction with BAY treatment in normoxia (**Figure \@ref(fig:m3)**).
+
+In addition to these differential effects on polar metabolite levels, perhaps the most significant contrast between hypoxia and BAY treatment is the impact of hypoxia on cellular redox state.
+As oxygen deprivation causes reductive stress [@RN2409], we next measured the impact of these treatments on intracellular NAD(H) as a surrogate for cellular redox state (**Figures \@ref(fig:m5)I-K**).
+As expected, hypoxia increased the NADH/NAD^+^ ratio, driven primarily by a decrease in intracellular NAD^+^.
+Interestingly, while BAY treatment increased the levels of NADH, a concomitant increase in NAD^+^ resulted in preservation of the NADH/NAD^+^ ratio.
+As NADH accumulation is a putative inhibitor of glycolytic flux [@RN2210], this may be one mechanism by which glycolytic flux is decreased in hypoxia but not following BAY treatment.
 

### 2022-07-25 R/1_figures.R

@@ -1016,7 +1016,7 @@ plot_m5_citrate <- function(df) {
     dplyr::mutate(
       x = 1.5,
       y = Inf,
-      vjust = 1,
+      vjust = 1.5,
       label = annot_p(p.value)
     )
 
@@ -1086,7 +1086,273 @@ plot_m5_citrate <- function(df) {
       # ylim = c(0, NA),
       clip = "off"
     ) +
+    ggplot2::theme(
+      axis.line = ggplot2::element_blank(),
+      panel.border = ggplot2::element_rect(color = "black", fill = NA, size = 0.25)
+    ) +
     NULL
+}
+
+format_time_course_mids <- function(model_mids) {
+  tracer_labels <- c(
+    expression(paste("[1,2-"^13, "C"[2], "] glucose")),
+    expression(paste("[U-"^13, "C"[6], "] glucose")),
+    expression(paste("[U-"^13, "C"[5], "] glutamine")),
+    expression(paste("[U-"^13, "C"[3], "] lactate"))
+  )
+  tracer_levels <- c("glc2", "glc6", "q5", "lac3")
+
+  model_mids |>
+    tidyr::unnest(c(data)) |>
+    dplyr::filter(metabolite %in% c("FBP", "PYR", "CIT", "MAL")) |>
+    dplyr::filter(tracer != "lac3") |>
+    # dplyr::filter(isotope != "M6") |>
+    dplyr::mutate(
+      tracer = factor(
+        tracer,
+        levels = tracer_levels,
+        labels = tracer_labels
+      ),
+      metabolite = factor(metabolite, levels = c("FBP", "PYR", "CIT", "MAL"))
+    )
+}
+
+plot_mid_time_course <- function(df, cells, o2, treat, color) {
+  df |>
+    dplyr::filter(
+      cell_type == cells &
+        oxygen %in% o2 &
+        treatment == treat &
+        isotope == "M0"
+    ) |>
+    ggplot2::ggplot() +
+    ggplot2::aes(
+      x = time,
+      y = 1 - mean,
+      color = oxygen,
+      fill = oxygen
+    ) +
+    ggplot2::facet_grid(
+      metabolite ~ tracer,
+      labeller = ggplot2::label_parsed
+    ) +
+    ggplot2::geom_linerange(
+      ggplot2::aes(
+        ymin = 1 - mean - se,
+        ymax = 1 - mean + se
+      ),
+      color = "black",
+      size = 0.25,
[…]

### 2022-07-25 R/1_model.R

@@ -6,8 +6,8 @@ clean_model_fluxes <- function(map_flux_files, model_reactions){
     dplyr::select(-equation) |>
     dplyr::add_row(name = "BIOMASS", pathway = factor("biomass"), index = 0)
 
-  map_flux_files[stringr::str_detect(map_flux_files, "(lf|pasmc)_.*_model\\.csv")] %>%
-    rlang::set_names(stringr::str_extract(basename(.), pattern = ".*(?=_model\\.csv)")) |>
+  map_flux_files[stringr::str_detect(map_flux_files, "(lf|pasmc)_.*_model\\.csv")] |>
+    {\(x) rlang::set_names(x, stringr::str_extract(basename(x), pattern = ".*(?=_model\\.csv)"))}() |>
     purrr::map_dfr(readr::read_csv, .id = "experiment") |>
     tidyr::separate(
       experiment,
@@ -77,6 +77,59 @@ calculate_flux_differences <- function(map_fluxes, cell, control, experiment){
     )
 }
 
+make_cell_ratio_graph <- function(map_fluxes, nodes) {
+  edges <-
+    map_fluxes |>
+    dplyr::filter(treatment == "21%") |>
+    dplyr::mutate(
+      cell_type = dplyr::case_when(
+        cell_type == "lf" ~ "ctl",
+        cell_type == "pasmc" ~ "exp"
+      ),
+      equation = stringr::str_replace_all(equation, "\\d+\\.?\\d+\\*", "")
+    ) |>
+    tidyr::pivot_wider(names_from = cell_type, values_from = c(flux, lb, ub)) |>
+    dplyr::mutate(
+      ratio = dplyr::if_else(
+        pmax(lb_ctl, lb_exp) - pmin(ub_ctl, ub_exp) > 0,
+        flux_exp/flux_ctl,
+        NA_real_
+      ),
+      treatment_ctl = "lf",
+      treatment_exp = "pasmc"
+    ) |>
+    dplyr::ungroup() |>
+    dplyr::select(
+      pathway:equation,
+      treatment_ctl,
+      treatment_exp,
+      tidyselect::contains("ctl"),
+      tidyselect::contains("exp"),
+      ratio
+    ) |>
+    tidyr::pivot_longer(
+      tidyselect::matches("_(ctl|exp)"),
+      names_to = c(".value", NA),
+      names_sep = "_"
+    ) |>
+    dplyr::filter(treatment == "lf") |>
+    parse_eq() |>
+    dplyr::select(
[…]

### 2022-07-25 _targets.R

@@ -25,7 +25,7 @@ future::plan(future.callr::callr(workers = future::availableCores() - 1))
 
 # target-specific options
 tar_option_set(
-  packages = c("tidyverse", "patchwork", "scales"),
+  packages = c("tidyverse", "patchwork", "scales", "grid"),
   # packages = c("tidyverse", "patchwork", "xcms"),
   # imports = c("rnaseq.lf.hypoxia.molidustat"),
   format = "qs",
@@ -367,7 +367,7 @@ list(
 
   tar_target(
     map_flux_files,
-    path_to_data("model"),
+    path_to_data("model\\.csv"),
     format = "file"
   ),
   tar_target(
@@ -382,6 +382,63 @@ list(
     map_flux_difference_report,
     path = path_to_reports("flux-differences.Rmd"),
     output_dir = system.file("analysis/pdfs", package = "Copeland.2022.hypoxia.flux")
+  ),
+  tar_target(
+    node_file,
+    path_to_data("nodes\\.csv"),
+    format = "file"
+  ),
+  tar_target(
+    nodes,
+    readr::read_csv(node_file)
+  ),
+  tar_target(
+    lf_hypoxia_graph,
+    make_graph(map_flux_differences, nodes, cell = "lf", treat = "21%", normalizer = "none")
+  ),
+  tar_target(
+    lf_hypoxia_graph_ratio_plot,
+    plot_ratio_network(lf_hypoxia_graph, "Hypoxia/Normoxia")
+  ),
+  tar_target(
+    bay_graph,
+    make_graph(map_flux_differences, nodes, cell = "lf", treat = "DMSO", normalizer = "none")
+  ),
+  tar_target(
+    bay_graph_ratio_plot,
+    plot_ratio_network(bay_graph, "BAY/DMSO")
+  ),
+  tar_target(
+    lf_normoxia_graph_plot,
+    plot_normoxia_network(lf_hypoxia_graph, "LF\nNormoxia")
+  ),
+  tar_target(
+    lf_hypoxia_growth_graph,
+    make_graph(map_flux_differences, nodes, cell = "lf", treat = "0.5%", normalizer = "growth")
+  ),
+  tar_target(
+    lf_hypoxia_growth_graph_plot,
+    plot_ratio_network(lf_hypoxia_growth_graph, "Hypoxia/Normoxia\nGrowth Rate Normalized", edges = FALSE)
+  ),
+  tar_target(
[…]

### 2022-07-25 manuscript/manuscript.Rmd
update metabolism results and figures

@@
 
-To reconcile the differential effects of prolyl hydroxylase inhibition by hypoxia and BAY, we next addressed whether hypoxia could suppress the effects of BAY on glucose and lactate fluxes
-
-(**Figures \@ref(fig:m5)A-C**).
+To reconcile the differential effects of prolyl hydroxylase inhibition by hypoxia and BAY, we next addressed whether hypoxia could suppress the effects of BAY on glucose and lactate fluxes (**Figs \@ref(fig:m5)A-C**).
 LFs cultured in standard growth medium were treated with BAY and placed in either 21% or 0.5% oxygen.
 Similar to previous experiments, BAY treatment decreased cell growth rate, increased glucose uptake, and increased lactate efflux in 21% oxygen.
-Interestingly, when combined with 0.5% oxygen, BAY treatment was unable to enhance lactate efflux.
-These data suggest that hypoxia antagonizes the effects of HIF-1 activation on glycolytic flux in these primary cells.
+However, when combined with 0.5% oxygen, BAY treatment was unable to enhance lactate efflux.
+These data indicate that hypoxia antagonizes the effects of HIF-1α activation on glycolytic flux in these cells.
 
-To investigate these metabolic differences further, we performed metabolomic profiling of LFs treated for 72 h with hypoxia or BAY separately or in combination (**Figures \@ref(fig:m5)D-F**).
-Both 0.5% oxygen and BAY treatment induced marked changes in intracellular metabolite levels (**Figure S8**).
-Of 133 metabolites, 98 were differentially regulated by hypoxia and 54 were differentially regulated by BAY.
-Of the differentially regulated metabolites, 44 were affected by both 0.5% oxygen and BAY treatment (**Figure S8C**).
-Metabolite set enrichment analysis of KEGG biochemical pathways was consistent with the results of the metabolic models demonstrating significant enrichment of the pentose phosphate pathway with 0.5% oxygen or BAY treatment; decreased alanine, aspartate, and glutamate metabolism with hypoxia; and increased glycolysis/gluconeogenesis with BAY (**Figures S8D, S8E**).
-Indeed, aspartate was the most significantly decreased metabolite with both treatments, consistent with prior reports demonstrating an important role for HIF-1 regulation of aspartate biosynthesis in cancer cells [@RN634; @RN1438].
+To investigate these metabolic differences further, we performed metabolomic profiling of LFs treated for 72 h with hypoxia or BAY separately or in combination.
+Both 0.5% oxygen and BAY treatment induced marked changes in intracellular metabolite levels (**Fig S8**).
+Of 133 total metabolites, 99 were differentially regulated by hypoxia and 54 were differentially regulated by BAY.
+Of the differentially regulated metabolites, 44 were affected by both 0.5% oxygen and BAY treatment.
+Metabolite set enrichment analysis of KEGG biochemical pathways identified increased enrichment of arginine and proline metabolism and fatty acid biosynthesis pathways with hypoxia (**Fig S8D**).
+By contrast, BAY treatment was enriched for metabolites involved in pentose/glucuronate interconversions and glycolysis (**Fig S8E**).
+Aspartate was the most significantly decreased metabolite with both treatments, consistent with prior reports demonstrating an important role for HIF-1 regulation of aspartate biosynthesis in cancer cells [@garcia-bermudez2018; @melendez-rodriguez2019].
 
-Principal component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen than among the BAY-treatment groups (**Figure \@ref(fig:m5)D**).
+Principal component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen than among the BAY-treatment groups (**Fig \@ref(fig:m5)D**).
 Moreover, these hypoxia-treated cells were well-segregated from BAY-treated cells.

### 2022-07-25 R/1_extracellular-fluxes.R

@@ -1,8 +1,8 @@
 # 1_extracellular-fluxes.R
 
 clean_flux_meta <- function(file_list){
-  file_list %>%
-    rlang::set_names(stringr::str_extract(basename(.), "^.+(?=_)")) |>
+  file_list |>
+    {\(x) rlang::set_names(x, stringr::str_extract(basename(x), "^.+(?=_)"))}() |>
     purrr::map_dfr(
       readr::read_csv,
       col_types = "dccccdc",
@@ -26,7 +26,7 @@ clean_fluxes <- function(data_list){
   df <-
     data_list[c("dna", "glc", "lac", "pyr")] |>
     dplyr::bind_rows(.id = "metabolite") |>
-    dplyr::filter(!is.na(.data$a)) %>%
+    dplyr::filter(!is.na(.data$a)) |>
     dplyr::select(where(~any(!is.na(.)))) |>
     clean_technical_replicates() |>
     tidyr::separate(.data$experiment, c("cell_type", "experiment", "batch", "date"), "_")
@@ -152,8 +152,8 @@ clean_flux_std <- function(df){
       "glycine", "05", "2017-11-06", "a", "b", "fld", 10
     )
 
-  df %>%
-    dplyr::filter(!(detector == "fld" & conc > 900)) %>%
+  df |>
+    dplyr::filter(!(detector == "fld" & conc > 900)) |>
     dplyr::anti_join(outliers, by = c(
       "metabolite",
       "experiment",
@@ -162,7 +162,7 @@ clean_flux_std <- function(df){
       "run",
       "detector",
       "conc"
-    )) %>%
+    )) |>
     make_std_curves()
 }
 
@@ -501,8 +501,8 @@ plot_raw_curves <- function(data, title, y, xlab = "Time (h)", ylab, ...){
 }
 
 plot_growth_curves <- function(flux_measurements){
-  flux_measurements %>%
-    dplyr::filter(metabolite == "cells") %>%
+  flux_measurements |>
+    dplyr::filter(metabolite == "cells") |>
     plot_masses(
       plot_raw_curves,
       y = conc,
@@ -512,8 +512,8 @@ plot_growth_curves <- function(flux_measurements){
 }
 
 plot_degradation_curves <- function(flux_measurements){
-  flux_measurements %>%
-    dplyr::filter(type == "empty") %>%
+  flux_measurements |>
+    dplyr::filter(type == "empty") |>
[…]

### 2022-07-25 R/1_figures.R

@@ -187,18 +187,10 @@ plot_growth_rates <- function(
       label = annot_p(p.value)
     )
 
-  x |>
-    dplyr::group_by(date) |>
-    dplyr::mutate(
-      grand_mean = mean(x$mu),
-      group_mean = mean(mu),
-      adj = group_mean - grand_mean,
-      mu_corr = mu + adj
-    ) |>
-    ggplot2::ggplot() +
+  ggplot2::ggplot(x) +
     ggplot2::aes(
       x = group,
-      y = mu_corr
+      y = mu
     ) +
     ggplot2::stat_summary(
       ggplot2::aes(fill = group),
@@ -528,19 +520,10 @@ plot_high_fluxes <- function(
       label = annot_p(p.value)
     )
 
-  x |>
-    dplyr::group_by(abbreviation) |>
-    dplyr::mutate(grand_mean = mean(flux)) |>
-    dplyr::group_by(abbreviation, date) |>
-    dplyr::mutate(
-      exp_mean = mean(flux),
-      adj = exp_mean - grand_mean,
-      flux_corr = flux - adj
-    ) |>
-    ggplot2::ggplot() +
+  ggplot2::ggplot(x) +
     ggplot2::aes(
       x = reorder(toupper(abbreviation), flux),
-      y = flux_corr
+      y = flux
     ) +
     ggplot2::stat_summary(
       ggplot2::aes(fill = group),
@@ -647,16 +630,7 @@ plot_low_fluxes <- function(
       label = annot_p(p.value)
     )
 
-  x |>
-    dplyr::group_by(abbreviation) |>
-    dplyr::mutate(grand_mean = mean(flux)) |>
-    dplyr::group_by(abbreviation, date) |>
-    dplyr::mutate(
-      exp_mean = mean(flux),
-      adj = exp_mean - grand_mean,
-      flux_corr = flux - adj
-    ) |>
-    ggplot2::ggplot() +
+  ggplot2::ggplot(x) +
     ggplot2::aes(
       x = reorder(toupper(abbreviation), flux),
       y = flux
@@ -1020,19 +994,10 @@ plot_m5_citrate <- function(df) {
       label = annot_p(p.value)
     )
 
-  x |>
-    dplyr::group_by(cell_type) |>
-    dplyr::mutate(grand_mean = mean(mid)) |>
-    dplyr::group_by(cell_type, date) |>
-    dplyr::mutate(
-      exp_mean = mean(mid),
[…]

### 2022-07-25 R/1_metab.R

@@ -49,8 +49,8 @@ tbl_to_se <- function(tbl_se, assay_name){
       rt_min = min(rt, na.rm = TRUE),
       rt_max = max(rt, na.rm = TRUE)
     ) |>
-    tibble::column_to_rownames(attr(tbl_se, "f_names")) %>%
-    .[match(rownames(assay_data), rownames(.)), ]
+    tibble::column_to_rownames(attr(tbl_se, "f_names")) |>
+    {\(x) x[match(rownames(assay_data), rownames(x)), ]}()
 
   sample_data <-
     tbl_se |>
@@ -59,8 +59,8 @@ tbl_to_se <- function(tbl_se, assay_name){
       attr(tbl_se, "s_data")
     ) |>
     dplyr::distinct() |>
-    tibble::column_to_rownames(attr(tbl_se, "s_names")) %>%
-    .[match(colnames(assay_data), rownames(.)), ]
+    tibble::column_to_rownames(attr(tbl_se, "s_names")) |>
+    {\(x) x[match(colnames(assay_data), rownames(x)), ]}()
 
   SummarizedExperiment::SummarizedExperiment(
     assays = assay_data,
@@ -221,21 +221,21 @@ log_transform <- function(pqn){
 
 annot_metabs <- function(se){
   fdata <-
-    SummarizedExperiment::rowData(se) %>%
-    data.frame() %>%
+    SummarizedExperiment::rowData(se) |>
+    data.frame() |>
     tibble::as_tibble(rownames = "HMDB")
 
   ah <- AnnotationHub::AnnotationHub()
   df <-
-    ah[["AH91792"]] %>%
-    dplyr::filter(HMDB %in% fdata$HMDB) %>%
-    dplyr::group_by(HMDB) %>%
-    dplyr::arrange(HMDB, KEGG, ChEBI, .by_group = TRUE) %>%
-    dplyr::slice(1) %>%
+    ah[["AH91792"]] |>
+    dplyr::filter(HMDB %in% fdata$HMDB) |>
+    dplyr::group_by(HMDB) |>
+    dplyr::arrange(HMDB, KEGG, ChEBI, .by_group = TRUE) |>
+    dplyr::slice(1) |>
     dplyr::select(-Name)
 
   annot_fdata <-
-    dplyr::left_join(fdata, df, by ="HMDB") %>%
+    dplyr::left_join(fdata, df, by ="HMDB") |>
     tibble::column_to_rownames("HMDB")
 
   SummarizedExperiment::rowData(se) <- annot_fdata
[…]

### 2022-07-25 _targets.R

@@ -29,7 +29,8 @@ tar_option_set(
   # packages = c("tidyverse", "patchwork", "xcms"),
   # imports = c("rnaseq.lf.hypoxia.molidustat"),
   format = "qs",
-  error = "continue"
+  # error = "continue"
+  error = NULL
 )
 
 # targets -----------------------------------------------------------------
@@ -572,7 +573,7 @@ list(
   ),
   tar_target(
     metab_tar_pca_plot,
-    plot_metab_pca(metab_tar_pca, metab_tar_clean)
+    plot_metab_pca(metab_tar_clean, metab_tar_pca)
   ),
   tar_target(
     metab_tar_limma,
@@ -591,7 +592,8 @@ list(
     values = list(
       names = list("hyp", "bay", "hyp_bay", "int"),
       colors = list(clrs[2:1], clrs[4:3], clrs[2:1], clrs[c(2, 4)]),
-      xlab = list("Hypoxia/Normoxia", "BAY/DMSO", "Hypoxia/Normoxia", "ΔHypoxia/ΔBAY")
+      xlab = list("Hypoxia/Normoxia", "BAY/DMSO", "Hypoxia/Normoxia", "ΔHypoxia/ΔBAY"),
+      title = list("Hypoxia", "BAY", "Hypoxia/Normoxia in BAY", "ΔHypoxia/ΔBAY")
     ),
     names = names,
     tar_target(
@@ -606,12 +608,16 @@ list(
       metab_tar_msea,
       run_msea(metab_tar_res, metab_pathways)
     ),
+    tar_target(
+      metab_tar_msea_table,
+      plot_msea_table(metab_tar_msea, title, colors, names)
+    ),
     NULL
   ),
-  # tar_target(
-  #   metab_venn,
-  #   plot_metab_venn(metab_hyp, metab_bay)
-  # ),
+  tar_target(
+    metab_venn,
+    plot_metab_venn(metab_tar_res_hyp, metab_tar_res_bay)
+  ),
   tar_render(
     metabolomics_report,
     path = path_to_reports("metabolomics-targeted.Rmd"),
@@ -1151,6 +1157,75 @@ list(
   tar_target(
     m4_figure,
     write_figures(m4, "m4.png")
+  ),
+
+  # m5 ----------------------------------------------------------------------
+
+  tar_target(
+    hyp_bay_fluxes_stats,
[…]

### 2022-07-26 manuscript/manuscript.Rmd
finish up metabolomics results and figures

@@
 
-Principal component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen than among the BAY-treatment groups (**Fig \@ref(fig:m5)D**).
-Moreover, these hypoxia-treated cells were well-segregated from BAY-treated cells.
-These observations are, again, consistent with the results of the metabolic flux models demonstrating an overriding effect of hypoxia *per se* on metabolic flux and highlighting important differences between hypoxic and pharmacologic PHD inhibition.
-To identify the key differences among these treatments, we performed linear modeling focusing on those metabolites with a significant interaction term in the model described by treatment × oxygen (**Figure \@ref(fig:m5)E**) (*i.e.*, metabolites where the change with hypoxia treatment was different from the change with BAY treatment).
-Of 133 metabolites, 77 were significantly differentially regulated by hypoxia and BAY treatments.
-Several patterns emerged from this analysis (**Figure \@ref(fig:m5)F**).
-Interestingly, the hypoxia-regulated metabolite, 2-hydroxyglutarate [@RN1603], increased following BAY treatment, but was unaffected by hypoxia treatment in proliferating LFs.
-The tricarboxylic acid (TCA) cycle metabolite aconitate was decreased in hypoxia, but increased by BAY treatment, while the opposite effect was observed for hydroxyproline and taurine.
-The glycolytic intermediate glyceraldehyde 3-phosphate (GAP) was decreased by both hypoxia and BAY, while hypoxia reversed the BAY-mediated decrease in γ-aminobutyric acid (GABA).
-A metabolite set enrichment analysis of these differentially regulated metabolites revealed the TCA cycle to be the most enriched KEGG metabolite set (**Figure \@ref(fig:m5)G**).
-Leading edge analysis demonstrated negative enrichment scores associated with all of the TCA metabolites detected by our platform (**Figure \@ref(fig:m5)H**).
-This result indicates a more modest impact of BAY treatment on the TCA cycle than hypoxia, as suggested by our metabolic flux models where hypoxia resulted in a 1.5-2-fold reduction of TCA flux compared to a 1.1-1.5-fold reduction with BAY treatment in normoxia (**Figure \@ref(fig:m3)**).
+Principal component analysis revealed greater similarity among both treatment groups cultured in 0.5% oxygen than among the BAY-treatment groups (**Fig \@ref(fig:m5)D**).
+Moreover, these hypoxia-treated cells were well-segregated from cells treated with BAY alone.
+These observations are, again, consistent with the results of the metabolic flux models demonstrating an overriding effect of hypoxia *per se* on cell metabolism and highlighting important differences between hypoxic and pharmacologic PHD inhibition.
+To identify the metabolic changes that depend on hypoxia rather than BAY inhibition, we identified differentially regulated metabolites in BAY treated cells cultured in normoxia and hypoxia (**Fig \@ref(fig:m5)E**).
+Of 133 metabolites, 83 were significantly differentially regulated by hypoxia in BAY-treated cells treatments.
+An enrichment analysis of these differentially regulated metabolites demonstrated up-regulation of arginine and proline metabolism and down-regulation of the TCA cycle as the most impacted by hypoxia in BAY treated cells (**Fig \@ref(fig:m5)F**).
+Indeed, leading edge analysis highlights negative enrichment scores associated with all of the TCA metabolites detected by our platform (**Fig \@ref(fig:m5)G**).
+This result indicates better preservation of TCA cycle flux in normoxic BAY-treated cells than in hypoxic cells, as suggested by our metabolic flux models where hypoxia resulted in a 1.5-2-fold reduction of TCA flux compared to a 1.1-1.5-fold reduction with BAY treatment  (**Fig \@ref(fig:m3)**).
 
-In addition to these differential effects on polar metabolite levels, perhaps the most significant contrast between hypoxia and BAY treatment is the impact of hypoxia on cellular redox state.
-As oxygen deprivation causes reductive stress [@RN2409], we next measured the impact of these treatments on intracellular NAD(H) as a surrogate for cellular redox state (**Figures \@ref(fig:m5)I-K**).
+In addition to these differential effects on polar metabolite levels, we reasoned that another critical difference between hypoxia and BAY treatment is the impact of hypoxia on cellular redox state.
+As oxygen deprivation causes reductive stress [@xiao2020], we next measured the impact of these treatments on intracellular NAD(H) (**Figs \@ref(fig:m5)H-J**).
 As expected, hypoxia increased the NADH/NAD^+^ ratio, driven primarily by a decrease in intracellular NAD^+^.
 Interestingly, while BAY treatment increased the levels of NADH, a concomitant increase in NAD^+^ resulted in preservation of the NADH/NAD^+^ ratio.
-As NADH accumulation is a putative inhibitor of glycolytic flux [@RN2210], this may be one mechanism by which glycolytic flux is decreased in hypoxia but not following BAY treatment.
+As NADH accumulation is a putative inhibitor of glycolytic flux [@tilton1991], this may be one mechanism by which glycolytic flux is decreased in hypoxia but not following BAY treatment.
+
+## Transcriptomic analysis identifies regulators of metabolism in hypoxia
+
+To identify the upstream regulators of the observed metabolic changes, we next performed transcriptomic analysis of LFs treated with hypoxia or BAY, separately or together (**Fig \@ref(fig:m6), S9**).
+
+As anticipated, both hypoxia and BAY treatment induced substantial changes in gene expression (**Figure S9A, S9B**).
+Of the 7,789 differentially expressed genes across both conditions, 891 (11%) were unique to BAY treatment in normoxia, 1,649 (21%) were shared between BAY and hypoxia, while 5,249 (67%) were unique to 0.5% hypoxia culture (**Figure S9C**).
+Gene set enrichment analysis of these differentially regulated metabolites was performed using Molecular Signatures Database "Hallmark" gene sets [@RN3112] (**Figures S9D, S9E, S9F**).
+As expected, both treatments were associated with enrichment of the "hypoxia" and "glycolysis" gene sets.
+
+Given the disparate effects of hypoxic and pharmacologic PHD inhibition on cellular metabolism described above, we focused our transcriptomics analyses on the differences between hypoxia and BAY treatments.
+Principal component analysis again demonstrates clear separation among the four treatment groups (**Figure \@ref(fig:m6)A**).
+The first and second principal components correspond to 0.5% oxygen and BAY treatments, respectively.
+Consistent with our prior observations, the combination of 0.5% oxygen plus BAY was more similar to 0.5% oxygen alone with decreased distance between both hypoxia-treated groups along the axis of the second principal component, again, consistent with the hypothesis that hypoxia uncouples the effects of BAY treatment.
+To identify those transcripts that were differentially affected by hypoxia compared to BAY treatment alone, we performed linear modeling to identify transcripts with a significant interaction term in the model described by treatment × oxygen (**Figure \@ref(fig:m6)B**).
+With this analysis, we found several patterns of transcriptional changes (**Figure \@ref(fig:m6)C**).
+For example, EPAS1, the gene encoding the HIF-2α protein, was markedly decreased by hypoxia, compared to its modest reduction following BAY treatment.
+A similar, but opposite, effect on RBM3 (RNA-binding protein 3) expression was observed.
+RBM3 supports the viability and proliferation of neural stem cells in hypoxia [@RN3111].
+Histone deacetylase 9 (HDAC9) and prolyl 4-hydroxylase subunit α2 (P4HA2) demonstrated opposing effects following hypoxia and BAY treatment, where BAY decreased HDAC9 expression and increased P4HA2 expression.
+Together, this subset of transcriptional changes illustrates important differences between hypoxia and HIF stabilization in normoxia.
+
+Interestingly, an enrichment analysis of transcripts differentially regulated by 0.5% oxygen and BAY treatment actually demonstrated greater enrichment of the "hypoxia" gene set following BAY treatment alone than with hypoxia itself (**Figure \@ref(fig:m6)D**).
+Similarly, the gene set for glycolysis was also more enriched following BAY treatment than hypoxia treatment.
+The most enriched gene sets associated with hypoxia included pro-proliferative "E2F targets" and "G2/M checkpoint" proteins as well as "MYC targets".
+These findings were further supported by a transcription factor enrichment analysis identifying enrichment of MYC target proteins associated with hypoxia (**Figure \@ref(fig:m6)E**).
 

### 2022-07-26 R/1_extracellular-fluxes.R

@@ -734,7 +734,7 @@ analyze_hyp_bay_fluxes <- function(growth, fluxes) {
     dplyr::mutate(
       oxygen = factor(oxygen, levels = c("21%", "0.5%")),
       y_pos = Inf,
-      vjust = 2,
+      vjust = 1,
       lab = annot_p(adj.p.value)
     )
 

### 2022-07-26 R/1_figures.R

@@ -1393,6 +1393,108 @@ plot_hyp_bay_fluxes <- function(df, annot, metab, ylab) {
     NULL
 }
 
+plot_nad <- function(df, annot, metab, ylab) {
+  annot <- dplyr::filter(annot, measurement == metab)
+  annot1 <-
+    dplyr::filter(annot, oxygen != ".") |>
+    dplyr::mutate(
+      treatment = "BAY",
+      treatment = factor(treatment, levels = c("DMSO", "BAY")),
+      oxygen = factor(oxygen, levels = c("21%", "0.5%"))
+    )
+  annot2 <-
+    dplyr::filter(annot, treatment != ".") |>
+    dplyr::mutate(
+      oxygen = "21%",
+      treatment = factor(treatment, levels = c("DMSO", "BAY")),
+      oxygen = factor(oxygen, levels = c("21%", "0.5%"))
+    )
+
+  df |>
+    dplyr::filter(treatment != "None" & measurement == metab) |>
+    ggplot2::ggplot() +
+    ggplot2::aes(
+      x = treatment,
+      y = value,
+      fill = oxygen
+    ) +
+    ggplot2::stat_summary(
+      # ggplot2::aes(fill = treatment),
+      geom = "col",
+      fun = "mean",
+      # width = 0.6,
+      position = ggplot2::position_dodge2(),
+      show.legend = TRUE,
+      alpha = 0.5
+    ) +
+    ggbeeswarm::geom_beeswarm(
+      # ggplot2::aes(fill = treatment),
+      dodge.width = 0.9,
+      pch = 21,
+      size = 1,
+      stroke = 0.25,
+      cex = 4,
+      color = "white",
+      show.legend = FALSE
+    ) +
+    ggplot2::stat_summary(
+      # ggplot2::aes(group = treatment),
+      geom = "errorbar",
+      fun.data = ggplot2::mean_se,
+      position = ggplot2::position_dodge(width = 0.9),
+      width = 0.2,
+      size = 0.25,
+      show.legend = FALSE
+    ) +
+    ggplot2::geom_text(
+      data = annot1,
+      ggplot2::aes(
+        color = oxygen,
+        y = y,
+        label = lab,
+        vjust = vjust
+      ),
+      family = "Calibri",
+      size = 8/ggplot2::.pt,
+      position = ggplot2::position_dodge(width = 0.6),
[…]

### 2022-07-26 R/1_metab.R

@@ -387,7 +387,7 @@ plot_metab_volcano <- function(
     colors = NULL,
     xlab = NULL,
     nudge = 5.5
-    ) {
+) {
   left <-
     results |>
     dplyr::filter(logFC < 0) |>
@@ -459,7 +459,7 @@ plot_metab_volcano <- function(
     ggplot2::scale_y_continuous(
       trans = c("log10", "reverse"),
       labels = scales::label_log()
-      ) +
+    ) +
     ggplot2::scale_x_continuous(
       breaks = scales::pretty_breaks(n = 7),
       limits = c(-nudge - 0.25, nudge + 0.25),
@@ -682,4 +682,124 @@ plot_msea_table <- function(df, title, clr, filename) {
     filename = path_to_manuscript(stringr::str_c("ai/msea_", filename, ".png"))
   )
   tab
+}
+
+plot_leading_edge <- function(tt, pathways, nm) {
+
+  x <-
+    tt |>
+    dplyr::filter(!is.na(KEGG)) |>
+    dplyr::mutate(KEGG = stringr::str_c("kegg:", KEGG))
+
+  stats <-
+    x |>
+    dplyr::select(KEGG, t) |>
+    tibble::deframe()
+
+  rnk <- rank(-stats)
+  ord <- order(rnk)
+
+  stats_adj <- stats[ord]
+  stats_adj <- stats_adj / max(abs(stats_adj))
+
+  pathway <- pathways[[nm]]
+  pathway <- unname(as.vector(na.omit(match(pathway, names(stats_adj)))))
+  pathway <- sort(pathway)
+
+  gsea_res <-
+    fgsea::calcGseaStat(
+      stats_adj,
+      selectedStats = pathway,
+      returnAllExtremes = TRUE
+    )
+
+  bottoms <- gsea_res$bottoms
+  tops <- gsea_res$tops
+
+  n <- length(stats_adj)
+  xs <- as.vector(rbind(pathway - 1, pathway))
+  ys <- as.vector(rbind(bottoms, tops))
+  nms <- as.vector(rbind(names(bottoms), names(tops)))
+  df <-
+    tibble::tibble(
+      x = c(0, xs, n + 1),
+      y = c(0, ys, 0),
+      names = c(NA_character_, nms, NA_character_)
+    ) |>
+    dplyr::left_join(
+      dplyr::select(x, metabolite, KEGG), by = c("names" = "KEGG")
+    )
+
+  annot <-
+    df |>
[…]

### 2022-07-26 R/1_nad.R

@@ -57,95 +57,6 @@ annot_nad <- function(nad_final){
     dplyr::mutate(
       lab = annot_p(pval),
       y = Inf,
-      vjust = 1.5
+      vjust = 1
     )
-}
-
-plot_nad <- function(nad_final, annot, metab, ylab){
-
-  annot <- dplyr::filter(annot, measurement == metab)
-
-  annot1 <-
-    dplyr::filter(annot, oxygen != ".") |>
-    dplyr::mutate(
-      treatment = "BAY",
-      treatment = factor(treatment, levels = c("DMSO", "BAY")),
-      oxygen = factor(oxygen, levels = c("21%", "0.5%"))
-    )
-  annot2 <-
-    dplyr::filter(annot, treatment != ".") |>
-    dplyr::mutate(
-      oxygen = "21%",
-      treatment = factor(treatment, levels = c("DMSO", "BAY")),
-      oxygen = factor(oxygen, levels = c("21%", "0.5%"))
-    )
-
-  nad_final |>
-    dplyr::filter(treatment != "None" & measurement == metab) |>
-    ggplot2::ggplot() +
-    ggplot2::aes(
-      x = treatment,
-      y = value,
-      fill = oxygen
-    ) +
-    ggplot2::stat_summary(
-      geom = "col",
-      fun = "mean",
-      position = ggplot2::position_dodge(width = 0.6),
-      width = 0.6,
-      show.legend = TRUE
-    ) +
-    ggplot2::stat_summary(
-      ggplot2::aes(group = oxygen),
-      geom = "errorbar",
-      fun.data = "mean_se",
-      position = ggplot2::position_dodge(width = 0.6),
-      width = 0.2,
-      size = 0.25,
-      show.legend = FALSE,
-      color = "black"
-    ) +
-    ggplot2::geom_text(
-      data = annot1,
-      ggplot2::aes(
-        color = oxygen,
-        y = y,
-        label = lab,
-        vjust = vjust
-      ),
-      family = "Calibri",
-      size = 6/ggplot2::.pt,
-      position = ggplot2::position_dodge(width = 0.6),
-      show.legend = FALSE
-    ) +
-    ggplot2::geom_text(
-      data = annot2,
-      ggplot2::aes(
-        y = y,
-        label = lab,
-        vjust = vjust
-      ),
-      family = "Calibri",
-      size = 6/ggplot2::.pt,
-      color = "black",
[…]

### 2022-07-26 _targets.R

@@ -478,6 +478,18 @@ list(
     nad_annot,
     annot_nad(nad_final)
   ),
+  tar_target(
+    nad_plot,
+    plot_nad(nad_final, nad_annot, "NAD", "NAD\n(nmol/cell)")
+  ),
+  tar_target(
+    nadh_plot,
+    plot_nad(nad_final, nad_annot, "NADH", "NADH\n(nmol/cell)")
+  ),
+  tar_target(
+    nadh_ratio_plot,
+    plot_nad(nad_final, nad_annot, "Ratio", "NADH/NAD ratio")
+  ),
 
   # rnaseq ------------------------------------------------------------------
 
@@ -612,6 +624,18 @@ list(
       metab_tar_msea_table,
       plot_msea_table(metab_tar_msea, title, colors, names)
     ),
+    tar_target(
+      msea,
+      patchwork::wrap_elements(
+        full =
+          plot_image(
+            path_to_manuscript(stringr::str_c("ai/msea_", names, ".png")),
+            vjust = 0
+          ) +
+          ggplot2::coord_fixed()
+      ) +
+        theme_plots()
+    ),
     NULL
   ),
   tar_target(
@@ -623,6 +647,22 @@ list(
     path = path_to_reports("metabolomics-targeted.Rmd"),
     output_dir = system.file("analysis/pdfs", package = "Copeland.2022.hypoxia.flux")
   ),
+  tar_target(
+    tca_leading_edge,
+    plot_leading_edge(
+      metab_tar_res_hyp_bay,
+      metab_pathways,
+      "KEGG | Citrate cycle (TCA cycle) - Homo sapiens (human)"
+    )
+  ),
+  # tar_target(
+  #   pro_leading_edge,
+  #   plot_leading_edge(
+  #     metab_tar_res_hyp_bay,
+  #     metab_pathways,
+  #     "KEGG | Arginine and proline metabolism - Homo sapiens (human)"
+  #   )
+  # ),
 
   # myc ---------------------------------------------------------------------
 
@@ -1185,7 +1225,11 @@ list(
       hyp_bay_fluxes_lac,
       metab_tar_pca_plot,
       metab_tar_vol_hyp_bay,
-      metab_tar_msea_hyp_bay
+      msea_hyp_bay,
+      tca_leading_edge,
+      nad_plot,
+      nadh_plot,
+      nadh_ratio_plot
     )
   ),
   tar_target(
@@ -1195,24 +1239,6 @@ list(
 
[…]

### 2022-07-26 manuscript/manuscript.Rmd
complete transcriptomics results and figures

@@
 
-To identify the upstream regulators of the observed metabolic changes, we next performed transcriptomic analysis of LFs treated with hypoxia or BAY, separately or together (**Fig \@ref(fig:m6), S9**).
-
-As anticipated, both hypoxia and BAY treatment induced substantial changes in gene expression (**Figure S9A, S9B**).
-Of the 7,789 differentially expressed genes across both conditions, 891 (11%) were unique to BAY treatment in normoxia, 1,649 (21%) were shared between BAY and hypoxia, while 5,249 (67%) were unique to 0.5% hypoxia culture (**Figure S9C**).
-Gene set enrichment analysis of these differentially regulated metabolites was performed using Molecular Signatures Database "Hallmark" gene sets [@RN3112] (**Figures S9D, S9E, S9F**).
+To identify the upstream regulators of the observed metabolic changes, we next performed RNA-seq transcriptomic analysis of LFs treated with hypoxia or BAY, separately or together (**Fig \@ref(fig:m6), S9**).
+As anticipated, both hypoxia and BAY treatment induced substantial changes in gene expression (**Fig S9A, S9B**).
+Of the 9,923 differentially expressed genes across both conditions, 869 (9%) were unique to BAY treatment in normoxia, 4,002 (40%) were shared by BAY and hypoxia, while 5,052 (51%) were unique to 0.5% hypoxia culture (**Fig S9C**).
+This distribution of transcriptional changes was nearly identical to that observed for the metabolic changes (**Fig S8C**) where half of the hypoxia-mediated changes overlapped with BAY treatment and half were unique to hypoxia.
+Gene set enrichment analysis of these differentially regulated metabolites was performed using Molecular Signatures Database "Hallmark" gene sets [@liberzon2015] (**Figs S9D, S9E, S9F**).
 As expected, both treatments were associated with enrichment of the "hypoxia" and "glycolysis" gene sets.
@@
 Given the disparate effects of hypoxic and pharmacologic PHD inhibition on cellular metabolism described above, we focused our transcriptomics analyses on the differences between hypoxia and BAY treatments.
-Principal component analysis again demonstrates clear separation among the four treatment groups (**Figure \@ref(fig:m6)A**).
+Principal component analysis again demonstrated clear separation among the four treatment groups (**Fig \@ref(fig:m6)A**).
 The first and second principal components correspond to 0.5% oxygen and BAY treatments, respectively.
-Consistent with our prior observations, the combination of 0.5% oxygen plus BAY was more similar to 0.5% oxygen alone with decreased distance between both hypoxia-treated groups along the axis of the second principal component, again, consistent with the hypothesis that hypoxia uncouples the effects of BAY treatment.
-To identify those transcripts that were differentially affected by hypoxia compared to BAY treatment alone, we performed linear modeling to identify transcripts with a significant interaction term in the model described by treatment × oxygen (**Figure \@ref(fig:m6)B**).
-With this analysis, we found several patterns of transcriptional changes (**Figure \@ref(fig:m6)C**).
-For example, EPAS1, the gene encoding the HIF-2α protein, was markedly decreased by hypoxia, compared to its modest reduction following BAY treatment.
-A similar, but opposite, effect on RBM3 (RNA-binding protein 3) expression was observed.
-RBM3 supports the viability and proliferation of neural stem cells in hypoxia [@RN3111].
-Histone deacetylase 9 (HDAC9) and prolyl 4-hydroxylase subunit α2 (P4HA2) demonstrated opposing effects following hypoxia and BAY treatment, where BAY decreased HDAC9 expression and increased P4HA2 expression.
-Together, this subset of transcriptional changes illustrates important differences between hypoxia and HIF stabilization in normoxia.
+Consistent with our prior observations, the combination of 0.5% oxygen plus BAY was more similar to 0.5% oxygen alone with decreased distance between both hypoxia-treated groups along the axis of the second principal component, again, consistent with the hypothesis that hypoxia overrides the effects of BAY treatment.
+To identify the transcripts driving these differences, we identified genes differentially expressed following hypoxia in BAY-treated cells (**Fig \@ref(fig:m6)B**).
+Interestingly, an enrichment analysis of these differentially expressed transcripts identified pro-proliferative gene sets like "E2F targets", "G2/M checkpoint", and "MYC targets" associated with hypoxia (**Fig \@ref(fig:m6)C-E**).
+These findings were further supported by a transcription factor enrichment analysis identifying enrichment of MYC transcription factor activity associated with hypoxia, but not BAY treatment (**Fig \@ref(fig:m6)F**).
+Classically, hypoxia and HIF activation are thought to inhibit cell proliferation by inhibiting pro-proliferative MYC signaling [@koshiji2004].
+These results indicate that hypoxia-induced MYC activation may be sustaining proliferation in these LFs.
+We reasoned that these pro-proliferative signals may also account for the unexpected effects of hypoxia on glycolysis that we observed.
 
-Interestingly, an enrichment analysis of transcripts differentially regulated by 0.5% oxygen and BAY treatment actually demonstrated greater enrichment of the "hypoxia" gene set following BAY treatment alone than with hypoxia itself (**Figure \@ref(fig:m6)D**).
-Similarly, the gene set for glycolysis was also more enriched following BAY treatment than hypoxia treatment.
-The most enriched gene sets associated with hypoxia included pro-proliferative "E2F targets" and "G2/M checkpoint" proteins as well as "MYC targets".
-These findings were further supported by a transcription factor enrichment analysis identifying enrichment of MYC target proteins associated with hypoxia (**Figure \@ref(fig:m6)E**).
+## MYC antagonizes HIF-dependent glycolytic fluxes
+
+To test the role of hypoxia-induced MYC activation in the metabolic response to hypoxia in proliferating primary cells, we first measured MYC protein levels by immunoblot.
+
+Consistent with our bioinformatic results, immunoblotting demonstrated increased MYC protein levels in hypoxia-treated cells, but not with BAY-treatment alone, where MYC was decreased (**Figure \@ref(fig:m7)A-B**).
+To test the hypothesis that hypoxia-induced MYC expression inhibits glycolysis in hypoxic primary cells, we first combined MYC knockdown with hypoxia treatment (**Figure \@ref(fig:m7)C-E**).
+As expected, MYC-deficient cells proliferated more slowly in normoxia and MYC was absolutely essential for sustaining cell proliferation in hypoxia (**Figure \@ref(fig:m7)D**).
+Consistent with our hypothesis, MYC knockdown cells demonstrated increased lactate efflux upon hypoxia treatment, unlike control siRNA-treated cells (**Figure \@ref(fig:m7)E**).
+We next performed the complementary experiment to determine whether MYC overexpression could attenuate the increase in glycolysis observed with BAY treatment (**Figure \@ref(fig:m7)F-H**).
+MYC increased the proliferation rate of DMSO-treated cells, although it did not augment the proliferation rate of BAY-treated cells.
+As expected, MYC overexpression blocked the BAY-stimulated increase in lactate efflux.
+Together, these data suggest that hypoxia-induced MYC expression may be one factor that uncouples the HIF transcriptional program from glycolytic flux in proliferating primary cells.
 

### 2022-07-26 R/1_figures.R

@@ -119,6 +119,7 @@ plot_time_lines <- function(
       pch = 21,
       color = "white",
       size = 1.5,
+      stroke = 0.2,
       show.legend = FALSE
     ) +
     ggplot2::labs(
@@ -204,7 +205,7 @@ plot_growth_rates <- function(
       ggplot2::aes(fill = group),
       pch = 21,
       size = 1,
-      stroke = 0.25,
+      stroke = 0.2,
       cex = 4,
       color = "white",
       show.legend = FALSE
@@ -294,7 +295,8 @@ plot_cells_per_dna <- function(dna_per_cell_clean, cell = c("lf", "pasmc")) {
       color = "white",
       fill = "black",
       size = 1.5,
-      show.legend = FALSE
+      show.legend = FALSE,
+      stroke = 0.2
     ) +
     ggplot2::labs(
       x = "Cell count",
@@ -328,7 +330,9 @@ plot_dna_count_hypoxia <- function(dna_count_hypoxia) {
       color = "white",
       # alpha = 0.3,
       size = 1.5,
-      show.legend = FALSE) +
+      stroke = 0.2,
+      show.legend = FALSE
+      ) +
     ggplot2::labs(
       x = "Cell count",
       y = "DNA (ng)"
@@ -386,6 +390,7 @@ plot_evap_data <- function(evap_clean) {
       pch = 21,
       color = "white",
       size = 1.5,
+      stroke = 0.2,
       show.legend = FALSE
     ) +
     ggplot2::labs(
@@ -544,7 +549,7 @@ plot_high_fluxes <- function(
       dodge.width = 0.9,
       pch = 21,
       size = 1,
-      stroke = 0.25,
+      stroke = 0.2,
       cex = 4,
       color = "white",
       show.legend = FALSE
@@ -1015,7 +1020,7 @@ plot_m5_citrate <- function(df) {
       ggplot2::aes(fill = oxygen),
       pch = 21,
       size = 1,
-      stroke = 0.25,
+      stroke = 0.2,
       cex = 4,
       color = "white",
       show.legend = FALSE
@@ -1118,6 +1123,7 @@ plot_mid_time_course <- function(df, cells, o2, treat, color) {
       pch = 21,
       color = "white",
       size = 1.5,
+      stroke = 0.2,
       show.legend = TRUE
     ) +
[…]

### 2022-07-26 R/1_metab.R

@@ -304,7 +304,8 @@ plot_metab_pca <- function(clean, df) {
       pch = 21,
       color = "white",
       size = 2,
-      show.legend = FALSE
+      show.legend = FALSE,
+      stroke = 0.2
     ) +
     ggplot2::labs(
       x = paste0("PC1: ", percent_variance[1], "% variance"),
@@ -441,19 +442,22 @@ plot_metab_volcano <- function(
       data = subset(results, adj.P.Val > 0.05),
       pch = 21,
       color = "white",
-      fill = "grey80"
+      fill = "grey80",
+      stroke = 0.2
     ) +
     ggplot2::geom_point(
       data = subset(results, logFC > 0 & adj.P.Val < 0.05),
       pch = 21,
       color = "white",
-      fill = colors[[1]]
+      fill = colors[[1]],
+      stroke = 0.2
     ) +
     ggplot2::geom_point(
       data = subset(results, logFC < 0 & adj.P.Val < 0.05),
       pch = 21,
       color = "white",
-      fill = colors[[2]]
+      fill = colors[[2]],
+      stroke = 0.2
     ) +
     ggplot2::scale_color_manual(values = c("black", "darkred")) +
     ggplot2::scale_y_continuous(

### 2022-07-26 R/1_rnaseq.R

@@ -86,6 +86,7 @@ plot_rnaseq_pca <- function(pca_data){
       pch = 21,
       color = "white",
       size = 2,
+      stroke = 0.2,
       show.legend = FALSE
     ) +
     ggplot2::labs(
@@ -107,7 +108,9 @@ plot_rnaseq_pca <- function(pca_data){
     theme_plots() +
     ggplot2::theme(
       legend.position = "bottom",
-      legend.key.size = ggplot2::unit(1, units = "lines")
+      legend.key.size = ggplot2::unit(1, units = "lines"),
+      axis.line = ggplot2::element_blank(),
+      panel.border = ggplot2::element_rect(color = "black", fill = NA, size = 0.25)
     )
 }
 
@@ -146,7 +149,7 @@ identify_deg <- function(dds, comp){
     dplyr::arrange(padj)
 }
 
-plot_rnaseq_volcano <- function(results, gois = NULL, xlab = NULL){
+plot_rnaseq_volcano <- function(results, gois = NULL, xlab = NULL, nudge = 8){
   df <-
     tibble::as_tibble(results)
 
@@ -174,7 +177,7 @@ plot_rnaseq_volcano <- function(results, gois = NULL, xlab = NULL){
       size = 5/ggplot2::.pt,
       max.overlaps = 20,
       segment.size = 0.1,
-      nudge_x = -8 - left$log2FoldChange,
+      nudge_x = -nudge - left$log2FoldChange,
       hjust = 0,
       direction = "y",
       family = "Calibri",
@@ -190,7 +193,7 @@ plot_rnaseq_volcano <- function(results, gois = NULL, xlab = NULL){
       size = 5/ggplot2::.pt,
       max.overlaps = 20,
       segment.size = 0.1,
-      nudge_x = 8 - right$log2FoldChange,
+      nudge_x = nudge - right$log2FoldChange,
       hjust = 1,
       direction = "y",
       segment.color = "black",
@@ -198,23 +201,30 @@ plot_rnaseq_volcano <- function(results, gois = NULL, xlab = NULL){
       show.legend = FALSE
     ) +
     ggplot2::geom_hex(
-      bins = c(250, 15),
+      bins = c(50, 2),
       show.legend = FALSE
     ) +
     ggplot2::scale_fill_viridis_c(trans = "log10") +
[…]

### 2022-07-26 R/1_utils.R

@@ -184,3 +184,9 @@ plot_image <- function(img, scale = 1, hjust = 0, vjust = 0) {
       axis.line = ggplot2::element_blank()
     )
 }
+
+write_table <- function(table, filename) {
+  file <- path_to_manuscript(stringr::str_c("ai/", filename, ".png"))
+  gt::gtsave(table, filename = file)
+  file
+}

### 2022-07-26 _targets.R

@@ -30,7 +30,8 @@ tar_option_set(
   # imports = c("rnaseq.lf.hypoxia.molidustat"),
   format = "qs",
   # error = "continue"
-  error = NULL
+  error = NULL,
+  # debug = "gsea_deg_oxphos"
 )
 
 # targets -----------------------------------------------------------------
@@ -529,9 +530,11 @@ list(
       xlab = c(
         "Hypoxia/Normoxia",
         "BAY/DMSO",
-        "Hypoxia/Normoxia",
+        "Hypoxia/Normoxia in BAY",
         "ΔHypoxia/ΔBAY"
-      )
+      ),
+      colors = list(clrs[2:1], clrs[4:3], clrs[2:1], clrs[c(2, 4)]),
+      filename = stringr::str_c("gsea_", c("hyp", "bay", "hyp_bay", "int"))
     ),
     names = names,
     tar_target(
@@ -540,17 +543,44 @@ list(
     ),
     tar_target(
       rnaseq_vol,
-      plot_rnaseq_volcano(deg, xlab)
+      plot_rnaseq_volcano(deg, xlab = xlab)
     ),
     tar_target(
       gsea,
       run_gsea(deg, hallmark_pathways)
     ),
     tar_target(
+      gsea_table,
+      plot_gsea_table(gsea, title = xlab, clr = colors)
+    ),
+    tar_target(
+      gsea_table_file,
+      write_table(gsea_table, filename),
+      format = "file"
+    ),
+    tar_target(
+      gsea_table_plot,
+      patchwork::wrap_elements(full = plot_image(gsea_table_file, vjust = 0)) +
+        ggplot2::coord_fixed() +
+        theme_plots()
+    ),
+    tar_target(
       tfea_res,
       index_tfea(tfea_fit, names)
     ),
+    tar_target(
+      tfea_plot,
+      plot_tfea_volcanoes(tfea_res, xlab = xlab, colors = colors, nudge = 10)
+    ),
     NULL
+  ),
+  tar_target(
+    rnaseq_venn,
+    plot_rnaseq_venn(deg_hyp, deg_bay)
+  ),
+  tar_target(
+    gsea_venn,
+    plot_gsea_venn(gsea_hyp, gsea_bay)
   ),
   tar_render(
     rnaseq_report,
@@ -605,7 +635,8 @@ list(
       names = list("hyp", "bay", "hyp_bay", "int"),
       colors = list(clrs[2:1], clrs[4:3], clrs[2:1], clrs[c(2, 4)]),
[…]

### 2022-07-28 manuscript/manuscript.Rmd
add simyc and oemyc results and figure

@@
 To test the role of hypoxia-induced MYC activation in the metabolic response to hypoxia in proliferating primary cells, we first measured MYC protein levels by immunoblot.
-
-Consistent with our bioinformatic results, immunoblotting demonstrated increased MYC protein levels in hypoxia-treated cells, but not with BAY-treatment alone, where MYC was decreased (**Figure \@ref(fig:m7)A-B**).
-To test the hypothesis that hypoxia-induced MYC expression inhibits glycolysis in hypoxic primary cells, we first combined MYC knockdown with hypoxia treatment (**Figure \@ref(fig:m7)C-E**).
-As expected, MYC-deficient cells proliferated more slowly in normoxia and MYC was absolutely essential for sustaining cell proliferation in hypoxia (**Figure \@ref(fig:m7)D**).
-Consistent with our hypothesis, MYC knockdown cells demonstrated increased lactate efflux upon hypoxia treatment, unlike control siRNA-treated cells (**Figure \@ref(fig:m7)E**).
-We next performed the complementary experiment to determine whether MYC overexpression could attenuate the increase in glycolysis observed with BAY treatment (**Figure \@ref(fig:m7)F-H**).
-MYC increased the proliferation rate of DMSO-treated cells, although it did not augment the proliferation rate of BAY-treated cells.
+Consistent with our bioinformatic results, we observed increased MYC protein levels in hypoxia-treated cells, but not with BAY-treatment alone, where MYC was decreased (**Fig \@ref(fig:m7)A-B**).
+To test the hypothesis that hypoxia-induced MYC expression inhibits glycolysis in primary cells, we first combined MYC knockdown with hypoxia treatment (**Fig \@ref(fig:m7)C-E**).
+As expected, MYC-deficient cells proliferated more slowly in normoxia and MYC was essential for sustaining cell proliferation in hypoxia (**Fig \@ref(fig:m7)D**).
+Consistent with our hypothesis, MYC-knockdown cells demonstrated increased lactate efflux with hypoxia culture, unlike control siRNA-treated cells (**Fig \@ref(fig:m7)E**).
+We next performed the complementary experiment to determine whether MYC overexpression could attenuate the increase in glycolysis observed with BAY treatment (**Fig \@ref(fig:m7)F-H**).
+MYC increased the proliferation rate of DMSO-treated cells, although it did not augment the proliferation rate of BAY-treated cells (**Fig \@ref(fig:m7)G**).
 As expected, MYC overexpression blocked the BAY-stimulated increase in lactate efflux.

### 2022-07-28 R/1_expression.R

@@ -57,7 +57,7 @@ normalize_densities <- function(blot_raw){
       group = factor(group, levels = c("21%", "0.5%", "0.2%", "DMSO", "BAY"))
     ) |>
     dplyr::group_by(experiment, oxygen, treatment, group, time, protein) |>
-    wmo::remove_nested_outliers(fold_change, remove = TRUE) |>
+    # wmo::remove_nested_outliers(fold_change, remove = TRUE) |>
     dplyr::relocate(group, .after = treatment)
 }
 
@@ -96,4 +96,46 @@ normalize_qpcr <- function(raw_mrna){
     dplyr::group_by(experiment, oxygen, treatment, group, time, protein) |>
     wmo::remove_nested_outliers(fold_change, remove = TRUE) |>
     dplyr::relocate(group, .after = treatment)
+}
+
+analyze_hyp_bay_densities <- function(x, prot) {
+  df <-
+    x |>
+    dplyr::filter(experiment == "lf_05-bay") |>
+    dplyr::filter(protein == prot) |>
+    dplyr::group_by(protein, oxygen, treatment) |>
+    # wmo::remove_nested_outliers(fold_change, remove = TRUE) |>
+    {\(x) x}()
+
+  annot <-
+    df |>
+    dplyr::group_by(protein) |>
+    tidyr::nest() |>
+    dplyr::mutate(
+      m = purrr::map(data, ~lmerTest::lmer(fold_change ~ oxygen * treatment + (1 | gel), data = .x)),
+      res = purrr::map(m, ~emmeans::emmeans(
+        .x,
+        "pairwise" ~ oxygen * treatment,
+        simple = "each",
+        adjust = "mvt",
+        combine = TRUE
+      )[["contrasts"]]
+      ),
+      out = purrr::map(res, broom::tidy)
+    ) |>
+    tidyr::unnest(c(out)) |>
+    dplyr::select(protein, oxygen, treatment, adj.p.value) |>
+    dplyr::mutate(
+      oxygen = replace(oxygen, oxygen == ".", "0.5%"),
+      oxygen = factor(oxygen, levels = c("21%", "0.5%")),
+      treatment = factor(treatment, levels = c("DMSO", "BAY")),
+      y_pos = Inf,
+      vjust = 1,
+      lab = dplyr::case_when(
+        adj.p.value < 0.05 ~ "*",
+        TRUE ~ NA_character_
+      )
+    )
+
+  list(data = df, annot = annot)
 }

### 2022-07-28 R/1_extracellular-fluxes.R

@@ -722,7 +722,7 @@ analyze_hyp_bay_fluxes <- function(growth, fluxes) {
         .x,
         "pairwise" ~ oxygen * treatment,
         simple = "each",
-        adjust = "Sidak",
+        adjust = "mvt",
         combine = TRUE
       )[["contrasts"]]
       ),

### 2022-07-28 R/1_figures.R

@@ -332,7 +332,7 @@ plot_dna_count_hypoxia <- function(dna_count_hypoxia) {
       size = 1.5,
       stroke = 0.2,
       show.legend = FALSE
-      ) +
+    ) +
     ggplot2::labs(
       x = "Cell count",
       y = "DNA (ng)"
@@ -1400,6 +1400,111 @@ plot_hyp_bay_fluxes <- function(df, annot, metab, ylab) {
     NULL
 }
 
+plot_hyp_bay_densities <- function(df, annot, prot, ylab) {
+  annot1 <-
+    dplyr::filter(annot, !is.na(treatment)) |>
+    dplyr::mutate(
+      treatment = factor(treatment, levels = c("DMSO", "BAY")),
+      oxygen = factor(oxygen, levels = c("21%", "0.5%"))
+    )
+  annot2 <-
+    dplyr::filter(annot, is.na(treatment)) |>
+    dplyr::mutate(
+      treatment = factor(treatment, levels = c("DMSO", "BAY")),
+      oxygen = factor(oxygen, levels = c("21%", "0.5%"))
+    )
+
+  df |>
+    dplyr::mutate(
+      treatment = factor(treatment, levels = c("DMSO", "BAY")),
+      oxygen = factor(oxygen, levels = c("21%", "0.5%"))
+    ) |>
+    ggplot2::ggplot() +
+    ggplot2::aes(
+      x = oxygen,
+      y = fold_change
+    ) +
+    ggplot2::stat_summary(
+      ggplot2::aes(fill = treatment),
+      geom = "col",
+      fun = "mean",
+      # width = 0.6,
+      position = ggplot2::position_dodge2(),
+      show.legend = TRUE,
+      alpha = 0.5
+    ) +
+    ggbeeswarm::geom_beeswarm(
+      ggplot2::aes(fill = treatment),
+      dodge.width = 0.9,
+      pch = 21,
+      size = 1,
+      stroke = 0.2,
+      cex = 4,
+      color = "white",
+      show.legend = FALSE
+    ) +
+    ggplot2::stat_summary(
+      ggplot2::aes(group = treatment),
+      geom = "errorbar",
+      fun.data = ggplot2::mean_se,
+      position = ggplot2::position_dodge(width = 0.9),
+      width = 0.2,
+      size = 0.25,
+      show.legend = FALSE
+    ) +
+    ggplot2::geom_text(
+      data = annot1,
+      ggplot2::aes(
[…]

### 2022-07-28 R/1_myc.R

@@ -1,7 +1,5 @@
 # 1_myc.R
 
-# 1_myc.R
-
 combine_fluxes <- function(growth_rates, fluxes, exp) {
   growth_rates |>
     dplyr::filter(experiment == exp) |>
@@ -76,17 +74,27 @@ plot_myc <- function(df, annot, metab, ylab, x, fill) {
       ),
       geom = "col",
       fun = "mean",
-      position = ggplot2::position_dodge(width = 0.6),
-      width = 0.6,
+      position = ggplot2::position_dodge2(),
+      alpha = 0.5,
       show.legend = TRUE
+    ) +
+    ggbeeswarm::geom_beeswarm(
+      ggplot2::aes(fill = {{fill}}),
+      dodge.width = 0.9,
+      pch = 21,
+      size = 1,
+      stroke = 0.2,
+      cex = 4,
+      color = "white",
+      show.legend = FALSE
     ) +
     ggplot2::stat_summary(
       ggplot2::aes(
         group = {{fill}}
       ),
       geom = "errorbar",
-      fun.data = "mean_se",
-      position = ggplot2::position_dodge(width = 0.6),
+      fun.data = ggplot2::mean_se,
+      position = ggplot2::position_dodge(width = 0.9),
       width = 0.2,
       size = 0.25,
       show.legend = FALSE,
@@ -100,9 +108,10 @@ plot_myc <- function(df, annot, metab, ylab, x, fill) {
         vjust = vjust,
         label = lab,
       ),
+      position = ggplot2::position_dodge(width = 0.9),
       family = "Calibri",
       color = "black",
-      size = 6/ggplot2::.pt,
+      size = 8/ggplot2::.pt,
       show.legend = FALSE
     ) +
     ggplot2::labs(
@@ -111,17 +120,23 @@ plot_myc <- function(df, annot, metab, ylab, x, fill) {
       color = NULL,
       fill = NULL
     ) +
-    ggplot2::scale_fill_manual(values = clrs, limits = force) +
+    ggplot2::scale_fill_manual(
+      values = clrs,
+      limits = force
+      ) +
     ggplot2::scale_y_continuous(
       expand = ggplot2::expansion(mult = c(0.05, 0.1)),
       breaks = scales::pretty_breaks(n = 6)
     ) +
-    # ggplot2::coord_cartesian(ylim = c(-750, 1250)) +
+    ggplot2::guides(
[…]

### 2022-07-28 R/1_rnaseq.R

@@ -333,10 +333,9 @@ get_msigdb_pathways <- function(species = 'Homo sapiens', category = NULL, subca
         gs_name
       )
     ) |>
-    tidyr::nest()
-
-  purrr::map(x$data, ~unlist(unname(as.list(.x)), recursive = FALSE)) |>
-    rlang::set_names(x$gs_name)
+    tidyr::nest() |>
+    dplyr::summarise(gs_symbol = unlist(data, recursive = FALSE)) |>
+    tibble::deframe()
 }
 
 run_gsea <- function(results, pathways){

### 2022-07-28 _targets.R

@@ -718,46 +718,46 @@ list(
 
   # cosmos ------------------------------------------------------------------
 
-  tar_target(
-    carnival_options,
-    set_carnival_options()
-  ),
-  tar_target(
-    cosmos_network,
-    get_cosmos_network()
-  ),
-  tar_target(
-    cosmos_tf,
-    format_tf(tfea_res_int)
-  ),
-  tar_target(
-    cosmos_metab,
-    format_metab(metab_tar_res_int)
-  ),
-  tar_target(
-    cosmos_deg,
-    format_deg(deg_int)
-  ),
-  tar_target(
-    cosmos_prep_forward,
-    preprocess("forward", cosmos_network, cosmos_tf, cosmos_metab, cosmos_deg, carnival_options)
-  ),
-  tar_target(
-    cosmos_prep_reverse,
-    preprocess("reverse", cosmos_network, cosmos_tf, cosmos_metab, cosmos_deg, carnival_options)
-  ),
-  tar_target(
-    cosmos_forward,
-    run_cosmos("forward", cosmos_prep_forward, carnival_options)
-  ),
-  tar_target(
-    cosmos_reverse,
-    run_cosmos("reverse", cosmos_prep_reverse, carnival_options)
-  ),
-  tar_target(
-    cosmos_res,
-    format_cosmos(cosmos_forward, cosmos_reverse)
-  ),
+  # tar_target(
+  #   carnival_options,
+  #   set_carnival_options()
+  # ),
+  # tar_target(
+  #   cosmos_network,
+  #   get_cosmos_network()
+  # ),
+  # tar_target(
+  #   cosmos_tf,
+  #   format_tf(tfea_res_int)
+  # ),
+  # tar_target(
+  #   cosmos_metab,
+  #   format_metab(metab_tar_res_int)
+  # ),
+  # tar_target(
+  #   cosmos_deg,
+  #   format_deg(deg_int)
+  # ),
+  # tar_target(
+  #   cosmos_prep_forward,
+  #   preprocess("forward", cosmos_network, cosmos_tf, cosmos_metab, cosmos_deg, carnival_options)
+  # ),
+  # tar_target(
+  #   cosmos_prep_reverse,
+  #   preprocess("reverse", cosmos_network, cosmos_tf, cosmos_metab, cosmos_deg, carnival_options)
+  # ),
+  # tar_target(
+  #   cosmos_forward,
+  #   run_cosmos("forward", cosmos_prep_forward, carnival_options)
+  # ),
+  # tar_target(
[…]

### 2022-07-29 manuscript/manuscript.Rmd
draft of manuscript assembled

@@
   path = "manuscript/manuscript.Rmd",
+  ignore = c("R-base"), 
   path_bib = "manuscript/library.json"
@@
 
-# targets::tar_load(
-#   c(
-#     m1_figure, 
-#   )
-# )
+# data for paper
+biomass <- 
+  targets::tar_read(biomass) |> 
+  tibble::deframe()
+resources <- targets::tar_read(resources_table)
+
+# figure dependencies
+targets::tar_load(
+  c(
+    m1_figure, 
+    m2_figure, 
+    m3_figure,
+    m4_figure, 
+    m5_figure, 
+    m6_figure, 
+    m7_figure
+  )
+)
 ```
+
+\newpage
 
@@
 
+Hypoxia requires metabolic adaptations to sustain cellular functions that underlie numerous physiologic and pathologic processes.
+While many studies have explored the metabolic consequences of hypoxia in cancer models, comparatively little is known about the metabolic response of primary cells to hypoxia.
+Here, we performed metabolic flux analyses of proliferating human lung fibroblasts and pulmonary artery smooth muscle cells in hypoxia.
+Unexpectedly, glycolytic flux was decreased in hypoxic cells despite activation of the hypoxia-inducible factor (HIF) transcriptional program and increased expression of glycolytic enzymes.
+Pharmacologic activation of HIF with the prolyl hydroxylase (PHD) inhibitor molidustat in normoxia did increase glycolytic flux, but hypoxia abrogated this effect.
+Multi-omic profiling of cells treated with hypoxia or molidustat, separately or together, revealed distinct molecular responses to hypoxia and pharmacologic PHD inhibition and suggested a critical role for MYC in modulating the HIF response in hypoxia.
+MYC knockdown in hypoxia increased lactate efflux while MYC overexpression in normoxia blunted the effects of molidustat treatment.
+Together, these data suggest that other factors, notably MYC, supersede the anticipated effects of HIF-dependent up-regulation of glycolytic gene expression on glycolytic flux in hypoxic proliferating primary cells.
+
 # Keywords
 
-/ /
+hypoxia / metabolic flux analysis / hypoxia-inducible factor / prolyl hydroxylase / MYC
 
@@
 
+In this work, we used ^13^C metabolic flux analysis to identify hypoxia-mediated metabolic changes in proliferating human primary cells.
+Our principal finding was that hypoxia reduced, rather than increased, carbon flux through glycolysis and lactate fermentation pathways despite robust activation of the HIF transcriptional program and up-regulation of glycolytic genes.
+Certainly, the LFs studied here are capable of augmenting glycolysis in response to HIF stabilization, as demonstrated by experiments with the PHD inhibitor BAY; however, these effects are completely attenuated when BAY-treated cells are cultured in hypoxia.
+Together, these findings suggest that changes in enzyme levels alone are insufficient to alter metabolic flux in hypoxia and point to the importance of regulatory mechanisms that supersede the effects of HIF-dependent gene transcription in primary cells.
+
+<!-- MYC role in metabolism -->
+
+Our data indicate that hypoxia-induced MYC expression is one such regulatory mechanism.
+MYC is a transcription factor that regulates the expression of numerous genes involved in many biological processes, including metabolism, proliferation, apoptosis, and differentiation [@dang2006; @li2020d; @stine2015].
+As deregulated MYC activity has been associated with the majority of human cancers [@vita2006], much of our understanding of MYC regulation comes from studies using cancer cell models [@li2020d; @stine2015; @dang2012a; @madden2021], while the role of MYC in the biology of untransformed cells is less well understood.
+The literature describes a complex and reciprocal relationship between HIF and MYC that depends on both environmental (*e.g.*, hypoxia) and cellular context [@li2020d].
+Generally, HIF-1 has been observed to inhibit MYC through multiple mechanisms [@koshiji2004; @koshiji2005; @gordan2007; @zhang2007], and this previous work is consistent with our present observations that HIF stabilization following BAY treatment decreased MYC protein and target gene expression.
+Conversely, MYC has been implicated in increased HIF activity through transcriptional and post-transcriptional mechanisms [@zhang2009a; @chen2013; @doe2012], primarily in the context of malignant transformation.
+The observation that MYC may antagonize the transcriptional effects of HIF to sustain primarily cell proliferation and metabolism in hypoxia suggests a substantially different regulatory relationship than has been previously described.
+
+Understanding how MYC transcriptional activity affects hypoxic primary cell metabolism is imperative to our understanding of cellular adaptation to hypoxia.
+MYC stimulates the expression of nuclear-encoded mitochondrial genes and promotes mitochondrial biogenesis, both directly and through activation of mitochondrial transcriptional factor A (TFAM) [@li2005a].
+Indeed, we found that the oxidative phosphorylation gene set was relatively enriched with hypoxia culture of BAY-treated cells (**Fig \@ref(fig:m6)C**).
+In this way, hypoxic MYC activation may sustain energy production by oxidative phosphorylation, thereby decreasing the energetic demands that would otherwise drive increased glycolytic flux.
+Beyond oxidative phosphorylation, MYC targets genes involved in many other intermediary metabolic pathways, including amino acids, nucleotides, and lipids [@stine2015], that may also impact the central pathways of carbon metabolism studied in this work.
+
+<!-- Important non-HIF mechanisms -->
+
+Beyond MYC, the identification of other HIF-independent mechanisms regulating primary cell adaption to hypoxia is of critical importance.
+Cells express several oxygen-dependent enzymes in addition to PHD whose activities may be altered in hypoxia but not by PHD inhibition.
+For example, PHD is one of many α-ketoglutarate-dependent dioxygenase enzymes that rely on molecular oxygen for their catalytic activity [@islam2018].
+Jumonji-C (JmjC) domain-containing histone demethylases are other prominent members of this family whose inhibition by hypoxia has been shown to cause rapid and HIF-independent induction of histone methylation [@batie2019].
+Similarly, a recently described cysteamine dioxygenase has been shown to mediate the oxygen-dependent degradation of Regulators of G protein Signaling 4 and 5 and IL-32 [@masson2019].
+In addition to dioxygenase enzymes, electron transport chain dysfunction resulting from impaired Complex IV activity leads to increased ROS production in hypoxia [@chandel1998].
+Mitochondrial ROS increase the half-lives of several mRNAs in hypoxia, including MYC, as we observed in this work, independent of HIF stabilization [@guzy2005].
+Finally, hypoxia imposes a reductive stress on cells associated with an increase in the NADH/NAD^+^ ratio secondary to impaired electron transport (**Fig \@ref(fig:m5)K**)[@chance1955; @garofalo1988].
+NADH accumulation may slow glycolysis *via* feedback inhibition of GAPDH [@tilton1991].
+Any or all of these molecular mechanisms may also contribute to uncoupling glycolytic enzyme expression from glycolytic flux as observed in the experiments described here.
+
+<!-- HIF and cell cycle progression -->
+
+In addition to its effects on cellular metabolism, another canonical role of HIF-1 activation is slowing of cellular proliferation rate in the face of limited oxygen availability [@hubbi2015], which is precisely what we observed here (**Figs \@ref(fig:m1)B, \@ref(fig:m2)B, \@ref(fig:m5)A**).
+The effects of HIF-1 on cell proliferation rate are mediated, in part, by increased expression of cyclin-dependent kinase inhibitor p21 (*CDKN1A*), inhibition of E2F targets [@gardner2001], and inhibition of pro-proliferative MYC signaling [@koshiji2004].
+These transcriptional effects are precisely what is observed in BAY treated LFs in normoxia.
+By contrast, hypoxia culture was associated with decreased expression of p21, consistent with a previous report [@mizuno2009], as well as increased expression of MYC protein and enrichment of MYC target genes.
+Indeed, the most marked differences between hypoxia and BAY treatment on LF gene transcription were the up-regulation of pro-proliferative gene sets containing E2F targets and G2/M checkpoint proteins.
+Much of this transcriptional response may be mediated by hypoxia-induced up-regulation of MYC, which is known to stimulate cell cycle progression through its effects on the expression and activity of cyclins, cyclin-dependent kinases, and cyclin-dependent kinase inhibitors [@hydbring2017].
+Clarifying the complex interactions among HIFs, MYC, and cell proliferation will be important for understanding the cellular response of these mesenchymal cells to tissue injury.
+
+<!-- Why HIF? -->
+
+Taken together, these findings raise important questions regarding the cell-autonomous role of HIFs in the hypoxia response.
+On an organismal level, HIFs drive expression of angiogenic and erythropoietic factors to increase oxygen delivery to hypoxic tissues.
+Within individual cells, HIF-1α seems to be important for mitigating the adverse effects of ROS formation by dysfunctional electron transport in the mitochondria.
+Indeed, hypoxia increased oxygen consumption and ROS production in HIF-1α-null mouse embryonic fibroblasts (MEFs), which was associated with increased cell death [@zhang2008a].
+Interestingly, these cells also had increased ATP levels compared to wild type, suggesting that mitochondrial function was adequate under 1% oxygen culture conditions to support oxidative phosphorylation and meet the energy needs of the cells.
+Given the prominence of HIFs in mediating the transcriptional response to hypoxia, it is somewhat surprising that neither PHD, HIFs, nor their downstream targets were found to be selectively essential as a function of oxygen tension in a genome-wide CRISPR growth screen of K562 human lymphoblasts cultured in normoxia or hypoxia [@jain2020].
+Similarly, knockout of HIF signaling did not affect growth, internal metabolite concentrations, glucose consumption, or lactate production under hypoxia by human acute myeloid leukemia cells [@wierenga2019].
+Together with our results, these studies highlight the need for additional research linking hypoxia-induced metabolic changes to their transcriptional and post-transcriptional regulatory mechanisms, particularly in primary cells.
+
+<!-- Lactate consumption -->
+<!-- PASMC reductive carboxylation -->
+
+This work also highlights two specific metabolic features that appear to be important in the metabolic response of primary cells to hypoxia.
+First, both LFs and PASMCs demonstrated notable incorporation of lactate-derived carbon into intracellular metabolic pathways that increased with hypoxia and BAY treatments (**Figs \@ref(fig:m4), S4, S5**).
+This finding is consistent with increasing evidence suggesting an important role for lactate as a metabolic fuel in several organ systems [@faubert2013; @hui2017].
+Although typically considered a metabolic waste product [@rabinowitz2020], an important contribution of lactate *import* in supporting metabolic homeostasis in the face of an ischemic insult, which is associated with increased extracellular lactate, is an evolutionarily attractive hypothesis that merits further investigation.
+Second, PASMCs, but not LFs, demonstrated significant rates of reductive carboxylation that increased in 0.5% oxygen (**Fig S5**).
+Reductive carboxylation was first identified in hypoxic tumor cells where stable isotope tracing revealed ^13^C incorporation from labeled glutamine into lipids [@metallo2011; @gameiro2013; @scott2011; @wise2011].
+Hypoxia drives PASMC proliferation *in vivo* contributing to the development pulmonary hypertension in humans and animal models.
+Isocitrate dehydrogenase has previously been implicated in the pathobiology of this disease [@fessel2012], and our findings suggest that reductive carboxylation catalyzed by isocitrate dehydrogenase may be a metabolic vulnerability of hypoxic PASMCs associated with pulmonary vascular disease.
+
+<!-- Limitations -->
+
+Our finding that hypoxia was associated with decreased glycolysis and lactate fermentation was unexpected.
+Several aspects of our experimental design may have contributed to this finding.
+First, our goal was to understand how metabolic reprogramming may support cell proliferation in hypoxia.
+Thus, we measured metabolite fluxes in cells during the exponential growth phase accounting for cell growth rate, metabolite degradation rates, and medium evaporation with multiple measurements over a 72 h time course.
+Often, cells are studied near confluence, where metabolic contributions to biomass production are less and the rate of glycolysis in hypoxia may be higher.
+Second, we began our experimental treatments 24 h prior to collecting samples to ensure that the hypoxia metabolic program was established prior to labeling.
+Similar studies [@metallo2011; @grassian2014] typically placed cells into hypoxia at the time of labeling.
+Third, and perhaps most importantly, these flux determinations were performed in human primary cell cultures rather than immortalized cell lines.
+Although both cell types used in this study were derived from lung, we anticipate that many of our findings will be generalizable to primary cells from different tissues.
+
+<!-- Conclusion -->
+
+In summary, in this metabolic flux analysis of proliferating human primary cells *in vitro*, we have demonstrated that MYC uncouples an increase in HIF-dependent glycolytic gene transcription from glycolytic flux in hypoxia.
+Indeed, the degree of metabolic reprogramming in hypoxia was modest and suggests close coupling between proliferation and metabolism.
+In light of our findings, additional studies are warranted to clarify the role of HIFs in mediating the metabolic response to hypoxia, to determine how MYC activity is regulated by hypoxia, and to identify other key regulators of hypoxic metabolic reprogramming in primary cells.
+Moreover, these data strongly caution investigators against drawing conclusions about metabolite flux from measures of gene transcription alone.
+
 \newpage
@@
 
-## Data Availability
+## Cell culture
+
+Primary normal human lung fibroblasts (LFs) were purchased from Lonza (CC-2512) and cultured in FGM-2 (Lonza CC-3132) in a standard tissue culture incubator in 5% CO~2~ at 37 °C.
+Cells from two donors were used in these studies: #33652 (56 y.o., male) and #29132 (19 y.o., female).
+Cell authentication was performed by the vendor.
+Primary human pulmonary artery smooth muscle cells were purchased from Lonza (CC-2581) and cultured in SmGM-2 (Lonza CC-3182) in a standard tissue culture incubator in 5% CO~2~ at 37 °C.
+Cells from multiple donors were used in these studies: #30020 (64 y.o., male), #27662 (35 y.o., male), #26698 (51 y.o., male), and #19828 (51 y.o., male).
+Cell authentication was performed by the vendor.
+
+## Metabolic flux protocol
+
+For extracellular flux measurements, cells were seeded in either standard growth medium or MCDB131 medium without glucose, glutamine, or phenol red (genDEPOT) supplemented with 2% dialyzed fetal bovine serum (Mediatech) and naturally labeled glucose and glutamine ("light" labeling medium).
+For LFs, glucose was supplemented at 8 mM and glutamine was supplemented at 1 mM.
+For PASMCs, glucose was supplemented at 5.55 mM and glutamine was supplemented at 10 mM.
+These concentrations match the concentrations of these substrates determined in standard growth medium.
+Preliminary experiments were performed to identify the optimal cell seeding density, exponential growth phase, and labeling duration consistent with metabolic and isotopic steady state.
+On Day -1, 25,000 cells were seeded in a 35 mm dish in "light" labeling medium.
+Hypoxia-treated cells were transferred to a tissue culture glovebox set to 0.5% oxygen and 5% CO~2~ (Coy Lab Products).
+Medium was supplemented with DMSO 0.1% or BAY (10 μM) for DMSO and BAY treatment conditions.
+On Day 0, cells were washed with PBS and the medium was changed to either "light" medium for flux measurements or "heavy" medium for tracer experiments.
+For LFs, samples were collected on Day 0 and every 24 h for 72 h.
+For PASMCs, samples were collected on Day 0 and every 12 h for 48 h.
+Medium and cell lysates were collected at each time point for intra- and extracellular metabolite measurements and total DNA quantification.
+Dishes without cells were weighed daily to correct for evaporative medium losses and to empirically determine degradation and accumulation rates of metabolites.
+Medium samples and cell lysates for DNA measurement were stored at -80 °C until analysis.
+Each individual experiment included triplicate wells for each treatment and time point, and each experiment was repeated 4-8 times.
+## Cell count
+
+Direct cell counts of trypsinized cell suspensions in PBS were obtained following staining with propidium iodide and acridine orange using a LUNA-FL fluorescence cell counter (Logos Biosystems).
+Indirect cell counts for flux measurements were interpolated from total DNA quantified using the Quant-iT PicoGreen dsDNA Assay Kit (Thermo).
+Cells were washed once with two volumes of PBS, lysed with Tris-EDTA buffer containing 2% Triton X-100, and collected by scraping.
+Total DNA in 10 μL of lysate was determined by adding 100 μL of 1X PicoGreen dye in Tris-EDTA buffer and interpolating the fluorescence intensity with a standard curve generated using the λ DNA standard.
+Cell counts were interpolated from a standard curve of DNA obtained from known cell numbers seeded in basal medium (**Fig S1B**).
+No difference in total cellular DNA was identified between normoxia and hypoxia cultures (**Fig S1D**).
+
+## Immunoblots
+
+Cells were washed with one volume of PBS and collected by scraping in PBS.
+Cell suspensions were centrifuged at 5,000 ×*g* for 5 min at 4 °C.
+Pellets were lysed in buffer containing Tris 10 mM, pH 7.4, NaCl 150 mM, EDTA 1 mM, EGTA 1 mM, Triton X-100 1% v/v, NP-40 0.5% v/v, and Halt Protease Inhibitor Cocktail (Thermo).
+Protein concentrations were determined by BCA Protein Assay (Thermo).
+Lysates were normalized for protein concentration and subjected to SDS-PAGE separation on stain-free tris-glycine gels (Bio-Rad), cross-linked and imaged with the Chemidoc system (Bio-Rad), transferred to PVDF membranes with the Trans-Blot Turbo transfer system (Bio-Rad), imaged, blocked in 5% blocking buffer (Bio-Rad), blotted in primary and secondary antibodies, and developed using WesternBright ECL (Advansta).
+Band signal intensity was normalized to total protein per lane as determined from the stain-free gel or membrane images.
+
+## RT-qPCR
+
+Total RNA was isolated from cells with the RNeasy Mini Kit (Qiagen).
+cDNA was synthesized from 0.25-1.00 ng RNA with the High Capacity cDNA Reverse Transcription Kit (Applied Biosystems).
+RT-qPCR analysis was performed with an Applied Biosystems 7500 Fast Real Time PCR System with TaqMan Universal PCR Master Mix and pre-designed TaqMan gene expression assays (Life Technologies).
+Relative expression levels were calculated using the comparative cycle threshold method referenced to *ACTB*.
+
+##  Glucose assay
+
+Medium samples were diluted 10-fold in PBS.
+Glucose concentration was determined using the Glucose Colorimetric Assay Kit (Cayman) according to the manufacturer's protocol.
+Standards were prepared in PBS.
+
+## Lactate assay
+
+Medium samples were diluted 10-fold in PBS.
+Glucose concentration was determined using the ʟ-Lactate Assay Kit (Cayman).
+Medium samples did not require deproteinization, otherwise the samples were analyzed according to the manufacturer's protocol.
+Standards were prepared in PBS.
+## Pyruvate assay
+
+Pyruvate was measured using either an enzymatic assay (most samples) or an HPLC-based assay (medium from 0.2% oxygen experiments).
+For the enzymatic assay, medium samples were diluted 20-fold in PBS.
+Pyruvate concentration was determined using the Pyruvate Assay Kit (Cayman).
+Medium samples did not require deproteinization, otherwise the samples were analyzed according to the manufacturer's protocol.
+Standards were prepared in PBS.
+For the HPLC assay, 2-oxovaleric acid was added to medium samples as an internal standard.
+Samples were subsequently deproteinized with 2 volumes of ice-cold acetone.
+Supernatants were evaporated to < 50% of the starting volume at 43 °C in a SpeedVac concentrator (Thermo Savant) and reconstituted to the starting volume with HPLC-grade water prior to derivatization.
+Samples were derivatized 1:1 by volume with *o*-phenylenediamine (25 mM in 2 M HCl) for 30 min at 80 °C.
+Derivatized pyruvate was separated with a Poroshell HPH C-18 column (2.1 × 100 mm, 2.7 μm) on an Infinity II high-performance liquid chromatography system with fluorescence detection of OPD-derivatized α-keto acids as described previously [@guarino2019].
+
+## Amino acid assay
+
+Medium amino acid concentrations were determined following the addition of norvaline and sarcosine internal standards and deproteinization with 2 volumes of ice-cold acetone.
+Supernatants were evaporated to < 50% of the starting volume at 43 °C in a SpeedVac concentrator (Thermo Savant) and reconstituted to the starting volume with HPLC-grade water prior to analysis.
+Amino acids in deproteinized medium were derivatized with *o*-phthalaldehyde (OPA) and 9-fluorenylmethylchloroformate (FMOC) immediately prior to separation with a Poroshell HPH-C18 column (4.6 × 100 mm, 2.7 μm) on an Infinity II high-performance liquid chromatography system with ultraviolet and fluorescence detection of OPA- and FMOC-derivatized amino acids, respectively, according to the manufacturer's protocol (Agilent) [@long2017].
+
+## MYC knockdown
+
+Approximately 1.25 M LFs were reverse transfected in 6-cm dishes with 40 pmol siMYC or non-targeting siCTL pools (Dharmacon) in 20 uL RNAiMAX (Thermo).
+After 24 h, cells were collected by trypsinization and re-seeded as described in *Metabolic flux protocol* above for growth rate and lactate efflux measurements.
+
+## MYC overexpression
+
+LFs were seeded at 25,000 cells per 35 mm dish on Day -2.
+On Day -1, cells were transduced with adenovirus for MYC (Vector Biolabs) or YFP overexpression [@oldham2015].
+After 24 h, the medium was changed and samples were collected as described in *Metabolic flux protocol* above.
+
+## Flux calculations
+
+The growth rate ($\mu$) and flux (*v*) for each measured metabolite were defined as follows [@murphy2013a]:
+
+\begin{align}
+\frac{dX}{dt} &= \mu X (\#eq:1) \\[6 pt] 
+\frac{dM}{dt} &= -kM+vX (\#eq:2) 
+\end{align}
+
+where *X* is the cell density, *k* is the first-order degradation or accumulation rate, and *M* is the mass of the metabolite. These equations are solved as follows:
+
+\begin{align}
+X &= X_0e^{\mu t} (\#eq:3)\\[6 pt]
+Me^{k t} &= \frac{v X_0}{\mu + k} (e^{(\mu + k)t}-1) + M_0 (\#eq:4) 
+\end{align}
+
+Growth rate ($\mu$) and cell count at time 0 ($X_0$) were determined by robust linear modeling of the logarithm of cell count as a function of time ($t$).
+Metabolite mass was calculated from the measured metabolite concentrations and predicted well volume accounting for evaporative losses (**Fig S1E**).
+First-order degradation and accumulation rates were obtained from robust linear modeling of metabolite mass *v*.
+time in unconditioned culture medium.
+Rates that significantly differed from 0 using Student's *t*-test were incorporated into the flux calculations.
+Final fluxes were obtained by robust linear modeling of $Me^{k t}$ *versus* $(e^{(\mu + k)t}-1)$ to determine the slope from which $v$ was calculated using equation \@ref(eq:4).
+
+## Metabolomics
+
+### Metabolite extraction
+
+Intracellular metabolites were obtained after washing cells with 2 volumes of ice-cold PBS and floating on liquid nitrogen.
+Plates were stored at -80 °C until extraction.
+Metabolites were extracted with 1 mL 80% MeOH pre-cooled to -80 °C containing 10 nmol [D~8~]-valine as an internal standard (Cambridge Isotope Labs).
+Insoluble material was removed by centrifugation at 21,000 ×*g* for 15 min at 4 °C.
+The supernatant was evaporated to dryness at 42 °C using a SpeedVac concentrator (Thermo Savant).
+Samples were resuspended in 35 μL LC-MS-grade water prior to analysis.
+
+### Acquisition parameters
+
+LC-MS analysis was performed on a Vanquish ultra-high-performance liquid chromatography system coupled to a Q Exactive orbitrap mass spectrometer by a HESI-II electrospray ionization probe (Thermo).
+External mass calibration was performed weekly.
+Metabolite samples (2.5 μL) were separated using a ZIC-pHILIC stationary phase (2.1 × 150 mm, 5 μm) (Merck).
+The autosampler temperature was 4 °C and the column compartment was maintained at 25 °C.
+Mobile phase A was 20 mM ammonium carbonate and 0.1% ammonium hydroxide.
+Mobile phase B was acetonitrile.
+The flow rate was 0.1 mL/min.
+Solvent was introduced to the mass spectrometer *via* electrospray ionization with the following source parameters: sheath gas 40, auxiliary gas 15, sweep gas 1, spray voltage +3.0 kV for positive mode and -3.1 kV for negative mode, capillary temperature 275 °C, S-lens RF level 40, and probe temperature 350 °C.
+Data were acquired and peaks integrated using TraceFinder 4.1 (Thermo).
+
+### Stable isotope quantification
+
+All metabolites except fructose 2,6-bisphosphate (FBP) and 3-phosphoglycerate (3PG) were measured using the following mobile phase gradient: 0 min, 80% B; 5 min, 80% B; 30 min, 20% B; 31 min, 80% B; 42 min, 80% B.
+The mass spectrometer was operated in selected ion monitoring mode with an m/*z* window width of 9.0 centered 1.003355-times half the number of carbon atoms in the target metabolite.
+The resolution was set at 70,000 and AGC target was 1 × 10^5^ ions.
+Peak areas were corrected for quadrupole bias as previously described [@kim2015].
+Mass isotope distributions for FBP and 3PG were calculated from full scan chromatograms as described below.
+Raw mass isotopomer distributions were corrected for natural isotope abundance using a custom R package (mzrtools, https://github.com/oldhamlab/mzrtools) employing the method of Fernandez, *et al.* -[@fernandez1996].
+
+### Metabolomic profiling
+
+For metabolomic profiling and quantification of isotopic enrichment for FBP and 3PG, the following mobile phase gradient was used: 0 min, 80% B; 20 min, 20% B; 20.5 min, 80% B; 28 min, 80% B; 42 min, 80% B.
+The mass spectrometer was operated in polarity switching full scan mode from 70-1000 m/*z*.
+Resolution was set to 70,000 and the AGC target was 1x10^6^ ions.
+Peak identifications were based on an in-house library of authentic metabolite standards previously analyzed utilizing this method.
+For metabolomics studies, pooled quality control (QC) samples were injected at the beginning, end, and between every four samples of the run.
+Raw peak areas for each metabolite were corrected for instrument drift using a cubic spline model of QC peak areas.
+Low quality features were removed on the basis of a relative standard deviation greater than 0.2 in the QC samples and a dispersion ratio greater than 0.4 [@broadhurst2018].
+Missing values were imputed using random forest.
+Samples peak areas were normalized using probabilistic quotient normalization [@dieterle2006].
+Differentially regulated metabolites were identified using limma [@ritchie2015a].
+Metabolite set enrichment analysis was performed using the fgsea package [@korotkevich2021a] with KEGG metabolite pathways [@kanehisa2000].
+
+## Biomass determination
+
+The dry weight of LFs was determined to be `r round(biomass[["lf"]])` pg/cell.
+The dry weight of PASMCs was determined to be `r round(biomass[["pasmc"]])` pg/cell.
+These values were estimated by washing 3 × 10^6^ cells twice in PBS and thrice in ice-cold acetone prior to drying overnight in a SpeedVac.
+The composition of the dry cell mass was estimated from the literature [@quek2010; @sheikh2005], and stoichiometric coefficients were determined as described [@murphy2013; @zamorano2010].
+
+## Metabolic flux analysis
+
+Metabolic flux analysis was performed using the elementary metabolite unit-based software package INCA [@young2014].
+Inputs to the model include the chemical reactions and atom transitions of central carbon metabolism, extracellular fluxes, the identity and composition of ^13^C-labeled tracers, and the MIDs of labeled intracellular metabolites.
+The metabolic network was adapted from previously published networks [@murphy2013; @vacanti2014] and comprises `r nrow(dplyr::filter(model_reactions, !(pathway %in% c("mixing", "dilution"))))` reactions representing glycolysis, the pentose phosphate pathway, the tricarboxylic acid cycle, anaplerotic pathways, serine metabolism, and biomass synthesis.
+The network includes seven extracellular substrates (aspartate, cystine, glucose, glutamine, glycine, pyruvate, serine) and five metabolic products (alanine, biomass, glutamate, lactate, lipid).
+Models were fit using three ^13^C-labeled tracers, [1,2-^13^C~2~] glucose, [U-^13^C~6~] glucose, and [U-^13^C~5~] glutamine.
+The MIDs of twelve metabolites (2-oxoglutarate, 3-phosphoglycerate, alanine, aspartate, citrate, fructose bisphosphate, glutamate, glutamine, lactate, malate, pyruvate, serine) were used to constrain intracellular fluxes.
+The following assumptions were made:
+
+1. Metabolism was at steady state.
+1. Labeled CO~2~ produced during decarboxylation reactions left the system and did not re-incorporate during carboxylation reactions.
+1. Protein turnover occurred at a negligible rate compared to glucose and glutamine consumption.
+1. Acetyl-CoA, aspartate, fumarate, malate, oxaloacetate, and pyruvate existed in cytosolic and mitochondrial pools. Aspartate and malate were allowed to exchange freely between the compartments.
+1. The per cell biomass requirements of proliferating lung fibroblasts were similar to published estimated in other cells [@quek2010; @sheikh2005]. 
+1.
+Dilution of alanine, aspartate, glutamate, glutamine, lactate, and pyruvate was allowed to occur through reversible exchange with unlabeled substrates in the medium as these metabolites were present in unconditioned medium or serum.
+1. Succinate and fumarate are symmetric molecules that have interchangeable orientations when metabolized by TCA cycle enzymes.
+
+Flux estimation was repeated a minimum of 50 times from random initial values.
+Results were subjected to a χ^2^ statistical test to assess goodness-of-fit.
+Accurate 95% confidence intervals were computed for estimated parameters by evaluating the sensitivity of the sum-of-square residuals to parameter variations [@antoniewicz2006; @murphy2013].
+
+## NAD(H) assay
+
+Cellular NAD^+^ and NADH were measured using an enzymatic fluorometric cycling assay based on the reduction of NAD^+^ to NADH by alcohol dehydrogenase (ADH) and subsequent electron transfer to generate the fluorescent molecule resorufin [@oldham2015].
+Briefly, cells were washed twice with one volume PBS.
+Pyridine nucleotides were extracted on ice with buffer containing 50% by volume PBS and 50% lysis buffer (100 mM sodium carbonate, 20 mM sodium bicarbonate, 10 mM nicotinamide, 0.05% by volume Triton-X-100, 1% by mass dodecyltrimethylammonium bromide) and collected by scraping.
+Extracts were divided equally and 0.5 volume of 0.4 N HCl was added to one sample.
+Both extracts were heated at 65 °C for 15 min to degrade selectively either the oxidized (buffer) or reduced (HCl) nucleotides.
+The reaction was cooled on ice and quenched by adding 0.5 M Tris-OH to the acid-treated samples or 0.2 N HCl plus 0.25 M Tris-OH to the buffer samples.
+Samples were then diluted in reaction buffer (50 mM EDTA and 10 mM Tris, pH 7.06).
+Cell debris was pelleted by centrifugation, and 50 μL was incubated for 2 h with 100 μL reaction buffer containing 0.6 M EtOH, 0.5 mM phenazine methosulfate, 0.05 mM resazurin, and 0.1 mg/mL ADH.
+Fluorescence intensities were measured with a Spectramax Gemini XPS (Molecular Devices) with excitation 540 nm, emission 588 nm, and 550 nm excitation cut-off filter.
+Sample intensities were compared to a standard curve generated from known concentrations of NADH.
+The ratio of fluorescence in buffer-extracted to acid-extracted samples corresponds to the NADH/NAD^+^ ratio.
+Absolute NADH and NAD^+^ were normalized to cell count from cells estimated from total DNA quantification as described above.
+
+## RNA-seq
+
+RNA was collected from LFs treated for three days ± hypoxia ± BAY as described above.
+Four biological replicates were analyzed.
+Library construction and sequencing was performed by BGI Genomics using 100 bp paired end analysis and a read depth of 50M reads per sample.
+Sequences were deposited in the NIH SRA (PRJNA721596).
+Sequences were mapped to the human GRCh38 primary assembly and counts summarized using Rsubread [@liao2019].
+This data is available from the Oldham Lab GitHub repository (https://github.com/oldhamlab/rnaseq.lf.hypoxia.molidustat).
+Differentially expressed transcripts were identified using DESeq2 [@love2014].
+Gene set enrichment and transcription factor enrichment was performed using the fgsea and DoRothEA R packages, respectively [@korotkevich2021a; @garcia-alonso2019].
+
+## Quantification and Statistical Analysis
+
+The raw data and annotated analysis code necessary to reproduce this manuscript are contained in an R package research compendium available from the Oldham Lab GitHub repository (https://github.com/oldhamlab/Copeland.2021.hypoxia.flux).
+Data analysis, statistical comparisons, and visualization were performed in R [@R-base] using the packages referenced in the data supplement.
+Experiments included technical and biological replicates as noted above.
+The number of biological replicates (n) is indicated in the figure legends.
+Summary data show the mean ± SEM.
+Outliers were identified using twice the median absolute deviation as a cutoff threshold.
+Two group comparisons (*e.g.*, 21% *v.* 0.5% oxygen) were performed using Student's *t*-test on log-transformed values with pairing among biological replicates.
+Multifactor comparisons (*e.g.*, BAY and oxygen) were performed using linear mixed-effects models with *post hoc* comparisons based on the multivariate *t* distribution.
+Metabolomics and RNA-seq data were analyzed as described above.
+Probability values less than 0.05 were considered significant.
+
+\newpage
+
+## Key Resources Table
+
+```{r, results = "asis"}
+flextable::flextable_to_rmd(resources)
+```
 
@@
 
-# Author Contributions
+This work was supported by grants from the NIH (K08HL128802), American Lung Association, Pulmonary Hypertension Association, and the American Thoracic Society Foundation to W.M.O and from the NIH (U01HG007690, U54HL119145, R01HL155107, R01HL155096) and the American Heart Association (D700382, CV-19) to J.L.
 
-# Disclosures and competing interests
+# Author contributions
+
+W.M.O.
+conceived and designed the analysis.
+C.A.C., B.A.O., D.R.Z., S.M., K.L., and W.M.O.
+collected the data.
+J.D.Y.
+and W.M.O.
+contributed data or analysis tools.
+W.M.O.
+performed the analysis.
+W.M.O.
+drafted the manuscript.
+All authors participated in interpreting the results and revising the manuscript.
+All authors approve the final submission.
+
+# Conflict of interest
+
+The authors declare that they have no competing interests.
 
@@
 
-# Tables
-
-\newpage
-
 # Figure Legends
 
-(ref:m1) **Caption.** Info.
+(ref:m1) **Effects of hypoxia on extracellular metabolite fluxes in lung fibroblasts.** (**A**) Lung fibroblasts (LFs) were cultured in 21% (*red*) or 0.5% (*blue*) oxygen beginning 24 h prior to time 0.
+Samples were collected every 24 h for 72 h.
+(**B**) Growth curves of LFs in each experimental condition (n = 8).
+(**C**) Growth rates from (B) were determined by robust linear modeling of log-transformed growth curves.
+(**D**) Representative immunoblot of LF protein lysates cultured as in (A).
+(**E**) Relative change in HIF-1α protein levels from (D) normalized to 21% oxygen at time 0 (n = 4).
+(**F**) Relative change in GLUT1 mRNA levels normalized to 21% oxygen treatment at time 0 (n = 4).
+(**G**) Relative change in LDHA mRNA levels as in (F).
+(**H**) Relative change in LDHA protein levels as in (E).
+(**I**) Extracellular fluxes of glucose (GLC) and lactate (LAC) (n = 8).
+(**J**) Extracellular fluxes of pyruvate (PYR) and amino acids.
+Data are mean ± SEM.
+Comparisons were made using linear mixed effects models with oxygen as the fixed effect and biological replicate as a random effect with p-values adjusted using the multivariate *t* distribution (* p < 0.05).
+
+(ref:m2) **Effects of pharmacologic prolyl hydroxylase inhibition on extracellular metabolite fluxes in lung fibroblasts.** (**A**) Lung fibroblasts (LFs) were treated with PHD inhibition molidustat (BAY, 10 μM) (*purple*) or DMSO (*green*) beginning 24 h prior to time 0.
+Samples were collected every 24 h for 72 h.
+(**B**) Growth curves of LFs in each experimental condition (n = 8).
+(**C**) Growth rates from (B).
+(**D**) Representative immunoblot of LF protein lysates cultured as in (A).
+(**E**) Relative change in HIF-1α protein levels from (D) normalized to DMSO at time 0 (n = 4).
+(**F**) Relative change in GLUT1 mRNA levels normalized to DMSO at time 0 (n = 4).
+(**G**) Relative change in LDHA mRNA levels as in (F).
+(**H**) Relative change in LDHA protein levels as in (E).
+(**I**) Extracellular fluxes of glucose (GLC) and lactate (LAC) (n = 8).
+(**J**) Extracellular fluxes of pyruvate (PYR) and amino acids.
+Data are mean ± SEM.
+Comparisons were made using linear mixed effects models with treatment as the fixed effect and biological replicate as a random effect with p-values adjusted using the multivariate *t* distribution (* p < 0.05).
+
+(ref:m3) **Metabolic flux analysis of lung fibroblasts following hypoxic and pharmacologic PHD inhibition.** (**A**) Mass isotopomer distribution (MID) of pyruvate (PYR) following 72 h labeling of LFs with [U-^13^C~6~]-glucose (GLC).
+(**B**) MID of citrate after 72 h labeling of LFs with [U-^13^C~6~]-GLC (**C**) MID of citrate after 72 h labeling of LFs with [U-^13^C~5~]-glutamine (GLN).
+Data are mean ± SEM.
+Comparisons were made using linear mixed effects models with treatment as the fixed effect and biological replicate as a random effect with p-values adjusted using the multivariate *t* distribution (n = 4, p < 0.05 indicated as * 0.5% *v.* 21% oxygen, † BAY *v.* DMSO, ‡ Δoxygen *v.* ΔBAY).
+(**D**) Fraction of M5 citrate indicating reductive carboxylation after labeling with [U-^13^C~5~]-GLN in LFs and PASMCs (n = 3-4, * p < 0.05 0.5% *v.* 21% oxygen).
+(**E**) Ratio of metabolic fluxes in 0.5% oxygen compared to 21% oxygen.
+Fluxes with non-overlapping confidence intervals are highlighted with arrows colored according to the magnitude of the change.
+Arrow thickness corresponds to the absolute flux measured in hypoxia.
+(**F**) Ratio of metabolic fluxes in BAY-treated cells compared to DMSO-treated control.
+Arrows are colored as in (E) and arrow weights correspond to the absolute flux as measured in BAY-treated cells.
+
+(ref:m4) **PHD inhibition increases lactate uptake and oxidation.** (**A**) Exchange flux estimates for reductive carboxylation of α-ketoglutarate (AKG) to citrate (CIT).
+Data show the estimate with upper and lower bounds for LFs and PASMCs.
+(**B**) Exchange flux estimates for lactate import presented as in (A).
+(**C**) Mass isotopomer distributions of lactate (LAC), fructose-bisphosphate (FBP), pyruvate (PYR), and citrate (CIT) in LFs following 72 h labeling with [U-^13^C~3~] lactate.
+Data are mean ± SEM.
+Comparisons were made using linear mixed effects models with treatment as the fixed effect and biological replicate as a random effect with p-values adjusted using the multivariate *t* distribution (n = 4, p < 0.05 indicated as * 0.5% *v.* 21% oxygen, † BAY *v.* DMSO, ‡ Δoxygen *v.* ΔBAY).
+
+(ref:m5) **Metabolomic analysis of molidustat treatment in normoxia and hypoxia.** (**A**) LFs were cultured in standard growth medium and treated with molidustat (BAY, 10 μM) or vehicle (DMSO, 0.1%) in 21% or 0.5% oxygen conditions for 72 h.
+Growth rates were determined by linear fitting of log-transformed growth curves.
+(**B**) Glucose uptake in LFs treated with BAY cultured in hypoxia (note reversed *y*-axis).
+(**C**) Lactate efflux in LFs treated with BAY cultured in hypoxia.
+Data are mean ± SEM.
+Comparisons were made using linear mixed effects models with oxygen × treatment as the fixed effects and biological replicate as a random effect with p-values adjusted using the multivariate *t* distribution (* p < 0.05 BAY *v.* DMSO within a given oxygen exposure).
+(**D**) Principal component analysis of intracellular metabolites following 72 h of treatment as in (A) suggests a dominant effect of hypoxia over pharmacologic PHD inhibition on the metabolome (n = 7).
+(**E**) Volcano plot of intracellular metabolites of BAY-treated cells cultures in 21% or 0.5% oxygen.
+*Blue* circles indicate increased levels in hypoxia while *red* circles indicate decreases with hypoxia.
+(**F**) Metabolite set enrichment analysis of metabolites from (E).
+KEGG pathways with p-values < 0.05 are shown.
+(**G**) Leading edge analysis of the TCA cycle metabolite set.
+Negative values indicate relative enrichment associated with BAY treatment compared to hypoxia treatment.
+Abbreviations: PYR, pyruvate; SUC, succinate; PEP, phosphoenolpyruvate; CIT, citrate; AKG, α-ketoglutarate; MAL, malate; ACO, aconitate; FUM, fumarate.
+(**H**) NAD^+^ from LFs cultured as in (A).
+(**I**) NADH from LFs cultured as in (A).
+(**J**) NADH/NAD^+^ ratio from (H) and (I).
+NAD and NADH were determined by enzymatic assay.
+Data are mean ± SEM (n = 5).
+Comparisons were made using linear mixed effects models with treatment as the fixed effect and biological replicate as a random effect with p-values adjusted using the multivariate *t* distribution (p < 0.05 with *black* * indicating a significant effect of treatment within a given oxygen tension and *colored* * indicating a significant effect of treatment within a given oxygen tension denoted by the color).
+
+(ref:m6) **Transcriptomic analysis of molidustat treatment in normoxia and hypoxia.** (**A**) Principal component analysis of transcriptional changes following 72 h of treatment with 0.5% oxygen or molidustat (BAY), separately or together (n = 4).
+(**B**) Volcano plot illustrating the effects of hypoxia in BAY-treated cells on gene expression.
+(**C**) Gene set enrichment analysis of transcripts from (B).
+(**D**) Volcano plot of those transcripts comprising the E2F Targets and G2/M Checkpoint Hallmark gene sets.
+(**E**) Volcano plot of those transcripts comprising the MYC Targets V1 and V2 Hallmark gene sets.
+(**F**) Volcano plot illustrating the results of a transcription factor enrichment analysis suggests mechanisms for differential regulation of gene expression following hypoxia or BAY treatment.
+
+(ref:m7) **MYC regulates HIF-dependent glycolytic flux.** (**A**) Representative immunoblot of MYC protein expression in lung fibroblasts following 72 h of treatment with 0.5% oxygen or molidustat (BAY, B).
+(**B**) Quantification of band densities from (A).
+(**C**) Working model illustrating that hypoxia-induced MYC antagonizes HIF-dependent metabolic and transcriptional events.
+(**D**) Representative immunoblot of lung fibroblasts (LFs) treated with siRNA targeting MYC (M) demonstrating adequate protein knockdown.
+(**E**) Growth rates of MYC-knockdown cells cultured in hypoxia.
+(**F**) Lactate efflux rates of MYC-knockdown cells cultured in hypoxia.
+(**G**) Representative immunoblot of LFs treated with MYC adenovirus.
+(**H**) Growth rates of MYC overexpressing cells cultured with BAY.
+(**I**) Lactate efflux rates of MYC overexpressing cells cultured with BAY.
+Data are mean ± SEM of n = 3-7 biological replicates.
+Comparisons were made using linear mixed effects models with oxygen × treatment as the fixed effects and biological replicate as a random effect.
+The multivariate *t* distribution was used to determine significant differences (p < 0.05) as indicated by black * for differences within groups and colored * for differences between groups defined by the *x*-axis.
 
@@
       "manuscript/figures", 
-      package = "package"
+      package = "Copeland.2022.hypoxia.flux"
     ), 
@@
 ```{r m1, fig.cap = "(ref:m1)"}
-# knitr::include_graphics(
-#   system.file(
-#     "manuscript/figures/m1.png",
-#     package = "package"
-#   )
-# )
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/m1.png",
+    package = "Copeland.2022.hypoxia.flux"
+  )
+)
 ```
@@
 \newpage
+
+```{r m2, fig.cap = "(ref:m2)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/m2.png",
+    package = "Copeland.2022.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
+```{r m3, fig.cap = "(ref:m3)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/m3.png",
+    package = "Copeland.2022.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
+```{r m4, fig.cap = "(ref:m4)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/m4.png",
+    package = "Copeland.2022.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
+```{r m5, fig.cap = "(ref:m5)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/m5.png",
+    package = "Copeland.2022.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
+```{r m6, fig.cap = "(ref:m6)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/m6.png",
+    package = "Copeland.2022.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
+```{r m7, fig.cap = "(ref:m7)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/m7.png",
+    package = "Copeland.2022.hypoxia.flux"
+  )
+)
+```

### 2022-07-29 R/1_figures.R

@@ -178,7 +178,7 @@ plot_growth_rates <- function(
   annot <-
     lmerTest::lmer(mu ~ group + (1 | date), data = x) |>
     emmeans::emmeans(~ group) |>
-    pairs() |>
+    pairs(adjust = "mvt") |>
     broom::tidy() |>
     dplyr::mutate(
       group = stringr::str_extract(contrast, "(?<= - ).*"),
@@ -1589,7 +1589,7 @@ plot_nad <- function(df, annot, metab, ylab) {
       aesthetics = c("fill", "color")
     ) +
     ggplot2::labs(
-      x = "Oxygen",
+      x = "Treatment",
       y = ylab,
       fill = NULL,
       color = NULL
@@ -1787,7 +1787,7 @@ arrange_m6 <- function(p1, p2, p3, p4, p5, p6, p7) {
   p1 + p2 + p3 + p4 + p5 + p6 +
     theme_patchwork(
       design = layout,
-      widths = unit(2.5, "in"),
+      widths = unit(1.5, "in"),
       heights = unit(c(1.5), "in")
     )
 }
@@ -1805,6 +1805,74 @@ arrange_m7 <- function(p1, p2, p3, p4, p5, p6, p7, p8, p9) {
       widths = unit(1, "in"),
       heights = unit(c(1), "in")
     )
+}
+
+create_resources <- function() {
+  tibble::tribble(
+    ~category, ~`REAGENT or RESOURCE`, ~SOURCE, ~IDENTIFIER,
+    "Antibodies", "HIF-1α", "BD Biosciences", "610958",
+    "Antibodies", "c-MYC", "Cell Signaling Technologies", "D84C12",
+    "Antibodies", "LDHA", "Cell Signaling Technologies", "2012",
+    "Antibodies", "HRP-α-Rabbit IgG", "Cell Signaling Technologies", "7074",
+    "Antibodies", "HRP-α-Mouse IgG", "Cell Signaling Technologies", "7076",
+    "Bacterial and virus strains", "c-MYC adenovirus", "Vector Biolabs", "1285",
+    "Bacterial and virus strains", "YFP adenovirus", "Oldham et al., 2015", "",
+    "Chemicals, peptides, and recombinant proteins", "[1,2-^13^C~1~] glucose", "Cambridge Isotope Labs", "CLM-504-PK",
+    "Chemicals, peptides, and recombinant proteins", "[U-^13^C~6~] glucose", "Cambridge Isotope Labs", "CLM-1396-PK",
[…]

### 2022-07-29 _targets.R

@@ -718,46 +718,46 @@ list(
 
   # cosmos ------------------------------------------------------------------
 
-  # tar_target(
-  #   carnival_options,
-  #   set_carnival_options()
-  # ),
-  # tar_target(
-  #   cosmos_network,
-  #   get_cosmos_network()
-  # ),
-  # tar_target(
-  #   cosmos_tf,
-  #   format_tf(tfea_res_int)
-  # ),
-  # tar_target(
-  #   cosmos_metab,
-  #   format_metab(metab_tar_res_int)
-  # ),
-  # tar_target(
-  #   cosmos_deg,
-  #   format_deg(deg_int)
-  # ),
-  # tar_target(
-  #   cosmos_prep_forward,
-  #   preprocess("forward", cosmos_network, cosmos_tf, cosmos_metab, cosmos_deg, carnival_options)
-  # ),
-  # tar_target(
-  #   cosmos_prep_reverse,
-  #   preprocess("reverse", cosmos_network, cosmos_tf, cosmos_metab, cosmos_deg, carnival_options)
-  # ),
-  # tar_target(
-  #   cosmos_forward,
-  #   run_cosmos("forward", cosmos_prep_forward, carnival_options)
-  # ),
-  # tar_target(
-  #   cosmos_reverse,
-  #   run_cosmos("reverse", cosmos_prep_reverse, carnival_options)
-  # ),
-  # tar_target(
-  #   cosmos_res,
-  #   format_cosmos(cosmos_forward, cosmos_reverse)
-  # ),
+  tar_target(
+    carnival_options,
+    set_carnival_options()
+  ),
+  tar_target(
+    cosmos_network,
+    get_cosmos_network()
+  ),
+  tar_target(
+    cosmos_tf,
+    format_tf(tfea_res_int)
+  ),
+  tar_target(
+    cosmos_metab,
+    format_metab(metab_tar_res_int)
+  ),
+  tar_target(
+    cosmos_deg,
+    format_deg(deg_int)
+  ),
+  tar_target(
+    cosmos_prep_forward,
+    preprocess("forward", cosmos_network, cosmos_tf, cosmos_metab, cosmos_deg, carnival_options)
+  ),
+  tar_target(
+    cosmos_prep_reverse,
+    preprocess("reverse", cosmos_network, cosmos_tf, cosmos_metab, cosmos_deg, carnival_options)
+  ),
+  tar_target(
+    cosmos_forward,
[…]

### 2022-08-01 manuscript/supplement.Rmd
finish up revised manuscript draft

---
title: "Supplemental Information"
subtitle: "MYC overrides HIF to regulate proliferating primary cell metabolism in hypoxia"

author:
- Courtney A. Copeland: 
    institute: bwh, hms
- Benjamin A. Olenchock: 
    institute: bwh, hms
- David R. Ziehr:
    institute: bwh, mgh, hms
- Sarah McGarrity: 
    institute: bwh, hms, ui
- Kevin Leahy: 
    institute: bwh, hms
- Jamey D. Young: 
    institute: vu
- Joseph Loscalzo: 
    institute: bwh, hms
- William M. Oldham:
    institute: bwh, hms
    correspondence: true
    email: woldham@bwh.harvard.edu

institute:
- bwh: Department of Medicine, Brigham and Women's Hospital, Boston, MA 02115, U.S.A.
- hms: Department of Medicine, Harvard Medical School, Boston, MA 02115, U.S.A.
- mgh: Department of Medicine, Massachusetts General Hospital, Boston, MA, 02114 U.S.A.
- ui: Center for Systems Biology, School of Health Sciences, University of Iceland, Reykjavik, Iceland
- vu: Departments of Chemical & Biomolecular Engineering and Molecular Physiology & Biophysics, Vanderbilt University, Nashville, TN 37240, U.S.A.

bibliography_main: library.json
bibliography_software: packages.bib
csl: embo.csl
link-citations: yes

output:
  bookdown::word_document2:
    reference_docx: !expr system.file("manuscript", "template.docx", package = "Copeland.2022.hypoxia.flux")
    df_print: kable
    fig_caption: true
    number_sections: false
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
      - '--lua-filter=pagebreak.lua'
      # - '--lua-filter=multiple-bibliographies.lua'

compact-title: true
geometry: margin = 0.75in
fontsize: 11pt
mainfont: Calibri
graphics: yes

nocite: |
  @R-base, @R-bookdown, @R-devtools, @R-dplyr, @R-emmeans, @R-forcats, @R-ggplot2, 
  @R-kableExtra, @R-knitr, @R-lme4, @R-lmerTest, @R-lubridate, @R-magick, @R-magrittr, 
  @R-MASS, @R-patchwork, @R-pbkrtest, @R-purrr, @R-RColorBrewer, @R-readr, 
  @R-readxl, @R-renv, @R-rlang, @R-rmarkdown, @R-roxygen2, @R-stringr, @R-tibble, 
  @R-tidyr, @R-tidyverse, @R-tinytex, @R-usethis, @R-viridis, @R-wmo, 
  @bookdown2016, @ggplot22016, @knitr2015, @lme42015, @lmerTest2017, @lubridate2011, 
  @MASS2002, @pbkrtest2014, @rmarkdown2018, @tidyverse2019, tinytex2019
---

```{r setup, include = FALSE}
# library(Copeland.2022.hypoxia.flux)
devtools::load_all()

knitr::opts_chunk$set(
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE, 
  out.width = "100%", 
  out.extra = ""
)

rbbt::bbt_update_bib(
  path = "manuscript/manuscript.Rmd",
  ignore = c("R-base"), 
  path_bib = "manuscript/library.json"
)

options(
  tinytex.verbose = TRUE, 
  knitr.graphics.rel_path = FALSE
)

# data for paper
hyp <- targets::tar_read(lf_hypoxia_table)
bay <- targets::tar_read(lf_bay_table)
pasmc <- targets::tar_read(pasmc_hypoxia_table)

# figure dependencies
targets::tar_load(
  c(
    s1_figure, 
    s2_figure, 
    s3_figure,
    s4_figure, 
    s5_figure, 
    s6_figure, 
    s7_figure, 
    s8_figure, 
    s9_figure
  )
)
```

\newpage

# Tables

```{r t1, results = "asis"}
hyp %>% 
  flextable::set_caption(
    "Lung fibroblast fluxes in 21% and 0.5% oxygen", 
    autonum = officer::run_autonum()
  ) %>% 
  flextable::flextable_to_rmd()
```

\newpage

```{r t2, results = "asis"}
bay %>% 
  flextable::set_caption(
    "Lung fibroblast fluxes following DMSO and BAY treatment", 
    autonum = officer::run_autonum()
  ) %>% 
  flextable::flextable_to_rmd()
```

\newpage

```{r t3, results = "asis"}
pasmc %>% 
  flextable::set_caption(
    "PASMC fluxes in 21% and 0.5% oxygen", 
    autonum = officer::run_autonum()
  ) %>% 
  flextable::flextable_to_rmd()
```

\newpage

# Figure legends

(ref:s1) **Supporting data for extracellular flux calculations.** (**A**) Cell viability as assessed by live/dead cell staining with acridine orange plus propidium iodide staining did not differ between 21% and 0.5% oxygen culture conditions (n = 3 technical replicates). (**B**) Standard curve of lung fibroblast (LF) cell count *v.* total DNA by PicoGreen measurement used to interpolate cell numbers from DNA measurements. Data are mean ± SEM of three biological replicates. (**C**) Standard curve of PASMC cell count *v.* total DNA as in (B). (**D**) Total DNA measurements were compared to direct cell counts over the experimental time course. Cell counts and total DNA were obtained from the same sample wells. The slopes of the best-fit lines for 21% (*red*) and 0.5% (*blue*) samples were not statistically different. (**E**) Predicted well volumes were estimated from the change in culture plate mass over the experimental time course. Evaporation rates were different depending on the culture conditions and treatment. Although the mean evaporation rate is depicted, experiment-specific evaporation rates were used to calculate fluxes for each biological replicate (**F**) Metabolite accumulation (positive values) and degradation (negative values) rates. Data are mean ± SEM of 3-8 biological replicates. Rates significantly different from 0 (\*) based on a probability value < 0.05 using Student's one-sample *t*-test were incorporated into flux calculations.

(ref:s2) **Effects of hypoxia on extracellular metabolite fluxes in lung fibroblasts.** (**A**) Lung fibroblasts (LFs) were cultured in 21% (*red*) or 0.2% (*dark blue*) oxygen beginning 24 h prior to time 0. Samples were collected every 24 h for 72 h. (**B**) Growth curves of LFs in each experimental condition (n = 4). (**C**) Growth rates from (B) were determined by robust linear modeling of log-transformed growth curves. (**D**) Representative immunoblot of LF protein lysates cultured as in (A). (**E**) Relative change in HIF-1α protein levels from (D) normalized to 21% oxygen at time 0 (n = 4). (**F**) Relative change in GLUT1 mRNA levels normalized to 21% oxygen treatment at time 0 (n = 4). (**G**) Relative change in LDHA mRNA levels as in (F). (**H**) Relative change in LDHA protein levels as in (E). (**I**) Extracellular fluxes of glucose (GLC) and lactate (LAC) (n = 4). (**J**) Extracellular fluxes of pyruvate (PYR) and amino acids. Data are mean ± SEM. Comparisons were made using linear mixed effects models with oxygen as the fixed effect and biological replicate as a random effect with p-values adjusted using the multivariate *t* distribution (* p < 0.05).

(ref:s3) **Effects of hypoxia on extracellular metabolite fluxes in pulmonary artery smooth muscle cells.** (**A**) Pulmonary artery smooth muscle cells (PASMCs) were cultured in 21% (*red*) or 0.5% (*blue*) oxygen beginning 24 h prior to time 0. Samples were collected every 12 h for 48 h. (**B**) Growth curves of LFs in each experimental condition (n = 4). (**C**) Growth rates from (B) were determined by robust linear modeling of log-transformed growth curves. (**D**) Representative immunoblot of LF protein lysates cultured as in (A). (**E**) Relative change in HIF-1α protein levels from (D) normalized to 21% oxygen at time 0 (n = 4). (**F**) Relative change in GLUT1 mRNA levels normalized to 21% oxygen treatment at time 0 (n = 4). (**G**) Relative change in LDHA mRNA levels as in (F). (**H**) Relative change in LDHA protein levels as in (E). (**I**) Extracellular fluxes of glucose (GLC) and lactate (LAC) (n = 4). (**J**) Extracellular fluxes of pyruvate (PYR) and amino acids. Data are mean ± SEM. Comparisons were made using linear mixed effects models with oxygen as the fixed effect and biological replicate as a random effect with p-values adjusted using the multivariate *t* distribution (* p < 0.05).

(ref:s4) **Mass isotopomer distributions after 72 h of labeling in lung fibroblasts.** Lung fibroblasts (LFs) were labeled with the indicated tracers and intracellular metabolites were analyzed by LC-MS after 72 h. Mass isotopomer distributions were adjusted for natural abundance. Data are the mean ± SEM of 4 biological replicates. Significant differences in labeling patterns between 21% and 0.5% oxygen (\*), DMSO and BAY treatment (†), and 0.5% oxygen and BAY treatment (‡) for each combination of metabolite and tracer are highlighted.

(ref:s5) **Mass isotopomer distributions after 48 h of labeling in pulmonary artery smooth muscle cells.** Pulmonary artery smooth muscle cells (PASMCs) were labeled with the indicated tracers and intracellular metabolites were analyzed by LC-MS after 36 h. Mass isotopomer distributions were adjusted for natural abundance. Data are the mean ± SEM of 4 biological replicates. Significant differences in labeling patterns between 21% and 0.5% oxygen (\*) for each combination of metabolite and tracer are highlighted.

(ref:s6) **Isotope incorporation over the labeling time course.** LFs were cultured in 21% or 0.5% oxygen and labeled with the indicated tracers. Intracellular metabolites were analyzed by LC-MS (FBP, fructose-bisphosphate; PYR, pyruvate; CIT, citrate; MAL, malate). Mass isotopomer distributions were calculated and adjusted for natural abundance. Data show the total amount of metabolite labeling (*i.e.,* 1 - M0 fractional abundance). Data are the mean ± SEM of 4 biological replicates.

(ref:s7) **Isotopically non-stationary metabolic flux analysis.** (**A**) Metabolic flux model of LF metabolism in 21% oxygen. Arrows are colored by log~10~(flux). (**B**) Metabolic flux model of PASMC metabolism in 21% oxygen as in (A). (**C**) The ratio between PASMC and LF fluxes in normoxia were calculated. Fluxes with non-overlapping confidence intervals are highlighted with arrows colored according to the magnitude of the change. Arrow thickness corresponds to the absolute flux measured in LFs. (**D**) Ratio of metabolic fluxes in 0.5% oxygen compared to 21% oxygen in PASMCs. (**E**) LF fluxes were normalized to cell growth rate. Graph depicts the ratio of normalized metabolic fluxes in LFs cultured in 0.5% oxygen compared to 21% oxygen control. Fluxes with non-overlapping confidence intervals are highlighted to indicate significant changes.

(ref:s8) **Metabolomic profiling of hypoxia and BAY treated lung fibroblasts.** (**A**) Volcano plot of differentially regulated metabolites following 0.5% oxygen culture. (**B**) Volcano plot of differentially regulated metabolites by BAY treatment in 21% oxygen. (**C**) Venn diagram illustrating the overlap among metabolites differentially regulated by hypoxia (*blue*) or BAY treatment (*purple*). (**D**) Metabolite enrichment in KEGG pathways following hypoxia. (**E**) Metabolite enrichment in KEGG pathways following BAY. Only significantly enriched metabolite sets with p < 0.05 are shown. NES, normalized enrichment score.

(ref:s9) **Transcriptomic profiling of hypoxia and BAY treated lung fibroblasts.** (**A**) Volcano plot of differentially expressed genes following 0.5% oxygen culture. (**B**) Volcano plot of differentially expressed genes following BAY treatment. (**C**) Venn diagram illustrating the number of differentially expressed transcripts following hypoxia (*blue*) or BAY treatment (*purple*). (**D**) Venn diagram illustrating the number of differentially enriched Hallmark gene sets with  hypoxia (*blue*) or BAY treatment (*purple*). (**E**) Gene set enrichment results of hypoxia-treated cells. (**F**) Gene set enrichment results of BAY-treated cells. All gene sets listed were significantly enriched at FDR < 0.05. NES, normalized enrichment score. (**G**) Volcano plot illustrating the results of a transcription factor enrichment analysis in hypoxia-treated cells. (**H**) Volcano plot illustrating the results of a transcription factor enrichment analysis in BAY-treated cells. (**I**) Venn diagram illustrating the overlap among enriched transcription factors following hypoxia or BAY treatment.

```{r}
figs <- 
  list.files(
    system.file(
      "manuscript/figures", 
      package = "Copeland.2022.hypoxia.flux"
    ), 
    pattern = "s\\d{1}\\.png"
  ) %>% 
  rlang::set_names(stringr::str_extract(., "^.*(?=\\.png)"))
```

```{r results = "asis"}
out <- 
  purrr::imap(
    figs, 
    ~ cat("Supplementary Figure ", stringr::str_extract(.y, "\\d"), ": (ref:", .y, ")\n\n", sep = "")
  )
```

\newpage

# Figures

```{r s1, fig.cap = "(ref:s1)"}
knitr::include_graphics(
  system.file(
    "manuscript/figures/s1.png",
    package = "Copeland.2022.hypoxia.flux"
  )
)
```

\newpage

```{r s2, fig.cap = "(ref:s2)"}
knitr::include_graphics(
  system.file(
    "manuscript/figures/s2.png",
    package = "Copeland.2022.hypoxia.flux"
  )
)
```

\newpage

```{r s3, fig.cap = "(ref:s3)"}
knitr::include_graphics(
  system.file(
    "manuscript/figures/s3.png",
    package = "Copeland.2022.hypoxia.flux"
  )
)
```

\newpage

```{r s4, fig.cap = "(ref:s4)"}
knitr::include_graphics(
  system.file(
    "manuscript/figures/s4.png",
    package = "Copeland.2022.hypoxia.flux"
  )
)
```

\newpage

```{r s5, fig.cap = "(ref:s5)"}
knitr::include_graphics(
  system.file(
    "manuscript/figures/s5.png",
    package = "Copeland.2022.hypoxia.flux"
  )
)
```

\newpage

```{r s6, fig.cap = "(ref:s6)"}
knitr::include_graphics(
  system.file(
    "manuscript/figures/s6.png",
    package = "Copeland.2022.hypoxia.flux"
  )
)
```

\newpage

```{r s7, fig.cap = "(ref:s7)"}
knitr::include_graphics(
  system.file(
    "manuscript/figures/s7.png",
    package = "Copeland.2022.hypoxia.flux"
  )
)
```

\newpage

```{r s8, fig.cap = "(ref:s8)"}
knitr::include_graphics(
  system.file(
    "manuscript/figures/s8.png",
    package = "Copeland.2022.hypoxia.flux"
  )
)
```

\newpage

```{r s9, fig.cap = "(ref:s9)"}
knitr::include_graphics(
  system.file(
    "manuscript/figures/s9.png",
    package = "Copeland.2022.hypoxia.flux"
  )
)
```

\newpage

# References

::: {#refs_main} 
:::

## Software

::: {#refs_software} 
:::

### 2022-08-01 R/1_cosmos.R

@@ -77,7 +77,7 @@ preprocess <- function(
       metabolic_data = metabolites,
       diff_expression_data = transcripts,
       diff_exp_threshold = 1,
-      maximum_network_depth = 6,
+      maximum_network_depth = 5,
       remove_unexpressed_nodes = remove_nodes,
       CARNIVAL_options = carnival_options
     )

### 2022-08-01 R/1_figures.R

@@ -1751,7 +1751,7 @@ arrange_s8 <- function(p1, p2, p3, p4, p5) {
   layout <- "
   ad
   be
-  c#
+  ce
   "
 
   p1 + p2 + p3 + p4 + p5 +
@@ -1762,19 +1762,20 @@ arrange_s8 <- function(p1, p2, p3, p4, p5) {
     )
 }
 
-arrange_s9 <- function(p1, p2, p3, p4, p5, p6, p7, p8) {
+arrange_s9 <- function(p1, p2, p3, p4, p5, p6, p7, p8, p9) {
   layout <- "
   aeg
   beh
-  cf#
+  #e#
+  cfi
   df#
   "
 
-  p1 + p2 + p3 + p4 + p5 + p6 + p7 + p8 +
+  p1 + p2 + p3 + p4 + p5 + p6 + p7 + p8 + p9 +
     theme_patchwork(
       design = layout,
       widths = unit(2.5, "in"),
-      heights = unit(c(1.5), "in")
+      heights = unit(c(1.5, 1.5, 1.1, 1.5, 1.5), "in")
     )
 }
 
@@ -1883,8 +1884,8 @@ format_flux_table <- function(
     ssr_exp = NULL
 ) {
 
-  big_border = flextable::fp_border_default(color = "black", width = 1)
-  small_border = flextable::fp_border_default(color = "black", width = 0.25)
+  big_border <- flextable::fp_border_default(color = "black", width = 1)
+  small_border <- flextable::fp_border_default(color = "black", width = 0.25)
 
   conditions <- c(unique(flux_differences$ctl), unique(flux_differences$exp))
 

### 2022-08-01 R/1_rnaseq.R

@@ -630,6 +630,59 @@ plot_gsea_venn <- function(hyp, bay) {
     )
 }
 
+plot_tfea_venn <- function(hyp, bay) {
+  nm <- hyp$tf
+  hyp_pathway <-
+    hyp |>
+    dplyr::filter(adj.P.Val < 0.05) |>
+    dplyr::pull(tf)
+  bay_pathway <-
+    bay|>
+    dplyr::filter(adj.P.Val < 0.05) |>
+    dplyr::pull(tf)
+  bay_deg <- nm %in% bay_pathway
+  hyp_deg <- nm %in% hyp_pathway
+
+  tibble::tibble(nm, hyp_deg, bay_deg) |>
+    ggplot2::ggplot() +
+    ggplot2::aes(A = hyp_deg, B = bay_deg) +
+    ggvenn::geom_venn(
+      set_names = c("0.5%", "BAY"),
+      digits = 0,
+      show_percentage = TRUE,
+      fill_color = clrs[c(2, 4)],
+      fill_alpha = 0.5,
+      stroke_size = 0.25,
+      set_name_size = 8/ggplot2::.pt,
+      text_size = 6/ggplot2::.pt
+    ) +
+    ggplot2::annotate(
+      geom = "text",
+      x = 1,
+      y = -1.2,
+      label = "96 total",
+      size = 6/ggplot2::.pt
+    ) +
+    theme_plots() +
+    ggplot2::labs(
+      x = NULL,
+      y = NULL,
+      title = "Transcription factors"
+    ) +
+    ggplot2::coord_fixed(
+      xlim = c(-1.75, 1.75),
+      ylim = c(-1.2, 1.2),
+      clip = "off"
+    ) +
+    ggplot2::theme(
+      panel.border = ggplot2::element_blank(),
+      axis.text = ggplot2::element_blank(),
+      axis.ticks = ggplot2::element_blank(),
+      axis.line = ggplot2::element_blank(),
+      plot.title = ggplot2::element_text(size = 8),
+    )
+}
+
 plot_pathway_volcanoes <- function(deg, pathways, sets, title, nudge = 3) {
   targets <-
     pathways[sets] |>

### 2022-08-01 _targets.R

@@ -291,7 +291,7 @@ list(
 
   tar_target(
     reactions_file,
-    path_to_reports("modeling/matlab-input/reactions_02.csv"),
+    path_to_reports("modeling/matlab-input/reactions.csv"),
     format = "file"
   ),
   tar_target(
@@ -582,6 +582,10 @@ list(
     gsea_venn,
     plot_gsea_venn(gsea_hyp, gsea_bay)
   ),
+  tar_target(
+    tfea_venn,
+    plot_tfea_venn(tfea_res_hyp, tfea_res_bay)
+  ),
   tar_render(
     rnaseq_report,
     path = path_to_reports("rnaseq.Rmd"),
@@ -635,7 +639,7 @@ list(
       names = list("hyp", "bay", "hyp_bay", "int"),
       colors = list(clrs[2:1], clrs[4:3], clrs[2:1], clrs[c(2, 4)]),
       xlab = list("Hypoxia/Normoxia", "BAY/DMSO", "Hypoxia/Normoxia", "ΔHypoxia/ΔBAY"),
-      title = list("Hypoxia", "BAY", "Hypoxia/Normoxia in BAY", "ΔHypoxia/ΔBAY"),
+      title = list("Hypoxia/Normoxia", "BAY/DMSO", "Hypoxia/Normoxia in BAY", "ΔHypoxia/ΔBAY"),
       filename = stringr::str_c("msea_", list("hyp", "bay", "hyp_bay", "int"))
     ),
     names = names,
@@ -726,37 +730,47 @@ list(
     cosmos_network,
     get_cosmos_network()
   ),
-  tar_target(
-    cosmos_tf,
-    format_tf(tfea_res_int)
-  ),
-  tar_target(
-    cosmos_metab,
-    format_metab(metab_tar_res_int)
-  ),
-  tar_target(
-    cosmos_deg,
-    format_deg(deg_int)
-  ),
-  tar_target(
-    cosmos_prep_forward,
-    preprocess("forward", cosmos_network, cosmos_tf, cosmos_metab, cosmos_deg, carnival_options)
-  ),
-  tar_target(
-    cosmos_prep_reverse,
-    preprocess("reverse", cosmos_network, cosmos_tf, cosmos_metab, cosmos_deg, carnival_options)
-  ),
-  tar_target(
-    cosmos_forward,
-    run_cosmos("forward", cosmos_prep_forward, carnival_options)
-  ),
-  tar_target(
[…]

### 2022-08-01 analysis/modeling/lf_21.m

@@ -3,8 +3,8 @@ function m = normoxia()
 % setup
 oxygen = '21%';
 treatment = 'None';
-in_path = '/Users/will/Dropbox (Partners HealthCare)/_data/Copeland.2022.hypoxia.flux/analysis/modeling/matlab-input/';
-out_path = '/Users/will/Dropbox (Partners HealthCare)/_data/Copeland.2022.hypoxia.flux/analysis/modeling/';
+in_path = '/Users/will/Dropbox (Partners HealthCare)/Copeland.2021.hypoxia.flux/inst/analysis/modeling/matlab-input/';
+out_path = '/Users/will/Dropbox (Partners HealthCare)/Copeland.2021.hypoxia.flux/inst/analysis/modeling/';
 cell_type = 'lf';
 addpath(genpath(in_path));
 output = [out_path cell_type '_21/' cell_type '_21_' datestr(now,'yyyy-mm-dd-HH-MM-SS') '.mat'];
@@ -19,7 +19,7 @@ biomasseqn = [biomasseqn num2str(biomass{height(biomass), 2}) ' ' char(biomass{h
 clear biomass;
 
 % reaction equations
-reactions = readtable('reactions_03.csv');
+reactions = readtable('reactions.csv');
 % reactions = rmmissing(reactions);
 reaction_list = [biomasseqn; reactions.equation]; 
 reaction_ids = ['BIOMASS'; reactions.name];
@@ -85,8 +85,8 @@ for label = {'glc2' 'glc6' 'q5'}
             y = [y c];
             h = [h time];
         end
-        d{n}{char(metabolite)}.idvs = idv(y);
-        d{n}{char(metabolite)}.idvs.time = h;
+        d{n}{char(metabolite)}.mdvs = mdv(y);
+        d{n}{char(metabolite)}.mdvs.time = h;
     end
     n = 1 + n;
 end
@@ -114,7 +114,7 @@ clear a b c y n x z time compound label;
 sem_value = 0.013;
 for i=1:length(d)
     for j=1:length(d{i})
-        d{i}{d{i}.id{j}}.idvs.std = sem_value * ones(length(d{i}{d{i}.id{j}}.idvs.val),1);
+        d{i}{d{i}.id{j}}.mdvs.std = sem_value * ones(length(d{i}{d{i}.id{j}}.mdvs.val),1);
[…]

### 2022-08-01 manuscript/manuscript.Rmd
add page numbers

@@
 
-rbbt::bbt_update_bib(
-  path = "manuscript/manuscript.Rmd",
-  ignore = c("R-base"), 
-  path_bib = "manuscript/library.json"
-)
+# rbbt::bbt_update_bib(
+#   path = "manuscript/manuscript.Rmd",
+#   ignore = c("R-base"),
+#   path_bib = "manuscript/library.json"
+# )
 
@@
 Differentially regulated metabolites were identified using limma [@ritchie2015a].
-Metabolite set enrichment analysis was performed using the fgsea package [@korotkevich2021a] with KEGG metabolite pathways [@kanehisa2000].
+Metabolite set enrichment analysis was performed using the fgsea package [@korotkevich2021] with KEGG metabolite pathways [@kanehisa2000].
 
@@
 Differentially expressed transcripts were identified using DESeq2 [@love2014].
-Gene set enrichment and transcription factor enrichment was performed using the fgsea and DoRothEA R packages, respectively [@korotkevich2021a; @garcia-alonso2019].
+Gene set enrichment and transcription factor enrichment was performed using the fgsea and DoRothEA R packages, respectively [@korotkevich2021; @garcia-alonso2019].
 

### 2022-08-01 manuscript/manuscript.Rmd
corrections to text up to legends

@@
 
-Hypoxia requires metabolic adaptations to sustain cellular functions that underlie numerous physiologic and pathologic processes.
-While many studies have explored the metabolic consequences of hypoxia in cancer models, comparatively little is known about the metabolic response of primary cells to hypoxia.
+Hypoxia requires metabolic adaptations to sustain energetically demanding cellular activities that underpin many physiologic and pathologic processes.
+While many studies have explored the metabolic consequences of hypoxia in cancer cell models, comparatively little is known about the metabolic response of primary cells to hypoxia.
 Here, we performed metabolic flux analyses of proliferating human lung fibroblasts and pulmonary artery smooth muscle cells in hypoxia.
@@
 Multi-omic profiling of cells treated with hypoxia or molidustat, separately or together, revealed distinct molecular responses to hypoxia and pharmacologic PHD inhibition and suggested a critical role for MYC in modulating the HIF response in hypoxia.
-MYC knockdown in hypoxia increased lactate efflux while MYC overexpression in normoxia blunted the effects of molidustat treatment.
+MYC knockdown in hypoxia increased lactate efflux, while MYC overexpression in normoxia blunted the effects of molidustat treatment.
 Together, these data suggest that other factors, notably MYC, supersede the anticipated effects of HIF-dependent up-regulation of glycolytic gene expression on glycolytic flux in hypoxic proliferating primary cells.
@@
 
-Cellular responses to hypoxia propel many physiologic and pathologic activities from wound healing and angiogenesis to vascular remodeling and fibrosis [@semenza2012b; @lee2019].
+Cellular responses to hypoxia propel many physiologic and pathologic processes from wound healing and angiogenesis to vascular remodeling and fibrosis [@semenza2012b; @lee2019].
 These activities require cells to continue energetically demanding tasks, such as macromolecular biosynthesis and proliferation, despite limited oxygen availability.
@@
 
-In addition to metabolic changes designed to maintain energy supply, hypoxic cells also reduce energy demands through down-regulation of Na^+^/K^+^-ATPase, slowing protein translation, and attenuating cell proliferation [@wheaton2011; @hubbi2015].
-In particular, HIF-1α decreases cell proliferation by activating cyclin-dependent kinase inhibitor expression, inhibiting cell cycle checkpoint progression [@gardner2001], and antagonizing pro-proliferative MYC signaling [@koshiji2004].
+In addition to metabolic changes designed to maintain energy supply, hypoxic cells also reduce energy demand through down-regulation of Na^+^/K^+^-ATPase, slowing protein translation, and decreasing cell proliferation [@wheaton2011; @hubbi2015].
+In particular, HIF-1α decreases cell proliferation by activating cyclin-dependent kinase inhibitor expression, inhibiting cell-cycle checkpoint progression [@gardner2001], and antagonizing pro-proliferative MYC signaling [@koshiji2004].
 Despite these canonical effects of HIF-1α activation, there are many examples where cells continue to proliferate despite hypoxic stress, including cancer cells, stem cells, and lung vascular cells [@hubbi2015].
-How these cells meet the metabolic needs of sustained proliferation in hypoxia and how these adaptations are regulated are an active area of investigation [@jain2020; @oldham2015; @lee2020b].
+How these cells meet the metabolic needs of sustained proliferation in hypoxia and how these adaptations are regulated are active areas of investigation [@jain2020; @oldham2015; @lee2020b].
 Since hypoxia is a prominent feature of cancer biology as tumor growth outstrips blood supply, most detailed metabolic studies of hypoxic cell metabolism have used tumor cell models, yielding important insights into the metabolic pathobiology of cancer [@wise2011; @metallo2011; @melendez-rodriguez2019; @garcia-bermudez2018; @lee2019a; @jiang2016].
-For example, stable isotope tracing and metabolic flux analyses identified a critical role for reductive carboxylation of glutamine-derived α-ketoglutarate for lipid biosynthesis in supporting tumor growth [@metallo2011; @gameiro2013a; @scott2011; @wise2011], and metabolomic studies identified aspartate as a limiting metabolite for cancer cell proliferation under hypoxia [@garcia-bermudez2018].
+For example, stable isotope tracing and metabolic flux analyses identified a critical role for the reductive carboxylation of glutamine-derived α-ketoglutarate for lipid biosynthesis in supporting tumor growth [@metallo2011; @gameiro2013a; @scott2011; @wise2011], and metabolomic studies identified aspartate as a limiting metabolite for cancer cell proliferation under hypoxia [@garcia-bermudez2018].
 By contrast, comparatively little is known about metabolic adaptations of primary cells to hypoxia.
 Indeed, the importance of reductive carboxylation or aspartate biosynthesis remains to be elucidated in these cells.
-A more complete understanding of primary cell metabolic adaptation to hypoxia would provide an important context for understanding how metabolic reprogramming supports normal cellular responses to hypoxia, how these responses may be (mal)adaptive in a variety of disease contexts, and how the hypoxia metabolic program in primary cells differs from that observed in cancer cells.
+A more complete understanding of primary cell metabolic adaptations to hypoxia would provide an important context for understanding how metabolic reprogramming supports normal cellular responses to hypoxia, how these responses may be (mal)adaptive in a variety of disease contexts, and how the hypoxia metabolic program in primary cells differs from that observed in cancer cells.
 
 To address these questions, we have developed models of bioenergetic carbon flux in human lung fibroblasts (LFs) and pulmonary artery smooth muscle cells (PASMCs) cultured in 21% or 0.5% oxygen.
-These cells may be exposed to a wide range of oxygen concentrations *in vivo*, continue to proliferate despite hypoxic culture conditions *in vitro*, and play important roles in the pathology of non-cancerous diseases in which tissue hypoxia is a prominent feature.
+These cells may be exposed to a wide range of oxygen concentrations *in vivo*, continue to proliferate despite hypoxic culture conditions *in vitro*, and play important roles in the pathology of non-cancerous diseases in which tissue hypoxia is a prominent feature, including pulmonary hypertension and pulmonary fibrosis.
 We found that hypoxia fails to increase glycolysis in these primary cells despite robust up-regulation of the HIF-1α transcriptional program.
 In normoxia, HIF-1α stabilization by the PHD inhibitor molidustat (BAY-85-3934, "BAY") [@flamme2014] did increase glycolysis and lactate efflux; however, hypoxia blocked this response.
-These findings suggested the existence of important hypoxia-dependent regulatory mechanisms that override the metabolic consequences of HIF-1α-dependent up-regulation of glycolytic gene expression in human primary cells.
-Transcriptomic profiling suggested a critical role for the transcription factor MYC in the adaptive response to hypoxia.
-Using knockdown and overexpression approaches, we demonstrate that MYC attenuates HIF-driven glycolysis in hypoxia and following BAY treatment.
+These findings suggested that important hypoxia-dependent regulatory mechanisms override the metabolic consequences of HIF-1α-dependent glycolytic gene expression.
+Transcriptomic profiling identified a critical role for the transcription factor MYC in the adaptive response to hypoxia.
+Using knockdown and overexpression approaches, we demonstrated that MYC attenuates HIF-driven glycolysis in hypoxia and following HIF stabilization in normoxia.
 
@@
 These models assume that cells are at a metabolic pseudo-steady state over the experimental time course [@buescher2015].
-Exponential growth phase is thought to reflect metabolic pseudo-steady state as cells in the culture steadily divide at their maximal condition-specific rate, provided nutrient supply does not become limiting [@buescher2015; @ahn2011].
+Exponential growth phase is thought to reflect metabolic pseudo-steady state as cells in culture steadily divide at their maximal condition-specific rate, provided nutrient supply does not become limiting [@buescher2015; @ahn2011].
 Thus, we first set out to define experimental conditions to capture exponential growth phase in normoxic and hypoxic cultures.
@@
 
-Having identified experimental conditions for exponential growth, we next determined extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (**Figs \@ref(fig:m1)I-J**).
+Having identified experimental conditions for exponential growth, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (**Figs \@ref(fig:m1)I-J**).
 Flux calculations incorporated changes in cell number, extracellular metabolite concentrations, metabolite degradation rates, and medium evaporation over time [@murphy2013a] (**Figure S1**).
@@
 To test if more severe hypoxia would augment glycolysis, we culture cells in 0.2% ambient oxygen (**Fig S2**).
-These results were similar to 0.5% oxygen culture.
-We observed no change in glucose or lactate fluxes.
+Under these conditions, we observed no change in glucose or lactate fluxes, similar to 0.5% oxygen culture.
 To test if this unexpected response was unique to LFs, we next studied PASMCs under 0.5% oxygen conditions (**Fig S3**).
@@
 Cells were treated with BAY for 24 h to activate the HIF transcriptional program prior to sample collection (**Fig \@ref(fig:m2)A**).
-As with hypoxia, BAY decreased cell growth rate (**Figs \@ref(fig:m2)B-C**) and activated the HIF-1α transcriptional program (**Figs \@ref(fig:m1)D-H**).
+As with hypoxia, BAY decreased cell growth rate (**Figs \@ref(fig:m2)B-C**) and activated the HIF-1α transcriptional program (**Figs \@ref(fig:m2)D-H**).
 Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1α target gene transcription and protein expression.
-Unlike hypoxia, however, HIF-1α stabilization in normoxia markedly increased glucose uptake and lactate efflux (**Fig \@ref(fig:m1)I**), as expected based on the pattern of glycolytic gene expression that we observed.
+Unlike hypoxia, however, HIF-1α stabilization in normoxia markedly increased glucose uptake and lactate efflux (**Fig \@ref(fig:m2)I**).
+This finding was consistent with the increased expression of glycolytic genes that we observed.
 Interestingly, although hypoxia and BAY treatments resulted in similar increases in HIF-1α, GLUT1, and LDHA, the glycolytic response was markedly different.
@@
 
-In addition to glucose and lactate, we also determined the extracellular fluxes of pyruvate and amino acids (**Figs \@ref(fig:m1)J, \@ref(fig:m2)J**).
+In addition to glucose and lactate, we also determined the extracellular fluxes of pyruvate and amino acids (**Figs \@ref(fig:m1)J, \@ref(fig:m2)J, S2J, S3J**).
 To our knowledge, this is the first comprehensive extracellular flux profiling of key metabolic substrates in these primary cells.
@@
 
-In normoxia, the magnitude of intracellular metabolite fluxes was generally similar in LFs and PASMCs (**Figures S7A-C, Tables S1, S3**).
+In normoxia, the magnitude of intracellular metabolite fluxes was generally similar in LFs and PASMCs (**Figs S7A-C, Tables S1, S3**).
 Compared to LFs, PASMCs had slower rates of glycolysis and faster rates of TCA metabolism driven, in part, by increased glutamine uptake.
@@
 The increased TCA flux in PASMCs was driven by increased glutamine consumption.
-This finding is similar to a prior report of glutamine-driven oxidative phosphorylation in hypoxic cancer cells [@fan2013], where oxidative phosphorylation continued to provide the majority of cellular ATP even in 1% oxygen.
+This finding is similar to a prior report of glutamine-driven oxidative phosphorylation in hypoxic cancer cells [@fan2013], where oxidative phosphorylation continued to provide the majority of cellular ATP even at 1% oxygen.
 
 Given the global decrease in bioenergetic metabolic flux in hypoxic LFs, we hypothesized that these differences may be a consequence of decreased growth rate.
-After normalizing metabolite fluxes in normoxia and hypoxia to the cell growth rate, modest increases (~10%) in glycolytic flux were observed (**Figure S7E**).
-This finding suggests that, while glycolysis increases relative to growth rate in hypoxic cells, the regulators of cell proliferation rate override the consequences of the HIF-1α transcriptional program.
-Indeed, even after adjusting for cell growth rate, the relative increase in glycolytic flux is modest compared to the marked up-regulation of glycolytic protein levels.
-BAY treatment decreased cell proliferation rate (**Figure \@ref(fig:m2)B**), indicating that, unlike hypoxia, BAY treatment in normoxia uncouples cell proliferation and metabolic flux.
+After normalizing metabolite fluxes in normoxia and hypoxia to the cell growth rate, a modest increase (~10%) in glycolytic flux was observed (**Fig S7E**).
+This finding suggests that, while glycolysis increases relative to growth rate in hypoxic cells, regulators of cell proliferation rate override the consequences of the HIF-1α transcriptional program.
+Indeed, even after adjusting for cell growth rate, the relative increase in glycolytic flux is modest compared to the marked up-regulation of glycolytic protein levels and the glycolytic potential of these cells demonstrated by BAY treatment in normoxia.
+BAY treatment decreased cell proliferation rate (**Fig \@ref(fig:m2)B**), indicating that, unlike hypoxia, BAY treatment in normoxia uncouples cell proliferation and metabolic flux.
 
@@
 
-Although the metabolite exchange fluxes for bidirectional reactions tend to be poorly resolved by metabolic flux analysis, two observations are worth highlighting (**Tables S1-S3**).
+Although the metabolite exchange fluxes for bidirectional reactions tend to be poorly resolved by metabolic flux analysis [@wiechert2007a], two observations are worth highlighting (**Tables S1-S3**).
 First, consistent with the stable isotope tracing results, the modeled rate of reductive carboxylation through reverse flux by isocitrate dehydrogenase in LFs is low (~4 fmol/cell/h), unchanged by hypoxia, and modestly increased by BAY treatment.
-By contrast, the rate of reductive carboxylation increases 6-fold in PASMCs in hypoxia, highlighting an important role for this pathway in the metabolic response of PASMCs to hypoxia (**Fig \@ref(fig:m4)A**).
+By contrast, the rate of reductive carboxylation increases 6-fold in PASMCs in hypoxia, highlighting an important role for this pathway in the metabolic response of PASMCs to decreased oxygen availability (**Fig \@ref(fig:m4)A**).
 
@@
 Although lactate has been used less commonly than glucose and glutamine in stable isotope tracing studies, Faubert and colleagues [-@faubert2017] demonstrated lactate incorporation in human lung adenocarcinoma *in vivo*.
-In this study, lactate incorporation corresponded to regions of high glucose uptake as determined by [ ^18^F ]-fluorodeoxyglucose positron emission tomography, suggesting that lactate consumption can occur even in areas of high glucose utilization.
+In this study, lactate incorporation corresponded to regions of high glucose uptake as determined by [¹⁸F]-fluorodeoxyglucose positron emission tomography, suggesting that lactate consumption can occur even in areas of high glucose utilization.
 Subsequently, investigators have demonstrated the importance of lactate as a metabolic fuel *in vivo* [@hui2017; @hui2020].
@@
 However, when combined with 0.5% oxygen, BAY treatment was unable to enhance lactate efflux.
-These data indicate that hypoxia antagonizes the effects of HIF-1α activation on glycolytic flux in these cells.
+These data clearly demonstrate that hypoxia antagonizes the effects of HIF-1α activation on glycolytic flux in these cells.
 
@@
 To identify the metabolic changes that depend on hypoxia rather than BAY inhibition, we identified differentially regulated metabolites in BAY treated cells cultured in normoxia and hypoxia (**Fig \@ref(fig:m5)E**).
-Of 133 metabolites, 83 were significantly differentially regulated by hypoxia in BAY-treated cells treatments.
+Of 133 metabolites, 83 were significantly differentially regulated by hypoxia in BAY-treated cells.
 An enrichment analysis of these differentially regulated metabolites demonstrated up-regulation of arginine and proline metabolism and down-regulation of the TCA cycle as the most impacted by hypoxia in BAY treated cells (**Fig \@ref(fig:m5)F**).
@@
 Of the 9,923 differentially expressed genes across both conditions, 869 (9%) were unique to BAY treatment in normoxia, 4,002 (40%) were shared by BAY and hypoxia, while 5,052 (51%) were unique to 0.5% hypoxia culture (**Fig S9C**).
-This distribution of transcriptional changes was nearly identical to that observed for the metabolic changes (**Fig S8C**) where half of the hypoxia-mediated changes overlapped with BAY treatment and half were unique to hypoxia.
-Gene set enrichment analysis of these differentially regulated metabolites was performed using Molecular Signatures Database "Hallmark" gene sets [@liberzon2015] (**Figs S9D, S9E, S9F**).
+This distribution of transcriptional changes was nearly identical to that observed for the metabolite changes (**Fig S8C**), where half of the hypoxia-mediated changes overlapped with BAY treatment and half were unique to hypoxia.
+Gene set enrichment analysis of these differentially regulated metabolites was performed using Molecular Signatures Database "Hallmark" gene sets [@liberzon2015] (**Figs S9D-F**).
 As expected, both treatments were associated with enrichment of the "hypoxia" and "glycolysis" gene sets.
@@
 Consistent with our bioinformatic results, we observed increased MYC protein levels in hypoxia-treated cells, but not with BAY-treatment alone, where MYC was decreased (**Fig \@ref(fig:m7)A-B**).
-To test the hypothesis that hypoxia-induced MYC expression inhibits glycolysis in primary cells, we first combined MYC knockdown with hypoxia treatment (**Fig \@ref(fig:m7)C-E**).
-As expected, MYC-deficient cells proliferated more slowly in normoxia and MYC was essential for sustaining cell proliferation in hypoxia (**Fig \@ref(fig:m7)D**).
-Consistent with our hypothesis, MYC-knockdown cells demonstrated increased lactate efflux with hypoxia culture, unlike control siRNA-treated cells (**Fig \@ref(fig:m7)E**).
-We next performed the complementary experiment to determine whether MYC overexpression could attenuate the increase in glycolysis observed with BAY treatment (**Fig \@ref(fig:m7)F-H**).
-MYC increased the proliferation rate of DMSO-treated cells, although it did not augment the proliferation rate of BAY-treated cells (**Fig \@ref(fig:m7)G**).
+Together, these data suggest a model whereby hypoxia activates MYC and inhibits HIF-driven increases in glycolysis.
+Inhibiting MYC in hypoxia, therefore, should increase glycolysis.
+Conversely, HIF-dependent glycolysis after BAY treatment in normoxia should be sensitive to inhibition by MYC overexpression (**Fig \@ref(fig:m7)C**).
+To test the hypothesis that hypoxia-induced MYC expression inhibits glycolysis in primary cells, we first combined MYC knockdown with hypoxia treatment (**Fig \@ref(fig:m7)D-F**).
+As expected, MYC-deficient cells proliferated more slowly in normoxia and MYC was essential for sustaining cell proliferation in hypoxia (**Fig \@ref(fig:m7)E**).
+Consistent with our hypothesis, MYC-knockdown cells demonstrated increased lactate efflux with hypoxia culture, unlike control siRNA-treated cells (**Fig \@ref(fig:m7)F**).
+We next performed the complementary experiment to determine whether MYC overexpression could attenuate the increase in glycolysis observed with BAY treatment (**Fig \@ref(fig:m7)G-I**).
+MYC increased the proliferation rate of DMSO-treated cells, although it did not augment the proliferation rate of BAY-treated cells (**Fig \@ref(fig:m7)H**).
 As expected, MYC overexpression blocked the BAY-stimulated increase in lactate efflux.
@@
 The literature describes a complex and reciprocal relationship between HIF and MYC that depends on both environmental (*e.g.*, hypoxia) and cellular context [@li2020d].
-Generally, HIF-1 has been observed to inhibit MYC through multiple mechanisms [@koshiji2004; @koshiji2005; @gordan2007; @zhang2007], and this previous work is consistent with our present observations that HIF stabilization following BAY treatment decreased MYC protein and target gene expression.
-Conversely, MYC has been implicated in increased HIF activity through transcriptional and post-transcriptional mechanisms [@zhang2009a; @chen2013; @doe2012], primarily in the context of malignant transformation.
+Generally, HIF-1 has been observed to inhibit MYC through multiple mechanisms [@koshiji2004; @koshiji2005; @gordan2007; @zhang2007], and this previous work is consistent with our present observations that HIF stabilization following BAY treatment in normoxia decreased MYC protein and target gene expression.
+Conversely, MYC has been implicated in increased HIF activity through transcriptional and post-transcriptional mechanisms [@zhang2009a; @chen2013; @doe2012], primary in the context of malignant transformation.
 The observation that MYC may antagonize the transcriptional effects of HIF to sustain primarily cell proliferation and metabolism in hypoxia suggests a substantially different regulatory relationship than has been previously described.
@@
 In this way, hypoxic MYC activation may sustain energy production by oxidative phosphorylation, thereby decreasing the energetic demands that would otherwise drive increased glycolytic flux.
-Beyond oxidative phosphorylation, MYC targets genes involved in many other intermediary metabolic pathways, including amino acids, nucleotides, and lipids [@stine2015], that may also impact the central pathways of carbon metabolism studied in this work.
+Beyond oxidative phosphorylation, MYC regulates genes involved in many other intermediary metabolic pathways, including amino acids, nucleotides, and lipids [@stine2015], that may also impact the central pathways of carbon metabolism studied in this work.
 
@@
 In addition to dioxygenase enzymes, electron transport chain dysfunction resulting from impaired Complex IV activity leads to increased ROS production in hypoxia [@chandel1998].
-Mitochondrial ROS increase the half-lives of several mRNAs in hypoxia, including MYC, as we observed in this work, independent of HIF stabilization [@guzy2005].
+Mitochondrial ROS increase the half-lives of several mRNAs in hypoxia, including MYC, independent of HIF stabilization [@guzy2005].
 Finally, hypoxia imposes a reductive stress on cells associated with an increase in the NADH/NAD^+^ ratio secondary to impaired electron transport (**Fig \@ref(fig:m5)K**)[@chance1955; @garofalo1988].
@@
 
-In addition to its effects on cellular metabolism, another canonical role of HIF-1 activation is slowing of cellular proliferation rate in the face of limited oxygen availability [@hubbi2015], which is precisely what we observed here (**Figs \@ref(fig:m1)B, \@ref(fig:m2)B, \@ref(fig:m5)A**).
+In addition to its effects on cellular metabolism, another canonical role of HIF-1 activation is slowing of cellular proliferation rate in the face of limited oxygen availability [@hubbi2015], which we found here (**Figs \@ref(fig:m1)B, \@ref(fig:m2)B, \@ref(fig:m5)A**).
 The effects of HIF-1 on cell proliferation rate are mediated, in part, by increased expression of cyclin-dependent kinase inhibitor p21 (*CDKN1A*), inhibition of E2F targets [@gardner2001], and inhibition of pro-proliferative MYC signaling [@koshiji2004].
-These transcriptional effects are precisely what is observed in BAY treated LFs in normoxia.
+These transcriptional effects are precisely what we observed in BAY treated LFs in normoxia.
 By contrast, hypoxia culture was associated with decreased expression of p21, consistent with a previous report [@mizuno2009], as well as increased expression of MYC protein and enrichment of MYC target genes.
@@
 Indeed, hypoxia increased oxygen consumption and ROS production in HIF-1α-null mouse embryonic fibroblasts (MEFs), which was associated with increased cell death [@zhang2008a].
-Interestingly, these cells also had increased ATP levels compared to wild type, suggesting that mitochondrial function was adequate under 1% oxygen culture conditions to support oxidative phosphorylation and meet the energy needs of the cells.
+Interestingly, these cells also had increased ATP levels compared to wild type, suggesting that mitochondrial function was adequate under 1% oxygen culture conditions to support oxidative phosphorylation and to meet the energy needs of the cells.
 Given the prominence of HIFs in mediating the transcriptional response to hypoxia, it is somewhat surprising that neither PHD, HIFs, nor their downstream targets were found to be selectively essential as a function of oxygen tension in a genome-wide CRISPR growth screen of K562 human lymphoblasts cultured in normoxia or hypoxia [@jain2020].
@@
 
-Primary normal human lung fibroblasts (LFs) were purchased from Lonza (CC-2512) and cultured in FGM-2 (Lonza CC-3132) in a standard tissue culture incubator in 5% CO~2~ at 37 °C.
+Primary normal human lung fibroblasts (LFs) were purchased from Lonza (CC-2512) and cultured in FGM-2 (Lonza CC-3132).
 Cells from two donors were used in these studies: #33652 (56 y.o., male) and #29132 (19 y.o., female).
+Primary human pulmonary artery smooth muscle cells were purchased from Lonza (CC-2581) and cultured in SmGM-2 (Lonza CC-3182).
+Cells from three donors were used in these studies: #30020 (64 y.o., male), #27662 (35 y.o., male), #26698 (51 y.o., male), and #19828 (51 y.o., male).
 Cell authentication was performed by the vendor.
-Primary human pulmonary artery smooth muscle cells were purchased from Lonza (CC-2581) and cultured in SmGM-2 (Lonza CC-3182) in a standard tissue culture incubator in 5% CO~2~ at 37 °C.
-Cells from multiple donors were used in these studies: #30020 (64 y.o., male), #27662 (35 y.o., male), #26698 (51 y.o., male), and #19828 (51 y.o., male).
-Cell authentication was performed by the vendor.
+Cells were maintained in a standard tissue culture incubator in 5% CO~2~ at 37 °C.
 
@@
 
-For extracellular flux measurements, cells were seeded in either standard growth medium or MCDB131 medium without glucose, glutamine, or phenol red (genDEPOT) supplemented with 2% dialyzed fetal bovine serum (Mediatech) and naturally labeled glucose and glutamine ("light" labeling medium).
+For extracellular flux measurements, cells were seeded in either standard growth medium or MCDB131 medium lacking glucose, glutamine, and phenol red (genDEPOT) which was supplemented with 2% dialyzed fetal bovine serum (Mediatech) and naturally labeled glucose and glutamine ("light" labeling medium).
 For LFs, glucose was supplemented at 8 mM and glutamine was supplemented at 1 mM.
@@
 On Day -1, 25,000 cells were seeded in a 35 mm dish in "light" labeling medium.
-Hypoxia-treated cells were transferred to a tissue culture glovebox set to 0.5% oxygen and 5% CO~2~ (Coy Lab Products).
-Medium was supplemented with DMSO 0.1% or BAY (10 μM) for DMSO and BAY treatment conditions.
-On Day 0, cells were washed with PBS and the medium was changed to either "light" medium for flux measurements or "heavy" medium for tracer experiments.
+Hypoxia-treated cells were transferred to a tissue culture glovebox set at 0.5% oxygen and 5% CO~2~ (Coy Lab Products).
+Medium was supplemented with DMSO 0.1% or BAY (10 μM) for these treatment conditions.
+On Day 0, cells were washed with PBS and the medium was changed to either "light" labeling medium for flux measurements or "heavy" labeling medium containing [1,2-^13^C~2~]-glucose, [U-^13^C~6~]-glucose, [U-^13^C~5~]-glutamine, or [U-^13^C~2~]-lactate for tracer experiments.
 For LFs, samples were collected on Day 0 and every 24 h for 72 h.
@@
 Each individual experiment included triplicate wells for each treatment and time point, and each experiment was repeated 4-8 times.
+
 ## Cell count
@@
 Total DNA in 10 μL of lysate was determined by adding 100 μL of 1X PicoGreen dye in Tris-EDTA buffer and interpolating the fluorescence intensity with a standard curve generated using the λ DNA standard.
-Cell counts were interpolated from a standard curve of DNA obtained from known cell numbers seeded in basal medium (**Fig S1B**).
+Cell counts were interpolated from a standard curve of DNA obtained from known cell numbers seeded in basal medium (**Fig S1B-C**).
 No difference in total cellular DNA was identified between normoxia and hypoxia cultures (**Fig S1D**).
@@
 Standards were prepared in PBS.
+
 ## Pyruvate assay
@@
 
-Approximately 1.25 M LFs were reverse transfected in 6-cm dishes with 40 pmol siMYC or non-targeting siCTL pools (Dharmacon) in 20 uL RNAiMAX (Thermo).
+Approximately 1.25 M LFs were reverse transfected in 6-cm dishes with 40 pmol siMYC or non-targeting siCTL pools (Dharmacon) and 20 μL RNAiMAX (Thermo) in 500 μL OptiMEM (Thermo).
 After 24 h, cells were collected by trypsinization and re-seeded as described in *Metabolic flux protocol* above for growth rate and lactate efflux measurements.
@@
 time in unconditioned culture medium.
-Rates that significantly differed from 0 using Student's *t*-test were incorporated into the flux calculations.
+Rates that significantly differed from 0 by Student's *t*-test were incorporated into the flux calculations.
 Final fluxes were obtained by robust linear modeling of $Me^{k t}$ *versus* $(e^{(\mu + k)t}-1)$ to determine the slope from which $v$ was calculated using equation \@ref(eq:4).
@@
 The mass spectrometer was operated in selected ion monitoring mode with an m/*z* window width of 9.0 centered 1.003355-times half the number of carbon atoms in the target metabolite.
-The resolution was set at 70,000 and AGC target was 1 × 10^5^ ions.
-Peak areas were corrected for quadrupole bias as previously described [@kim2015].
+The resolution was set at 70,000 and AGC target was 1×10^5^ ions.
+Peak areas were corrected for quadrupole bias as in Kim *et al.* [-@kim2015].
 Mass isotope distributions for FBP and 3PG were calculated from full scan chromatograms as described below.
-Raw mass isotopomer distributions were corrected for natural isotope abundance using a custom R package (mzrtools, https://github.com/oldhamlab/mzrtools) employing the method of Fernandez, *et al.* -[@fernandez1996].
+Raw mass isotopomer distributions were corrected for natural isotope abundance using a custom R package (mzrtools, https://github.com/oldhamlab/mzrtools) employing the method of Fernandez, *et al.* [-@fernandez1996].
 
@@
 The mass spectrometer was operated in polarity switching full scan mode from 70-1000 m/*z*.
-Resolution was set to 70,000 and the AGC target was 1x10^6^ ions.
+Resolution was set to 70,000 and the AGC target was 1×10^6^ ions.
 Peak identifications were based on an in-house library of authentic metabolite standards previously analyzed utilizing this method.
@@
 1. Protein turnover occurred at a negligible rate compared to glucose and glutamine consumption.
-1. Acetyl-CoA, aspartate, fumarate, malate, oxaloacetate, and pyruvate existed in cytosolic and mitochondrial pools. Aspartate and malate were allowed to exchange freely between the compartments.
-1. The per cell biomass requirements of proliferating lung fibroblasts were similar to published estimated in other cells [@quek2010; @sheikh2005]. 
-1.
-Dilution of alanine, aspartate, glutamate, glutamine, lactate, and pyruvate was allowed to occur through reversible exchange with unlabeled substrates in the medium as these metabolites were present in unconditioned medium or serum.
+1. Acetyl-CoA and pyruvate existed in cytosolic and mitochondrial pools. Aspartate, fumarate, oxaloacetate, and malate were allowed to exchange freely between the compartments.
+1. The per cell biomass requirements of LFs and PASMCs were similar to published estimates from other cell types [@quek2010; @sheikh2005]. 
 1. Succinate and fumarate are symmetric molecules that have interchangeable orientations when metabolized by TCA cycle enzymes.
@@
 The ratio of fluorescence in buffer-extracted to acid-extracted samples corresponds to the NADH/NAD^+^ ratio.
-Absolute NADH and NAD^+^ were normalized to cell count from cells estimated from total DNA quantification as described above.
+Absolute NADH and NAD^+^ were normalized to estimated cell counts from total DNA quantification as described above.
 
@@
 Four biological replicates were analyzed.
-Library construction and sequencing was performed by BGI Genomics using 100 bp paired end analysis and a read depth of 50M reads per sample.
+Library construction and sequencing was performed by BGI Genomics using 100 bp paired end analysis and a read depth of 50 M reads per sample.
 Sequences were deposited in the NIH SRA (PRJNA721596).
@@
 
-The raw data and annotated analysis code necessary to reproduce this manuscript are contained in an R package research compendium available from the Oldham Lab GitHub repository (https://github.com/oldhamlab/Copeland.2021.hypoxia.flux).
+The raw data and annotated analysis code necessary to reproduce this manuscript are contained in an R package research compendium available from the Oldham Lab GitHub repository (https://github.com/oldhamlab/Copeland.2022.hypoxia.flux).
 Data analysis, statistical comparisons, and visualization were performed in R [@R-base] using the packages referenced in the data supplement.
@@
 Outliers were identified using twice the median absolute deviation as a cutoff threshold.
-Two group comparisons (*e.g.*, 21% *v.* 0.5% oxygen) were performed using Student's *t*-test on log-transformed values with pairing among biological replicates.
-Multifactor comparisons (*e.g.*, BAY and oxygen) were performed using linear mixed-effects models with *post hoc* comparisons based on the multivariate *t* distribution.
+Comparisons were performed using linear mixed-effects models with oxygen, treatment, and their interaction as fixed effects and biological replicate as a random effect.
+Significant differences in estimated marginal means were identified by comparisons to the multivariate *t* distribution.
 Metabolomics and RNA-seq data were analyzed as described above.
-Probability values less than 0.05 were considered significant.
+Probability values less than 0.05 were considered statistically significant.
 

### 2022-08-01 R/1_figures.R

@@ -1818,15 +1818,15 @@ create_resources <- function() {
     "Antibodies", "HRP-α-Mouse IgG", "Cell Signaling Technologies", "7076",
     "Bacterial and virus strains", "c-MYC adenovirus", "Vector Biolabs", "1285",
     "Bacterial and virus strains", "YFP adenovirus", "Oldham et al., 2015", "",
-    "Chemicals, peptides, and recombinant proteins", "[1,2-^13^C~1~] glucose", "Cambridge Isotope Labs", "CLM-504-PK",
-    "Chemicals, peptides, and recombinant proteins", "[U-^13^C~6~] glucose", "Cambridge Isotope Labs", "CLM-1396-PK",
-    "Chemicals, peptides, and recombinant proteins", "[U-^13^C~5~] glutamine", "Cambridge Isotope Labs", "CLM-1822-H-PK",
-    "Chemicals, peptides, and recombinant proteins", "[U-^13^C~3~] lactate", "Sigma", "485926",
+    "Chemicals, peptides, and recombinant proteins", "[1,2-^13^C~1~]-glucose", "Cambridge Isotope Labs", "CLM-504-PK",
+    "Chemicals, peptides, and recombinant proteins", "[U-^13^C~6~]-glucose", "Cambridge Isotope Labs", "CLM-1396-PK",
+    "Chemicals, peptides, and recombinant proteins", "[U-^13^C~5~]-glutamine", "Cambridge Isotope Labs", "CLM-1822-H-PK",
+    "Chemicals, peptides, and recombinant proteins", "[U-^13^C~3~]-lactate", "Sigma", "485926",
     "Chemicals, peptides, and recombinant proteins", "Molidustat (BAY-85-3934)", "Cayman", "15297",
     "Critical commercial assays", "Glucose colorimetric assay kit", "Cayman", "10009582",
     "Critical commercial assays", "ʟ-Lactate assay kit", "Cayman", "700510",
     "Critical commercial assays", "Pyruvate assay kit", "Cayman", "700470",
-    "Depositied data", "Raw and analyzed data", "This paper", "https://github.com/oldhamlab/Copeland.2021.hypoxia.flux",
+    "Depositied data", "Raw and analyzed data", "This paper", "https://github.com/oldhamlab/Copeland.2022.hypoxia.flux",
     "Depositied data", "RNA-seq reads", "This paper", "SRA: PRJNA721596",
[…]

### 2022-08-02 manuscript/manuscript.Rmd
finish edits prior to submission

@@
 
-# rbbt::bbt_update_bib(
-#   path = "manuscript/manuscript.Rmd",
-#   ignore = c("R-base"),
-#   path_bib = "manuscript/library.json"
-# )
+rbbt::bbt_update_bib(
+  path = "manuscript/manuscript.Rmd",
+  ignore = c("R-base"),
+  path_bib = "manuscript/library.json"
+)
 
@@
 This decrease in lactate efflux occurred despite activation of the HIF-1α transcriptional program as reflected by increased expression of GLUT1 and LDHA.
-To test if more severe hypoxia would augment glycolysis, we culture cells in 0.2% ambient oxygen (**Fig S2**).
+To test if more severe hypoxia would augment glycolysis, we cultured cells in 0.2% ambient oxygen (**Fig S2**).
 Under these conditions, we observed no change in glucose or lactate fluxes, similar to 0.5% oxygen culture.
@@
 
-Although the metabolite exchange fluxes for bidirectional reactions tend to be poorly resolved by metabolic flux analysis [@wiechert2007a], two observations are worth highlighting (**Tables S1-S3**).
+Although the metabolite exchange fluxes for bidirectional reactions tend to be poorly resolved by metabolic flux analysis [@wiechert2007], two observations are worth highlighting (**Tables S1-S3**).
 First, consistent with the stable isotope tracing results, the modeled rate of reductive carboxylation through reverse flux by isocitrate dehydrogenase in LFs is low (~4 fmol/cell/h), unchanged by hypoxia, and modestly increased by BAY treatment.
@@
 Differentially regulated metabolites were identified using limma [@ritchie2015a].
-Metabolite set enrichment analysis was performed using the fgsea package [@korotkevich2021] with KEGG metabolite pathways [@kanehisa2000].
+Metabolite set enrichment analysis was performed using the fgsea package [@korotkevich2021a] with KEGG metabolite pathways [@kanehisa2000].
 
@@
 Differentially expressed transcripts were identified using DESeq2 [@love2014].
-Gene set enrichment and transcription factor enrichment was performed using the fgsea and DoRothEA R packages, respectively [@korotkevich2021; @garcia-alonso2019].
+Gene set enrichment and transcription factor enrichment was performed using the fgsea and DoRothEA R packages, respectively [@korotkevich2021a; @garcia-alonso2019].
 
@@
 
-(ref:m1) **Effects of hypoxia on extracellular metabolite fluxes in lung fibroblasts.** (**A**) Lung fibroblasts (LFs) were cultured in 21% (*red*) or 0.5% (*blue*) oxygen beginning 24 h prior to time 0.
+(ref:m1) **Effects of hypoxia on extracellular metabolite fluxes in lung fibroblasts.** (**A**) Lung fibroblasts (LFs) were cultured in 21% or 0.5% oxygen beginning 24 h prior to time 0.
 Samples were collected every 24 h for 72 h.
@@
 (**I**) Extracellular fluxes of glucose (GLC) and lactate (LAC) (n = 8).
+By convention, negative fluxes indicate metabolite consumption.
 (**J**) Extracellular fluxes of pyruvate (PYR) and amino acids.
-Data are mean ± SEM.
-Comparisons were made using linear mixed effects models with oxygen as the fixed effect and biological replicate as a random effect with p-values adjusted using the multivariate *t* distribution (* p < 0.05).
+Data are mean ± SEM (* p < 0.05).
 
-(ref:m2) **Effects of pharmacologic prolyl hydroxylase inhibition on extracellular metabolite fluxes in lung fibroblasts.** (**A**) Lung fibroblasts (LFs) were treated with PHD inhibition molidustat (BAY, 10 μM) (*purple*) or DMSO (*green*) beginning 24 h prior to time 0.
+(ref:m2) **Effects of pharmacologic prolyl hydroxylase inhibition on extracellular metabolite fluxes in lung fibroblasts.** (**A**) Lung fibroblasts (LFs) were treated with the prolyl hydroxlyase inhibitor molidustat (BAY, 10 μM) or DMSO beginning 24 h prior to time 0.
 Samples were collected every 24 h for 72 h.
@@
 (**I**) Extracellular fluxes of glucose (GLC) and lactate (LAC) (n = 8).
+By convention, negative fluxes indicate metabolite consumption.
 (**J**) Extracellular fluxes of pyruvate (PYR) and amino acids.
-Data are mean ± SEM.
-Comparisons were made using linear mixed effects models with treatment as the fixed effect and biological replicate as a random effect with p-values adjusted using the multivariate *t* distribution (* p < 0.05).
+Data are mean ± SEM (* p < 0.05).
 
-(ref:m3) **Metabolic flux analysis of lung fibroblasts following hypoxic and pharmacologic PHD inhibition.** (**A**) Mass isotopomer distribution (MID) of pyruvate (PYR) following 72 h labeling of LFs with [U-^13^C~6~]-glucose (GLC).
-(**B**) MID of citrate after 72 h labeling of LFs with [U-^13^C~6~]-GLC (**C**) MID of citrate after 72 h labeling of LFs with [U-^13^C~5~]-glutamine (GLN).
-Data are mean ± SEM.
-Comparisons were made using linear mixed effects models with treatment as the fixed effect and biological replicate as a random effect with p-values adjusted using the multivariate *t* distribution (n = 4, p < 0.05 indicated as * 0.5% *v.* 21% oxygen, † BAY *v.* DMSO, ‡ Δoxygen *v.* ΔBAY).
-(**D**) Fraction of M5 citrate indicating reductive carboxylation after labeling with [U-^13^C~5~]-GLN in LFs and PASMCs (n = 3-4, * p < 0.05 0.5% *v.* 21% oxygen).
-(**E**) Ratio of metabolic fluxes in 0.5% oxygen compared to 21% oxygen.
-Fluxes with non-overlapping confidence intervals are highlighted with arrows colored according to the magnitude of the change.
+(ref:m3) **Metabolic flux analysis of lung fibroblasts following hypoxic and pharmacologic PHD inhibition.** (**A**) Mass isotopomer distribution (MID) of pyruvate (PYR) following 72 h labeling with [U-^13^C~6~]-glucose (GLC).
+(**B**) MID of citrate after 72 h labeling with [U-^13^C~6~]-GLC (**C**) MID of citrate after 72 h labeling with [U-^13^C~5~]-glutamine (GLN).
+Data are mean ± SEM (n = 4, p < 0.05 indicated as * 0.5% *v.* 21% oxygen, † BAY *v.* DMSO, ‡ Δoxygen *v.* ΔBAY).
+(**D**) Fraction of M5 citrate indicating reductive carboxylation after labeling with [U-^13^C~5~]-GLN in LFs and PASMCs (n = 3-4, * p < 0.05).
+(**E**) Ratio of modeled metabolic fluxes in 0.5% oxygen compared to 21% oxygen.
+Fluxes with non-overlapping confidence intervals are highlighted with arrows colored according to the magnitude of the fold change.
 Arrow thickness corresponds to the absolute flux measured in hypoxia.
@@
 (**C**) Mass isotopomer distributions of lactate (LAC), fructose-bisphosphate (FBP), pyruvate (PYR), and citrate (CIT) in LFs following 72 h labeling with [U-^13^C~3~] lactate.
-Data are mean ± SEM.
-Comparisons were made using linear mixed effects models with treatment as the fixed effect and biological replicate as a random effect with p-values adjusted using the multivariate *t* distribution (n = 4, p < 0.05 indicated as * 0.5% *v.* 21% oxygen, † BAY *v.* DMSO, ‡ Δoxygen *v.* ΔBAY).
+Data are mean ± SEM (n = 4, p < 0.05 indicated as * 0.5% *v.* 21% oxygen, † BAY *v.* DMSO, ‡ Δoxygen *v.* ΔBAY).
 
@@
 (**C**) Lactate efflux in LFs treated with BAY cultured in hypoxia.
-Data are mean ± SEM.
-Comparisons were made using linear mixed effects models with oxygen × treatment as the fixed effects and biological replicate as a random effect with p-values adjusted using the multivariate *t* distribution (* p < 0.05 BAY *v.* DMSO within a given oxygen exposure).
+Data are mean ± SEM (* p < 0.05 BAY *v.* DMSO within a given oxygen exposure).
 (**D**) Principal component analysis of intracellular metabolites following 72 h of treatment as in (A) suggests a dominant effect of hypoxia over pharmacologic PHD inhibition on the metabolome (n = 7).
-(**E**) Volcano plot of intracellular metabolites of BAY-treated cells cultures in 21% or 0.5% oxygen.
-*Blue* circles indicate increased levels in hypoxia while *red* circles indicate decreases with hypoxia.
+(**E**) Volcano plot of intracellular metabolites of BAY-treated cells cultured in 21% or 0.5% oxygen.
+Circles indicate significantly increased (*blue*) and decreased (*red*) levels in hypoxia.
 (**F**) Metabolite set enrichment analysis of metabolites from (E).
-KEGG pathways with p-values < 0.05 are shown.
+All KEGG pathways with p-values < 0.05 are shown.
 (**G**) Leading edge analysis of the TCA cycle metabolite set.
-Negative values indicate relative enrichment associated with BAY treatment compared to hypoxia treatment.
+Negative values indicate the relative enrichment associated with BAY treatment compared to hypoxia treatment.
 Abbreviations: PYR, pyruvate; SUC, succinate; PEP, phosphoenolpyruvate; CIT, citrate; AKG, α-ketoglutarate; MAL, malate; ACO, aconitate; FUM, fumarate.
@@
 NAD and NADH were determined by enzymatic assay.
-Data are mean ± SEM (n = 5).
-Comparisons were made using linear mixed effects models with treatment as the fixed effect and biological replicate as a random effect with p-values adjusted using the multivariate *t* distribution (p < 0.05 with *black* * indicating a significant effect of treatment within a given oxygen tension and *colored* * indicating a significant effect of treatment within a given oxygen tension denoted by the color).
+Data are mean ± SEM (n = 5, p < 0.05 with *black* * indicating a significant effect of treatment within a given oxygen tension and *colored* * indicating a significant effect of treatment within a given oxygen tension denoted by the color).
 
-(ref:m6) **Transcriptomic analysis of molidustat treatment in normoxia and hypoxia.** (**A**) Principal component analysis of transcriptional changes following 72 h of treatment with 0.5% oxygen or molidustat (BAY), separately or together (n = 4).
+(ref:m6) **Transcriptomic analysis of molidustat treatment in normoxia and hypoxia.** (**A**) Principal component analysis of transcriptional changes in lung fibroblasts following 72 h of treatment with 0.5% oxygen or molidustat (BAY), separately or together (n = 4).
 (**B**) Volcano plot illustrating the effects of hypoxia in BAY-treated cells on gene expression.
@@
 
-(ref:m7) **MYC regulates HIF-dependent glycolytic flux.** (**A**) Representative immunoblot of MYC protein expression in lung fibroblasts following 72 h of treatment with 0.5% oxygen or molidustat (BAY, B).
+(ref:m7) **MYC regulates HIF-dependent glycolytic flux.** (**A**) Representative immunoblot of MYC protein expression in lung fibroblasts (LFs) following 72 h of treatment with 0.5% oxygen or molidustat (BAY, B).
 (**B**) Quantification of band densities from (A).
 (**C**) Working model illustrating that hypoxia-induced MYC antagonizes HIF-dependent metabolic and transcriptional events.
-(**D**) Representative immunoblot of lung fibroblasts (LFs) treated with siRNA targeting MYC (M) demonstrating adequate protein knockdown.
+(**D**) Representative immunoblot of LFs treated with siRNA targeting MYC (M) demonstrating adequate protein knockdown.
 (**E**) Growth rates of MYC-knockdown cells cultured in hypoxia.
@@
 (**I**) Lactate efflux rates of MYC overexpressing cells cultured with BAY.
-Data are mean ± SEM of n = 3-7 biological replicates.
-Comparisons were made using linear mixed effects models with oxygen × treatment as the fixed effects and biological replicate as a random effect.
-The multivariate *t* distribution was used to determine significant differences (p < 0.05) as indicated by black * for differences within groups and colored * for differences between groups defined by the *x*-axis.
+Data are mean ± SEM (n = 3-7, p < 0.05 as indicated by black * for differences within groups and colored * for differences between groups defined by the *x*-axis).
 

### 2022-08-02 manuscript/supplement.Rmd
finish edits prior to submission

@@
 
-(ref:s1) **Supporting data for extracellular flux calculations.** (**A**) Cell viability as assessed by live/dead cell staining with acridine orange plus propidium iodide staining did not differ between 21% and 0.5% oxygen culture conditions (n = 3 technical replicates).
+(ref:s1) **Supporting data for extracellular flux calculations.** (**A**) Cell viability as assessed by live/dead cell staining with acridine orange and propidium iodide did not differ between 21% and 0.5% oxygen culture conditions (n = 3 technical replicates).
 (**B**) Standard curve of lung fibroblast (LF) cell count *v.* total DNA by PicoGreen measurement used to interpolate cell numbers from DNA measurements.
@@
 
-(ref:s2) **Effects of hypoxia on extracellular metabolite fluxes in lung fibroblasts.** (**A**) Lung fibroblasts (LFs) were cultured in 21% (*red*) or 0.2% (*dark blue*) oxygen beginning 24 h prior to time 0.
+(ref:s2) **Effects of hypoxia on extracellular metabolite fluxes in lung fibroblasts.** (**A**) Lung fibroblasts (LFs) were cultured in 21% or 0.2% oxygen beginning 24 h prior to time 0.
 Samples were collected every 24 h for 72 h.
@@
 (**I**) Extracellular fluxes of glucose (GLC) and lactate (LAC) (n = 4).
+By convention, negative fluxes indicate metabolite consumption.
 (**J**) Extracellular fluxes of pyruvate (PYR) and amino acids.
-Data are mean ± SEM.
-Comparisons were made using linear mixed effects models with oxygen as the fixed effect and biological replicate as a random effect with p-values adjusted using the multivariate *t* distribution (* p < 0.05).
+Data are mean ± SEM (* p < 0.05).
 
-(ref:s3) **Effects of hypoxia on extracellular metabolite fluxes in pulmonary artery smooth muscle cells.** (**A**) Pulmonary artery smooth muscle cells (PASMCs) were cultured in 21% (*red*) or 0.5% (*blue*) oxygen beginning 24 h prior to time 0.
+(ref:s3) **Effects of hypoxia on extracellular metabolite fluxes in pulmonary artery smooth muscle cells.** (**A**) Pulmonary artery smooth muscle cells (PASMCs) were cultured in 21% or 0.5% oxygen beginning 24 h prior to time 0.
 Samples were collected every 12 h for 48 h.
@@
 (**I**) Extracellular fluxes of glucose (GLC) and lactate (LAC) (n = 4).
+By convention, negative fluxes indicate metabolite consumption.
 (**J**) Extracellular fluxes of pyruvate (PYR) and amino acids.
-Data are mean ± SEM.
-Comparisons were made using linear mixed effects models with oxygen as the fixed effect and biological replicate as a random effect with p-values adjusted using the multivariate *t* distribution (* p < 0.05).
+Data are mean ± SEM (* p < 0.05).
 
@@
 (**B**) Metabolic flux model of PASMC metabolism in 21% oxygen as in (A).
-(**C**) The ratio between PASMC and LF fluxes in normoxia were calculated.
+(**C**) Ratio of metabolic fluxes in PASMCs compared to LFs.
 Fluxes with non-overlapping confidence intervals are highlighted with arrows colored according to the magnitude of the change.

### 2022-08-02 R/1_figures.R

@@ -559,7 +559,7 @@ plot_high_fluxes <- function(
       geom = "errorbar",
       fun.data = ggplot2::mean_se,
       position = ggplot2::position_dodge(width = 0.9),
-      width = 0.2,
+      width = 0.3,
       size = 0.25,
       show.legend = FALSE
     ) +
@@ -659,7 +659,7 @@ plot_low_fluxes <- function(
       geom = "errorbar",
       fun.data = ggplot2::mean_se,
       position = ggplot2::position_dodge(width = 0.9),
-      width = 0.2,
+      width = 0.3,
       size = 0.25,
       show.legend = FALSE
     ) +
@@ -834,10 +834,10 @@ plot_mids <- function(df, cell, metab, t = 72, track) {
 
 plot_all_mids <- function(df, cell, t = 72, track) {
   tracer_labels <- c(
-    expression(paste("[1,2-"^13, "C"[2], "] glucose")),
-    expression(paste("[U-"^13, "C"[6], "] glucose")),
-    expression(paste("[U-"^13, "C"[5], "] glutamine")),
-    expression(paste("[U-"^13, "C"[3], "] lactate"))
+    expression(paste("[1,2-"^13, "C"[2], "]-glucose")),
+    expression(paste("[U-"^13, "C"[6], "]-glucose")),
+    expression(paste("[U-"^13, "C"[5], "]-glutamine")),
+    expression(paste("[U-"^13, "C"[3], "]-lactate"))
   )
   tracer_levels <- c("glc2", "glc6", "q5", "lac3")
   metab <- c(
@@ -1065,10 +1065,10 @@ plot_m5_citrate <- function(df) {
 
 format_time_course_mids <- function(model_mids) {
   tracer_labels <- c(
-    expression(paste("[1,2-"^13, "C"[2], "] glucose")),
-    expression(paste("[U-"^13, "C"[6], "] glucose")),
-    expression(paste("[U-"^13, "C"[5], "] glutamine")),
-    expression(paste("[U-"^13, "C"[3], "] lactate"))
+    expression(paste("[1,2-"^13, "C"[2], "]-glucose")),
+    expression(paste("[U-"^13, "C"[6], "]-glucose")),
+    expression(paste("[U-"^13, "C"[5], "]-glutamine")),
+    expression(paste("[U-"^13, "C"[3], "]-lactate"))
   )
   tracer_levels <- c("glc2", "glc6", "q5", "lac3")
 
[…]

### 2022-08-02 R/1_metab.R

@@ -86,6 +86,7 @@ format_metab_tar <- function(files){
     dplyr::mutate(
       metabolite = dplyr::case_when(
         metabolite == "glyceraldehyde 3-phosphate" ~ "GAP",
+        metabolite == "sedoheptulose 7-phosphate" ~ "S7P",
         TRUE ~ metabolite
       )
     ) |>
@@ -290,7 +291,7 @@ plot_metab_pca <- function(clean, df) {
         label = label
       ),
       expand = ggplot2::unit(2, "mm"),
-      label.fontsize = 5,
+      label.fontsize = 6,
       label.fontface = "plain",
       label.family = "Calibri",
       label.hjust = 0.5,
@@ -410,7 +411,7 @@ plot_metab_volcano <- function(
         label = metabolite,
         color = metabolite %in% mois
       ),
-      size = 5/ggplot2::.pt,
+      size = 6/ggplot2::.pt,
       max.overlaps = 20,
       segment.size = 0.1,
       # nudge_x = -4,
@@ -427,7 +428,7 @@ plot_metab_volcano <- function(
         label = metabolite,
         color = metabolite %in% mois
       ),
-      size = 5/ggplot2::.pt,
+      size = 6/ggplot2::.pt,
       max.overlaps = 20,
       segment.size = 0.1,
       segment.color = "black",
@@ -609,7 +610,7 @@ plot_metab_venn <- function(hyp, bay) {
       digits = 0,
       show_percentage = TRUE,
       fill_color = clrs[c(2, 4)],
-      fill_alpha = 0.5,
+      fill_alpha = 0.25,
       stroke_size = 0.25,
       set_name_size = 8/ggplot2::.pt,
       text_size = 6/ggplot2::.pt

### 2022-08-02 R/1_myc.R

@@ -61,12 +61,22 @@ annot_myc_fluxes <- function(df, intervention) {
 }
 
 plot_myc <- function(df, annot, metab, ylab, x, fill) {
-  df |>
+  z <-
+    df |>
+    dplyr::ungroup() |>
     dplyr::filter(metabolite == metab) |>
-    ggplot2::ggplot() +
+    dplyr::mutate(grand_mean = mean(flux)) |>
+    dplyr::group_by(date) |>
+    dplyr::mutate(
+      exp_mean = mean(flux),
+      adj = grand_mean - exp_mean,
+      flux_corr = flux + adj
+    )
+
+  ggplot2::ggplot(z) +
     ggplot2::aes(
       x = {{x}},
-      y = flux
+      y = flux_corr
     ) +
     ggplot2::stat_summary(
       ggplot2::aes(
@@ -123,7 +133,7 @@ plot_myc <- function(df, annot, metab, ylab, x, fill) {
     ggplot2::scale_fill_manual(
       values = clrs,
       limits = force
-      ) +
+    ) +
     ggplot2::scale_y_continuous(
       expand = ggplot2::expansion(mult = c(0.05, 0.1)),
       breaks = scales::pretty_breaks(n = 6)

### 2022-08-02 R/1_rnaseq.R

@@ -72,7 +72,7 @@ plot_rnaseq_pca <- function(pca_data){
         label = label
       ),
       expand = ggplot2::unit(2, "mm"),
-      label.fontsize = 5,
+      label.fontsize = 6,
       label.fontface = "plain",
       label.family = "Calibri",
       label.hjust = 0.5,
@@ -174,7 +174,7 @@ plot_rnaseq_volcano <- function(results, gois = NULL, xlab = NULL, nudge = 8){
         label = symbol,
         color = symbol %in% gois
       ),
-      size = 5/ggplot2::.pt,
+      size = 6/ggplot2::.pt,
       max.overlaps = 20,
       segment.size = 0.1,
       nudge_x = -nudge - left$log2FoldChange,
@@ -190,7 +190,7 @@ plot_rnaseq_volcano <- function(results, gois = NULL, xlab = NULL, nudge = 8){
         label = symbol,
         color = symbol %in% gois
       ),
-      size = 5/ggplot2::.pt,
+      size = 6/ggplot2::.pt,
       max.overlaps = 20,
       segment.size = 0.1,
       nudge_x = nudge - right$log2FoldChange,
@@ -201,7 +201,7 @@ plot_rnaseq_volcano <- function(results, gois = NULL, xlab = NULL, nudge = 8){
       show.legend = FALSE
     ) +
     ggplot2::geom_hex(
-      bins = c(50, 2),
+      bins = c(50, 4),
       show.legend = FALSE
     ) +
     ggplot2::scale_fill_viridis_c(trans = "log10") +
@@ -545,7 +545,7 @@ plot_rnaseq_venn <- function(hyp, bay) {
       digits = 0,
       show_percentage = TRUE,
       fill_color = clrs[c(2, 4)],
-      fill_alpha = 0.5,
+      fill_alpha = 0.25,
       stroke_size = 0.25,
       set_name_size = 8/ggplot2::.pt,
       text_size = 6/ggplot2::.pt
@@ -598,7 +598,7 @@ plot_gsea_venn <- function(hyp, bay) {
       digits = 0,
       show_percentage = TRUE,
       fill_color = clrs[c(2, 4)],
-      fill_alpha = 0.5,
+      fill_alpha = 0.25,
       stroke_size = 0.25,
       set_name_size = 8/ggplot2::.pt,
       text_size = 6/ggplot2::.pt
[…]

### 2022-08-02 _targets.R

@@ -638,7 +638,7 @@ list(
     values = list(
       names = list("hyp", "bay", "hyp_bay", "int"),
       colors = list(clrs[2:1], clrs[4:3], clrs[2:1], clrs[c(2, 4)]),
-      xlab = list("Hypoxia/Normoxia", "BAY/DMSO", "Hypoxia/Normoxia", "ΔHypoxia/ΔBAY"),
+      xlab = list("Hypoxia/Normoxia", "BAY/DMSO", "Hypoxia/Normoxia in BAY", "ΔHypoxia/ΔBAY"),
       title = list("Hypoxia/Normoxia", "BAY/DMSO", "Hypoxia/Normoxia in BAY", "ΔHypoxia/ΔBAY"),
       filename = stringr::str_c("msea_", list("hyp", "bay", "hyp_bay", "int"))
     ),

### 2022-08-03 manuscript/manuscript.Rmd
final revisions before journal upload

@@
 - packages.bib
-csl: cell-metabolism.csl
+csl: embo.csl
 link-citations: yes
@@
 
-Hypoxia requires metabolic adaptations to sustain energetically demanding cellular activities that underpin many physiologic and pathologic processes.
-While many studies have explored the metabolic consequences of hypoxia in cancer cell models, comparatively little is known about the metabolic response of primary cells to hypoxia.
-Here, we performed metabolic flux analyses of proliferating human lung fibroblasts and pulmonary artery smooth muscle cells in hypoxia.
-Unexpectedly, glycolytic flux was decreased in hypoxic cells despite activation of the hypoxia-inducible factor (HIF) transcriptional program and increased expression of glycolytic enzymes.
+Hypoxia requires metabolic adaptations to sustain energetically demanding cellular activities.
+While the metabolic consequences of hypoxia have been studied extensively in cancer cell models, comparatively little is known about the metabolic response of primary cells to hypoxia.
+We performed metabolic flux analyses of proliferating human lung fibroblasts and pulmonary artery smooth muscle cells in hypoxia.
+Unexpectedly, hypoxia decreased glycolytic flux despite activation of hypoxia-inducible factor (HIF) and increased glycolytic enzyme expression.
 Pharmacologic activation of HIF with the prolyl hydroxylase (PHD) inhibitor molidustat in normoxia did increase glycolytic flux, but hypoxia abrogated this effect.
-Multi-omic profiling of cells treated with hypoxia or molidustat, separately or together, revealed distinct molecular responses to hypoxia and pharmacologic PHD inhibition and suggested a critical role for MYC in modulating the HIF response in hypoxia.
+Multi-omic profiling revealed distinct molecular responses to hypoxia and pharmacologic PHD inhibition and suggested a critical role for MYC in modulating the HIF response in hypoxia.
 MYC knockdown in hypoxia increased lactate efflux, while MYC overexpression in normoxia blunted the effects of molidustat treatment.
@@
 
-The authors declare that they have no competing interests.
+The authors declare that they have no conflicts of interest.
 

### 2022-08-03 _targets.R

@@ -1478,10 +1478,6 @@ list(
         "--lua-filter=author-info-blocks.lua",
         "--lua-filter=pagebreak.lua"
       )
-    ),
-    params = list(
-      bibliography = c("manuscript/library.json", pkgs),
-      csl = csl
     )
   ),
 
@@ -1503,11 +1499,6 @@ list(
         "--lua-filter=pagebreak.lua",
         "--lua-filter=multiple-bibliographies.lua"
       )
-    ),
-    params = list(
-      bibliography_main = "manuscript/library.json",
-      bibliography_software = pkgs,
-      csl = csl
     )
   ),
   NULL
